### 2012-06-14 paper.tex
First version

\documentclass{article}

% *** CITATION PACKAGES ***

\usepackage{cite}

% *** GRAPHICS RELATED PACKAGES ***

\usepackage[pdftex]{graphicx}
\usepackage{tikz}

% *** MATH PACKAGES ***

\usepackage{amsmath} 
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{latexsym} 
\usepackage{amstext}
\usepackage{amsxtra} 
\usepackage{amsfonts} 
\usepackage{graphicx}

\usepackage{algorithmic}
\usepackage{algorithm}

% *** ALIGNMENT PACKAGES ***

\usepackage{array} 
\usepackage{mdwmath}
\usepackage{mdwtab}

\usepackage[tight,footnotesize,caption=false]{subfig}

% *** CITATION PACKAGE *** %
% ** IJRR asks for an author-date bibtex style ** %
\usepackage{harvard}

\newtheorem{theorem}{Theorem}

\begin{document}

\title{Whole-Body Manipulation Planning for Humanoid Robots}

\author{
  S\'ebastien Dalibard\and
  Antonio El Khoury\and
  Florent Lamiraux\and
  Alireza Nakhaei\and
  Michel Ta\"ix\and
  Jean-Paul Laumond
  \footnote{The authors are with CNRS ; LAAS ; 7 avenue du colonel Roche, F-31077
    Toulouse Cedex 4, France and Universit\'e de Toulouse ; UPS, INSA, INP, ISAE ;
    UT1, UTM, LAAS ; F-31077 Toulouse
    Cedex 4, France. S\'ebastien Dalibard is now with Nanyang Technological University, Singapore.}
  \footnote{
    This paper summarizes and and extends previous work that appeared in the 9th and 11th
    IEEE-RAS International Conference on Humanoid Robots, 2009 and 2011.
  }
}

\date{}

\maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

Redundancy, walk

Local planning

different levels of complexity

\subsection{Outline}
Section~\ref{sec:related} reviews the related work and states our contribution. 
Section~\ref{sec:wb} presents a constrained motion planning algorithm, and its
use on a humanoid robot manipulation problem. Section~\ref{sec:wb-step} generalizes
the previous algorithm to problems that require locomotion. The generalization is
well-grounded, and
based on a controllability property of legged robots demonstrated in the paper. 
Section~\ref{sec:exp} presents some experimental results, and 
Section~\ref{sec:limits} discusses the limitations and potential future work of our
method.

\section{Related Work and Contribution}
\label{sec:related}

This work is based on several fields of humanoid robotics
research: prioritized inverse kinematics, randomized  whole-body  motion
planning and walk pattern generators based on the Zero-Momentum
Point (ZMP) formalism. This section summarizes the related work for each of them.

\subsection{Prioritized Inverse Kinematics}

The problem of inverse kinematics for a humanoid robot, or any articulated
structure, is to compute a joint motion to achieve an end-effector pose. As the
robots we deal with are redundant, it is natural to take advantage of
this redundancy by specifying multiple tasks, potentially with
different priorities. This problem has been widely studied in robotics
planning and control literature, and many jacobian-based solutions have been
proposed, among which 
\cite{nakamura1986iks}, \cite{siciliano1991gfm},
\cite{baerlocher1998tpf} and \cite{khatib2004wbd}.
Obstacle avoidance can be taken into account with similar methods. To
do so, one has to include the obstacles as other task constraints to
be satisfied. A recent  contribution on this subject is \cite{kanehiro2008lca}.
These methods are prone to fall into local minima, global motion 
planning is needed to overcome this limitation.

\subsection{Whole-Body Motion Planning}

When  planning a  whole-body motion  for a  humanoid robot, one difficult
challenge is to cope with  the curse of dimensionality. The complexity
of   motion  planning  is   exponential  in   the  dimension   of  the
configuration  space ($\mathcal{C}$)  to explore.  When  dealing with
high-dimensional configuration  spaces, it is  typically impossible to
explicitly represent  them, leading to the use  of randomized sampling
techniques  to solve  global planning  problems. In  the  past fifteen
years,  \textit{Probabilistic Roadmaps} \cite{kavraki1996prp} and  
\textit{Rapidly exploring Random  Trees} (RRT) 
\cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
high-dimensional   planning  problems.
When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
difficulty  is to  take into  account stability  constraints,  i.e. to
generate  random  configurations   on  zero  volume  submanifolds  of
$\mathcal{C}$. This problem has been investigated with success during
the last few years, \cite{Berenson15032011} presents a exhaustive survey
of jacobian-based methods. Other recent contributions \cite{porta2012randomized}
present sophisticated constrained planning techniques based on higher-dimensional
continuation. Section \ref{sec:wb} presents a simple adaptation
of the RRT algorithm to constrained motion planning, that was first
introduced in \cite{dalibard09}.

\subsection{Walk Pattern Generation}

Another  field of  humanoid  robotics research  is  the generation  of
dynamically stable  walk patterns. Since  the introduction of  the ZMP
formalism  \cite{vukobratovic1969contribution},  several  methods  have  been  proposed  to
generate  walking  motions efficiently.   One  way  to  deal with  the
complexity  of a  humanoid  robotics  kinematic tree  is  to use  the
so-called "cart-table" simplified model \cite{kajita2003biped}. Based on such a
model,  planning a trajectory  for the  ZMP is  reduced to  planning a
trajectory  for  the Center  of  Mass (CoM)  of  the  robot.  Given  a
trajectory  of  the CoM  and  footstep  positions, inverse  kinematics
solvers can animate  the whole set of degrees of freemdom (DoFs) of the  
robot to generate a dynamically stable walk trajectory.

\subsection{Collision-Free Walk Planning}

Collision-free  locomotion   trajectories  are  usually   obtained  by
simplifying the model of the  robot or of its environment. By reducing 
a  humanoid robot to a bounding volume, that includes its swaying motions,
one can  use a simple  planar motion planner  on this bounding  volume and
generate  a valid  locomotion  trajectory. This  strategy  is used  in
\cite{pettre20032} in a computer animation context. Variants of this method
include  dynamic path reshaping  \cite{yoshida-humanoids05}: if  collisions appear
when animating  the locomotion  trajectory, it is  locally reshaped
and re-animated.  This two-stage  strategy does not guarantee that the
locomotion trajectory can be followed or that the local reshaping will
converge.

Simplifying  the environment  consists in  considering obstacles  at a
footstep   level.   \cite{chestnutt2005footstep,kuffner2005motion}
use   an  
A$^{*}$   algorithm  to   find
collision-free   footsteps.    In   \cite{perrinbiped},   the   authors   compute
collision-free motions  for the legs by using  an RRT$^{*}$ algorithm.

Some planning methods for free-climbing robots \cite{bretl2006motion}
can be seen as a general way 
to consider quasi-static multi-step planning. They are not directly applicable to
humanoid dynamically stable locomotion.
Other recent contributions to the field of locomotion planning  include algorithms 
considering the dynamics at the planning phase \cite{shkolnik2011bounding}. This leads to 
a growth of algorithmic complexity, particularly costly for high-dimensional
systems such as humanoid robots. In our work, we show that it is not always
necessary to take into account dynamics while planning, even though the final
output of our algorithm is a dynamically stable walk motion.

\subsection{Contribution}

The main contribution of this work is a  whole-body motion planner for humanoid robots
that computes collision-free walking trajectories, based on
exact models of both robot and environment. It is used to solve manipulation tasks
that may require walking. The first stage of our algorithm uses a
constrained sampling-based motion 
planner and  computes a collision-free path  for a statically stable robot on the
ground. Another  contribution of this  paper is the formal  proof that
this first path can always be  approximated by a dynamically balanced, collision-free
walking  trajectory in a second stage. We have implemented this well-grounded 
method and  used it on a model of  HRP-2 robot.  The results have been  
validated on a real platform.

\section{Randomized Motion Planning on Constrained Manifolds}
\label{sec:wb}

This section presents an algorithm for constrained motion planning. 
The problem solved here
differs from classic configuration approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
  set of inverse kinematic tasks;
\item the goal is also defined implicitely, by another set of inverse kinematic tasks.
\end{enumerate}
In this presentation, we use a prioritized inverse kinematic solver, or local planner,
 as a black box. We will not describe the details of its implementation. They are presented in 
\cite{yoshida2006tds}, and are out of scope of the present work.

During global planning, we will consider several types of
tasks and constraints for various reasons:
\begin{itemize}
\item  Static stability: the center of mass of the robot stays at
  the vertical of the support polygon center, the two feet have fixed position and
  orientation.
\item End-effector position and orientation: the goals of some problems presented
  in the experimental section of this paper are defined as robot hand configuration,
  or gaze direction.
\item Configuration task: our adaptation of randomized motion planning algorithms 
  uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
  This will be detailed in the following of the section.
\end{itemize}

The architecture of the constrained planning algorithm presented here follows
the one of randomized tree expansion strategies, introduced in both \cite{HsuLat99c} and
\cite{kuffner00rrtconnect}. Next subsection will recall the structure of the RRT algorithm,
a popular randomized motion planning algorithm.

\subsection{Rapidly exploring Random Trees (RRT)}

The classic RRT algorithm, as presented in  \cite{kuffner00rrtconnect}, grows 
a random tree inside a robot 
collision-free configuration space 
$\mathcal{C}_{free}$. Each iteration of the algorithm attempts to extend the tree
by adding new vertices in the direction of a randomly-selected configuration
$q_{rand}$. Algorithm~\ref{algRRT} shows the pseudo-code of the RRT algorithm.
It takes as input an initial configuration $q_0$ and grows a tree  $\mathcal{T}$ rooted 
in $q_0$.

\begin{algorithm}
\caption{RRT($q_0$)}
\label{algRRT}
\begin{algorithmic}
\STATE $\mathcal{T}.$Init$(q_0)$
\FOR{$i$ = 1 to $K$}
\STATE $q_{rand} \leftarrow $ Rand$(\mathcal{C})$
\STATE $q_{near} \leftarrow $ Nearest$(q_{rand},\mathcal{T})$
\STATE Extend$(q_{near},q_{rand})$
\ENDFOR

\end{algorithmic}
\end{algorithm}

One way to make the RRT algorithm more efficient is to grow trees from both the initial
and  goal configurations. This was first proposed in 
\cite{kuffner00rrtconnect}. Our formulation of manipulation planning does
not include an explicit goal configuration, so it is not possible to directly grow
a tree from the goal. To make use of the idea of growing multiple trees, we first
randomly sample the goal submanifold and generate several goal configurations. Then, 
we grow random
trees from the initial configuration and the random goal configurations. The idea of
generating several goals for manipulation planning was proposed in \cite{diankov2008bpc}.

Section~\ref{sec:goal-sampling}  describes the goal manifold sampling, and 
Section~\ref{sec:extension} the adaptation or RRTs random extensions to constrained planning.

\subsection{Goal Manifold Sampling}
\label{sec:goal-sampling}

The way we generate a goal configuration is the following:
\begin{enumerate}
\item Shoot a random configuration in $\mathcal{C}$ with uniform distribution.
\item Call the local planner on this configuration, with the static
  stability constraints and the goal task.
\item Check that all of those tasks are achieved.
\item Check for collisions.
\end{enumerate}

Fig.~\ref{fig:goal} shows different random configurations respecting stability 
constraints and solving a reaching task for an animated character.

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pics/goal.png}

\caption{Random goal configurations solving a reaching task. All the
  configurations are stable and collision-free, and the right hand of the character
  reaches the purple box.}
\label{fig:goal}
\end{figure}

\subsection{Random Extensions on a Constrained Manifold}
\label{sec:extension}

Fig.~\ref{fig:rrt-extend} shows an extension of the classic 
RRT algorithm, from a configuration already in the tree $q_{near}$ towards a random
configuration $q_{rand}$.

\begin{figure}[h]
  \centering

  \begin{tikzpicture}[x=0.5cm,y=0.5cm]

    \node [draw,circle,inner sep=2pt] (0) at (0,0) {};
    \node [draw,circle,inner sep=2pt] (1) at (1,1) {};
    \node [draw,circle,inner sep=2pt] (2) at (2.4,1) {};
    \node [draw,circle,inner sep=2pt,fill=gray] (qn) at (3.4,2) {};
    \node [above left] at (qn) {$q_{near}$};
    \node (t) at (1.5,-0.5) {$\mathcal{T}$};
    \draw (0) -- (1) -- (2) -- (qn) ;

    \node [draw,circle,inner sep=2pt] (qn1) at (4.8,2) {};
    \node [draw,circle,inner sep=2pt] (qn2) at (6.2,2) {};
    \node [draw,circle,inner sep=2pt] (qn3) at (7.6,2) {};

    \draw (qn) -- (qn1) -- (qn2) -- (qn3) ;

    \path [<->,thick,red] (3.4,1.5) edge  node [below,black]
          {$\varepsilon$}  (4.8,1.5) ;
          \path [<->,thick,red] (4.8,1.5) edge  node [below,black]
                {$\varepsilon$}  (6.2,1.5) ;
                \path [<->,thick,red] (6.2,1.5) edge  node [below,black]
                      {$\varepsilon$} (7.6,1.5) ;

                      \node [rectangle] (newconf) at (6.2,3.5) {\begin{small}new configurations\end{small}};

                      \draw [->,blue] (newconf) -- (qn1.north) ;
                      \draw [->,blue] (newconf) -- (qn2.north) ;
                      \draw [->,blue] (newconf) -- (qn3.north) ;

                      \node [draw,circle,inner sep=2pt] (qrand) at (13,2) {};
                      \node [below right] at (qrand) {$q_{rand}$};

                      \draw [dashed,thin] (qn3) -- (qrand) ;

  \end{tikzpicture}

  \caption{One step of extension of the RRT algorithm} 
  \label{fig:rrt-extend}
\end{figure}

The equivalent random  extension on a constrained manifold $\mathcal{M}$
starts from a valid configuration $q_{near} \in \mathcal{M}$, and extends the tree
towards a random configuration $q_{rand}$, while keeping the constraints
satisfied. To do so, we add a task to the local planner, whose value
is the distance to  $q_{rand}$ in $\mathcal{C}$. Algorithm~\ref{alg:constrained}
shows the pseudo-code of a constrained random extension from  $q_{near}$ towards
$q_{rand}$ while satisfying the constraints defining $\mathcal{M}$: $Constraints$.
The configuration space task used to explore $\mathcal{C}$ is 
noted $ConfigurationTask$.   This
task is added with the lowest priority. Each new configuration $q_{new}$ generated by the
local planner  is 
checked for constraint satisfaction and collision 
avoidance. If the new configuration is valid, each new edge is checked for collision
avoidance, as in a classic RRT algorithm.

\begin{algorithm}[h]
  \caption{Constrained-Extend($\mathcal{T},q_{near},q_{rand},Constraints$)}
  \label{alg:constrained}
  \begin{algorithmic}
    \STATE $Tasks.$Initialize$(Constraints)$
    \STATE $ConfigurationTask$.SetTarget($q_{rand}$)
    \STATE $Tasks.$Add$(ConfigurationTask)$
    \STATE $q_{current} \leftarrow q_{near}$
    \STATE $State \leftarrow$ Progressing
    \WHILE{$State =  $ Progressing}
    \STATE $q_{new} \leftarrow $ LocalPlannerPerformOneStep($Tasks,q_{current}$)
    \IF{$q_{new} \neq q_{current}$ \\ 
      and CollisionCheck($q_{new}$) = OK\\ 
      and CollisionCheckEdge($q_{current},q_{new}$) = OK \\
      and ConstraintsCheck($q_{new},Constraints$) = OK }
    \STATE $\mathcal{T}$.AddNode($q_{new}$)
    \STATE $\mathcal{T}$.AddEdge($q_{current},q_{new}$)
    \STATE $q_{current} \leftarrow q_{new}$
    \IF{$q_{target} = q_{rand}$}
    \STATE $State \leftarrow$ Reached
    \ENDIF
    \ELSE 
    \STATE $State \leftarrow$ Trapped
    \ENDIF
    \ENDWHILE

  \end{algorithmic}
\end{algorithm}

Fig. \ref{fig:gikrrt} shows one step of constrained
extension, and the prioritized stack of tasks and constraints solved by the local planner.

\begin{figure}[h]
\centering
  \begin{tikzpicture}[x=0.61cm,y=0.61cm]
    \definecolor{dg}{rgb}{0,0.3,0}
     \draw [dashed,color=dg,thick,text width=2cm] (1,1)
      arc(139:90:10.61) node [above right] {\small{Constrained manifold $\mathcal{M}$}} ; 
      \node [draw,circle,inner sep=2pt,fill=gray] (qnear) at (2,2) {};
      \node [left] (qn0) at (qnear) {$q_{near}$} ;
      \node [draw,circle,inner sep=2pt] (q1) at (3,3) {};
      \node [draw,circle,inner sep=2pt] (q2) at (4,4) {};
      \node [draw,circle,inner sep=2pt] (q3) at (5,5) {};
      \node [draw,circle,inner sep=2pt] (qrand) at (7,7) {};
      \node [below right] (qr0) at (qrand) {$q_{rand}$} ;
      \draw [dashed,thin] (qnear) -- (q1) -- (q2) -- (q3) -- (qrand) ;

    \node [draw,circle,inner sep=2pt,fill=gray] (qc1) at (3.11,2.83) {};
      \draw (qnear) -- (qc1);
      \node [draw,circle,inner sep=2pt,fill=gray] (qc2) at (4.26,3.49) {};
      \draw (qc1) -- (qc2);
     \node [draw,circle,inner sep=2pt,fill=gray] (qc3) at (5.37,3.97) {};
      \draw (qc2) -- (qc3);
      \node [rectangle] (newconf) at (6,2) {\small{new configurations}}; 
      \draw [->,blue] (newconf) -- (qc1) ;
      \draw [->,blue] (newconf) -- (qc2) ;
      \draw [->,blue] (newconf) -- (qc3) ;

      \draw [color=black, thick,fill = white] (15,1) rectangle (17,2);
      \draw [color=black, thick,fill = white] (15,2) rectangle (17,3);
      \draw [color=black, thick,fill = white] (15,3) rectangle (17,5);
      \draw [color=black, thick,fill = white] (15,5) rectangle (17,6);
      \draw [color=black, thick,fill = white] (15,6) rectangle (20,7);
      \draw [color=black, thick,fill = white] (16,1) rectangle (20,6);
      \draw [thick] (16,1) -- (20,1) -- (20,6);
      \draw (16,2) -- (20,2) ;
      \node [text width = 5cm,text centered] at (17.5,6.5) {\large Stack
        of tasks};
      \begin{scriptsize}
        \node at (15.5,5.5) {1};
        \node at (15.5,4) {$\vdots$}; 
        \node at (15.5,2.5) {$n$};
        \node at (15.5,1.5) {$n$+1};

        \node [text width = 2cm,color=dg] at (18.2,4) {\small Constraints defining
          $\mathcal{M}$};
        \node [text width = 3cm,text centered] at (18,1.5)
              { Configuration task towards $q_{rand}$};    
      \end{scriptsize}

\end{tikzpicture}

\caption{One step of constrained extension, and the stack of tasks used
  to generate new configurations with a local planner starting from $q_{near}$:
  first, the constraints defining $\mathcal{M}$, and at a lower priority
  a configuration task towards $q_{rand}$.}
\label{fig:gikrrt}
\end{figure}

\subsection{Example}

An experimental study, presented in  \cite{dalibard09}, showed that
in complex environments, randomized motion planners were faster than
collision avoiding local planners. Another benefit of using global motion
planning compared to local gradient descent is the ability to avoid local 
minima.

We present in Fig.~\ref{fig:wb-shelves} an illustration of 
the use of randomized motion planning 
on complex manipulation problems. The humanoid robot HRP-2, faces shelves. It has to: (i) grasp
a ball lying on a shelf, (ii) put it on a higher shelf, (iii) come back to a natural
rest configuration. The statically stable manifold on which the planning takes place is
defined by tasks constraining the robot feet and CoM. The goal of the first problem
is defined by a hand configuration task (position and orientation), and a gaze task
(the robot has to see the ball). The goal of the second problem is defined by a 
different hand configuration and the same gaze task,
and finally, the goal of the third problem is the rest configuration.
Extension 1 presents a video of this motion.

\begin{figure}[h]
\centering
\includegraphics[width=0.24\linewidth]{pics/wb-shelves/1.jpg}
\includegraphics[width=0.24\linewidth]{pics/wb-shelves/2.jpg}
\includegraphics[width=0.24\linewidth]{pics/wb-shelves/3.jpg}
\includegraphics[width=0.24\linewidth]{pics/wb-shelves/4.jpg}
\\
\vskip 0.08cm
\includegraphics[width=0.24\linewidth]{pics/wb-shelves/5.jpg}
\includegraphics[width=0.24\linewidth]{pics/wb-shelves/6.jpg}
\includegraphics[width=0.24\linewidth]{pics/wb-shelves/7.jpg}
\includegraphics[width=0.24\linewidth]{pics/wb-shelves/8.jpg}

\caption{HRP-2 grabs a ball on a shelf, puts it on another shelf, and comes back to 
  a rest position. Static stability constraints are respected along the path, and 
  the intermediary goals consisting in grasping and displacing the ball are defined
  implicitly as inverse kinematic tasks.}
\label{fig:wb-shelves}
\end{figure}

The motion presented in Fig.~\ref{fig:wb-shelves} looks very complicated.
The reasons are that the environment is complex and HRP-2 kinematic chain is highly constrained.
The motion has been optimized based on these constraints. Note that the
accessibility space of this robot is not wide enough to accomplish this task further 
away from the shelves, which could have ease the motion. One possibility to extend the
robot accessibility space is to automatically change the support polygon to solve the task.
This is an ongoing research field, \cite{OussamaKanoun06102010} proposes some 
local solutions. This has not been integrated in our global planning algorithm yet.

\section{From Statically Stable Paths to Dynamic Walk Trajectories}
\label{sec:wb-step}

The previous section has presented a simple algorithm that solves manipulation planning
problems on a given constrained manifold of $\mathcal{C}$.

If we use this algorithm with static 
stability constraints without fixing globally the robot foot positions, it generates 
statically stable paths for a robot sliding on the ground. Fig~\ref{fig:sliding} shows
an example of a whole-body collision-free path for a robot passing between two chairs.
Such paths are physically unfeasible. They are, however, easier to generate than feasible
dynamic trajectories because only geometric constraints are considered at planning time.

\begin{figure}[h]
  \centering

  \includegraphics[width=0.24\linewidth]{pics/chairs/sliding-perspective-1.png}
  \includegraphics[width=0.24\linewidth]{pics/chairs/sliding-perspective-2.png}
  \includegraphics[width=0.24\linewidth]{pics/chairs/sliding-perspective-3.png}
  \includegraphics[width=0.24\linewidth]{pics/chairs/sliding-perspective-4.png}

  \caption{Collision-free  statically stable path  for a  humanoid  robot  sliding on  the
    ground.}
  \label{fig:sliding}
\end{figure}

This section presents a \textit{constructive} proof that any such statically stable, 
collision-free path for a legged robot sliding
on the ground can be approximated by a collision-free dynamic walk trajectory.
The proof is based on ideas from control theory, in particular small-space 
controllability. It also uses the fact that balanced criteria for dynamic walk are different
from the ones for static stability. An important point is that our result and method do not hold
for a biped robot that walks in a quasi-static way.

Section~\ref{sec:ssc} recalls the definition of small-space controllability and its
use in motion planning. Section~\ref{sec:humanoid-ssc} proves that a legged robot
dynamically walking is small-space controllable, while a legged robot quasi-statically
walking is not. Section~\ref{sec:ssc-application} shows how this property is used to
approximate collision-free statically stable paths by dynamic walk trajectories.

\subsection{Small-Space Controllability}
\label{sec:ssc}

A  robotic system  is  controllable  if  for any two  configurations
$q_1$ and $q_2$,  there exists  a
trajectory  going  from  $q_1$ to  $q_2$.  It  is  
\textit{small-space  controllable} if for  any configuration  $q$, 
for any $\epsilon >0$, there
exist $\eta >0$ such that all the configurations contained in the ball of center
$q$ and radius $\eta$ are reachable by trajectories included in the
ball of center $q$ and radius $\epsilon$.

The main  consequence of  that property  
in  motion planning  is the following theorem, that shows how planning for
dynamic systems is reduced to geometric planning
thanks to small-space controllability:

\begin{theorem}
  \label{thm:ssc}
  Any collision-free path of a small-space controllable system can be approximated
  by a sequence of both collision-free and admissible trajectories. Thus, small-space 
  controllability reduces trajectory planning problems to geometric path planning problems.
\end{theorem}

Fig. \ref{fig:stc} shows an example of collision-free 
path approximation by admissible collision-free sub-trajectories. The fact 
that this algorithm  converges is guaranteed by the small-space
controllability property.

\begin{figure}[h]
  \centering

  \begin{minipage}{3cm}
    \input{stc1.tex}
  \end{minipage}
  \begin{minipage}{4cm}
    \input{stc2.tex}
  \end{minipage}

  \caption{Small-space controllability in motion planning. On the left,
    the local property: any configuration $q'$ at distance less than
    $\eta$ is reachable from $q$ by an admissible trajectory included in
    a ball of size $\epsilon$. On the right, a collision-free path from
    $q_1$ to $q_2$ is approximated by collision-free and admissible
    trajectories by using the local property.
  }
  \label{fig:stc}
\end{figure}

This result has been long known and used in motion planning, in particular in planning
for non-holonomic systems. A detailed proof of it can be found in
\cite{taix-94}. We will present a sketch of the proof to give an intuition about the 
corresponding algorithm.

\begin{proof}
  Let $\mathcal{C}$ be the configuration space of a small-space controllable robot, and 
  $\mathcal{C}_{free} \subset \mathcal{C}$ the set of collision-free configurations. We
  consider in-contact configurations as colliding, so $\mathcal{C}_{free}$ is an open set.
  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. For every $x \in [0,1]$,
  $\tau(x) \in \mathcal{C}_{free}$, so there exists $\epsilon_x$ such that the open ball 
  $B(\tau(x),\epsilon_x)$ of center $\tau(x)$ and radius $\epsilon_x$ is included in 
  $\mathcal{C}_{free}$. The small-space controllability property states that for every $x$,
  there exist $\eta_x$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
  from $\tau(x)$ by a trajectory included in $B(\tau(x),\epsilon_x)$.

  The set of open balls $\left( B(\tau(x),\eta_x) \right)_{x\in [0,1]}$ forms an open cover
  of $\tau([0,1])$ which is compact. The Heine-Borel theorem states that there exists a
  finite subcover $\left( B(\tau(x_i),\eta_{x_i}) \right)_{i\in \{ 1,\dots ,n \}}$ of $\tau([0,1])$. To this
  finite subcover corresponds a finite number of feasible trajectories, going from $\tau(0)$ to  
  $\tau(1)$, included in the union of 
  $\left( B(\tau(x_i),\epsilon_{x_i}) \right)_{i\in \{ 1,\dots ,n \}}$, and thus in 
  $\mathcal{C}_{free}$. This concludes the proof.
\end{proof}

\subsubsection{Small-Time \textit{versus} Small-Space Controllability}
In the control theory literature, the property used is usually \textit{small-time controllability}, 
which states that for  any configuration  $q$, for  any time
$T>0$, the set of configurations accessible from $q$ in time less than
$T$ forms a  neighborhood of $q$. When accelerations and velocities are bounded,
small-time controllability implies small-space controllability, which is why 
a lot of motion planning previous work only refers to the sufficient small-time controllability
property. However, the reciprocate is not necessarily true:  a system can be 
small-space controllable and
not small-time, if the trajectories generated by its controller are arbitrarily long.
The important property
in terms of motion planning application is the small-space controllability, as 
Theorem~\ref{thm:ssc} shows. In the following, we show that legged robots
are small-space controllable, but  not that they are small-time controllable.
In fact, the  control method that we present does not follow the small-time controllability
property. For the sake of clarity, we have chosen to make the distinction between these
two controllability properties.

\subsection{Walking Robot Controllability}
\label{sec:humanoid-ssc}

This section  discusses the small-space controllability of walking robots. To clarify
the presentation, we consider a simplified model of legged robot consisting of two feet
of null mass and a punctual mass free to move in three dimensions.
We do not consider 
the kinematic chains between the feet and the mass. The robot is walking on a flat terrain,
and the feet are assumed to have a positive surface. For our presentation, it is not 
necessary to consider the feet height, so the configuration space of the robot is:
\[
\mathcal{C} = SE(2) \times SE(2) \times \mathbb{R}^3
\]
It is of dimension 9.

The balanced walking conditions for a quasi-static walking robot are that the punctual
mass, or CoM, should always be over the support polygon (the convex hull of the two feet), and 
one foot can move iff the CoM is over the other foot. Similarly, the walking
conditions for a dynamic walking robot are that the ZMP should
always be in the robot support polygon, and on foot can move iff the ZMP is over the other 
foot. For a precise description of dynamically balanced walking conditions, the reader can refer
to \cite{wieber2002}. Under these assumptions, the following result holds:

\begin{theorem}
A quasi-static walking robot is not small-space controllable. A dynamic walking robot is.
\end{theorem}

\begin{proof}

The first claim is straightforward. Let the robot be in a configuration $q$ where the two
feet are separated from a positive distance. Let us assume that the CoM is at a horizontal
distance $L>0$ from the left foot (if the CoM is over the left foot, we can consider similarly 
the right foot). For any $\epsilon < L$, any valid trajectory starting from $q$, included in
the ball of center $q$ and radius $\epsilon$, is such that the CoM is never over the left foot.
Given the quasi-static walk conditions, the right foot of the robot is fixed along 
the trajectory. Thus, the set of accessible configurations from $q$ by staying inside 
$B(q,\epsilon)$ does not form a neighborhood of $q$. This shows that the robot is 
not small-space controllable.

\bigskip

Let us now consider a dynamic walking robot. Starting from any valid static configuration, 
the CoM can move vertically without affecting balance, so there is no need to 
consider this degree of freedom in the following. If the CoM is not over the edge of the support
polygon, it is possible to move it in a quasi-static way inside a neighborhood of its current position that projects itself over the support polygon. 
It is thus sufficient and necessary to prove that for any $\epsilon >0$, it is possible to move
the feet while keeping the CoM inside a neighborhood of size $\epsilon$. Let such $\epsilon >0$
be arbitrarily fixed.

The model of a walking robot with a punctual
mass at a fixed height is known in the literature as the cart-table model \cite{kajita2003biped}.
The equations  giving  the   ZMP  horizontal  coordinates  $(p_x,p_y)$  as
functions  of CoM  horizontal coordinates $(x,y)$  in the  cart-table  model were
presented in \cite{kajita2003biped}:
\begin{equation}
\label{eq:walk-zmp}
\left(
\begin{array}{c}
p_x\\ p_y
\end{array}
\right) = \displaystyle \left(
\begin{array}{c}
x - \frac{z_c}{g} \ddot{x}\\ y - \frac{z_c}{g} \ddot{y}
\end{array}
\right)
\end{equation}
where $z_c$ is  the constant height of the CoM and  $g$ is the gravity
constant.    In    the    following    we    will    note    $\omega_0
=\sqrt{\frac{g}{z_c}}$.

Without loss of generality, let us assume that the robot is in a configuration
in which the CoM is at the horizontal position $(0,0)$, the center of the feet are 
aligned with the $y$-axis and the horizontal distance between the CoM and either of the centers
of the feet is $L$. To achieve dynamically balanced walking, we aim at making
$p_y(t)$ oscillate  between $-L$ and $L$. To move the ZMP  under a
given foot, only  the $y$ coordinate of the CoM  is of interest. Thus,
we will keep the $x$ coordinates  of the CoM and ZMP constant equal to
$0$. We have made the hypothesis that the feet were of positive surface,
let $l>0$ be such that the length of the section of a foot along the 
$y$-axis is greater than $l$.

The idea of this proof is to use the form of Eq. (\ref{eq:walk-zmp}) to
apply a  scaling factor between  the amplitude of the  oscillations of
the CoM and of the ZMP. For $\omega >0$, if the CoM follows a trajectory given
by  $y(t) = \epsilon \sin(\omega t)$,  Eq. (\ref{eq:walk-zmp}) gives:
\[
p_y(t) =
(1+\left(\frac{\omega}{\omega_0}\right)^2)\epsilon\sin(\omega t)
\]

The
amplitude  of  the oscillations  of  $y$  is  multiplied by  a  factor
$(1+\left(\frac{\omega}{\omega_0}\right)^2)$.  If we choose  $\omega =
\omega_0 \sqrt{\frac{L}{\epsilon} -1}$,  $p_y$ oscillates between $-L$
and    $L$.   At    time   $t_l^{(n)}    =    n\frac{2\pi}{\omega}   +
\frac{\pi/2}{\omega}$, the  ZMP is located  at the center of  the left
foot,  the robot  can move  its right  foot and  at time  $t_r^{(n)} =
n\frac{2\pi}{\omega}  + \frac{3\pi/2}{\omega}$ the  ZMP is  located at
the center of the right foot, the robot can move its left foot.

Starting from a static configuration at time $(t=0)$, we cannot apply
directly  a  command  $y(t)  =  \epsilon \sin(\omega  t)$  because  it
generates  a discontinuity  in the  speed of  the CoM  at time $(t=0)$. To
overcome this  discontinuity, we go through a  transient state between
$(t=0)$ and  $(t=T)$ for some  $T >0$. Let  $f:[0,T] \rightarrow
[0,1]$  be an  increasing function of class $C^\infty$  such  that  $f(0)  =  0$,
$\dot{f}(0) = 0$, $f(T) =  1$, $\dot{f}(T) = 0$ and $\ddot{f}(T)
=  0$.  We can explicitly  construct  such  an $f$   with  a degree  4
spline.   We   also   request   that   for  all   $t   \in   [0,T]$,
$|2\epsilon\dot{f}(t)\frac{\omega}{\omega_0^2}|   \leq  \frac{l}{4}$
and   $|\epsilon\ddot{f}(t)/\omega_0^2|  \leq   \frac{l}{4}$.  These
inequalities  will be  used  to bound  the  trajectory of  the  ZMP. We  can
guarantee them by  choosing $T$ large enough. Let  us now consider the
following CoM motion:

\[
y(t) = \left\{
\begin{array}{ll}
f(t)\epsilon\sin(\omega t) 
& \text{if } t\in [0,T]
\\ 
\epsilon\sin(\omega t) 
& \text{if } t \geq T \end{array}
\right.
\]

One can  check that $y$  is of class $C^2$ over  $\mathbb{R}_+$, and
that $\dot{f}(0) = 0$. When $t\geq T$, the robot is in the permanent
state described above  and can successively move its feet  inside small neighborhoods.
The  last point to check
is that for $t \in  [0,T]$ $p_y(t)$ stays inside the support polygon
of the robot. The calculation of the successive derivatives of $y$ gives:

\[
\begin{array}{cl}
p_y(t) = &  f(t) \epsilon (1 + \left(\frac{\omega}{\omega_0}\right)^2)
\sin (\omega  t) \\ &  + 2\epsilon \dot{f}(t)\frac{\omega}{\omega_0^2}
\cos  (\omega t)  \\ &  +  \frac{\epsilon}{\omega_0^2}\ddot{f}(t) \sin
(\omega t)
\end{array}
\]

\begin{figure}
\centering
\input{fig-zmp-inplace.tex}

\caption{CoM motion (in plain red) along $y$ axis.  The CoM stays in the interval
  $[-\epsilon,\epsilon]$ while during  permanent state ($t \geq T$),
  the ZMP (dashed blue) oscillates between the centers of the feet, which allows
  in-place walk.}
\label{fig:zmp-inplace}
\end{figure}

For    all     $t    \in    [0,T]$,    $f(t)     \epsilon    (1    +
\frac{\omega}{\omega_0}^2)  \sin  (\omega t)$  lies  between $-L$  and
$L$. The bounds on the derivatives of $f$ guarantee that $p_y(t)$ lies
between  $-L- l/2$ and  $L+ l/2$,  which means  that the  ZMP stays
inside  the  support  polygon.  Fig.  \ref{fig:zmp-inplace}  shows  an
example  of CoM  motion  on the  $y$  axis and  the corresponding  ZMP
motion. Once in permanent walk in-place state, the robot can come back
to  a static  state by  applying a  symmetric transient  state  used to
decrease  gradually  the amplitude  of  the  oscillations  of the  CoM
without  generating a  discontinuity in  the first  derivative  of the
command.

We have thus exhibited a continuous control scheme that allows to move any of the feet
while keeping the CoM inside an arbitrarily small neighborhood, this concludes
the proof.
\end{proof}

\subsubsection{Remarks}
\paragraph{Generalization to a complete model:} More work is needed to generalize 
the previous proof to any legged robot model. First, because we use a model where
the mass is punctual, we did not have to consider the angular-momentum of the robot
around the CoM, while it accounts for dynamic balance. Moreover, when considering
an actual kinematic and dynamic model,
the positions of feet and CoM are not controlled degrees of freedom, but inputs 
for inverse kinematic tasks. Continuity and regularity results about the robot
controller -- i.e. how much the configuration changes when the input of an inverse
kinematic task changes--, would be needed to generalize the proof. We have not
proven such properties. The  proof presented above still gave the intuition for the
algorithm presented in the following.

\paragraph{Use of ZMP preview controller:} The control strategy presented in the  
previous proof may
generate very  long trajectories, because  of the transient  states at
the beginning and end of the locomotion. In the actual implementation,
we chose  to generate  CoM motions with  a ZMP preview  controller, as
presented in \cite{kajita2003biped}.  We  have observed experimentally that the
amplitude of  CoM trajectories decreases  when the frequency  of steps
increases.

\paragraph{Speed of CoM:} The theoretical result presented in this section implies
that any collision-free path can be approximated by a sequence of admissible 
and collision-free trajectories. However, the theorem depends on a control law
that will generate trajectories with unbounded velocities for the CoM, when the input
path is close to obstacles. The humanoid robot
hardware may be a limitation to such trajectories. To prevent the generated CoM 
oscillations from being too fast, one has to require that the statically stable 
path is included
inside an $\epsilon$-radius tube of the free space, where $\epsilon$ depends on the
physical capacities of the robot.

\subsection{Application: Dynamic Approximation of a Statically Stable Sliding Path}
\label{sec:ssc-application}

The  algorithm   that  animates  a  statically  stable   path  into  a
dynamically stable  walk trajectory has been inspired  by the previous
small-space controllability  proof. Given a statically  stable path $p$,
we  start  by placing  footsteps  corresponding  to  the nominal  walk
pattern  of the  robot. Given the footsteps, we  compute a ZMP trajectory, 
and a preview controller outputs  a corresponding CoM trajectory, using
the cart-table model.
The stack of tasks applied to
the robot is - in decreasing priority order:

\begin{enumerate}

\item Position and orientation of  feet,

\item Horizontal position of the CoM,

\item Height of the CoM,

\item Verticality of the waist,

\item Upper-body configuration task towards corresponding
  configuration  of $p$.

\end{enumerate}

Tasks (1)  and (2) generate a  dynamically stable motion  by using the
simplified cart-table model  and the ZMP criterion. Tasks  (3) and (4)
ensure that the  resulting motion is well described  by the cart-table
model. Task (5)  is used to approximate $p$ as  well as possible given
the walk parameters.

Because it comes at the  lowest priority, task (5) is not necessarily
fulfilled in  the resulting trajectory. Hence,  collisions may appear
when animating $p$, if the resulting trajectory diverges too much from
the initial sliding  path. If so, it is  necessary to approximate more
closely $p$  by a walk  trajectory.  To do  so, we use  the small-space
controllability  property   of  the  system  shown   in  the  previous
section. The way  we use this property is  inspired by similar results
in non-holonomic mobile robot control presented in \cite{taix-94}.

If the animated  trajectory collides with the environment,  we cut the
initial  path   $p$  into  two   sub-paths,  that  we  try   to  animate
recursively. When the  paths to animate are too  short for the robot
nominal  walk parameters, we  accelerate the  steps, and  decrease the
maximum height of  the moving foot. As shown  in previous section, the
walk trajectory  corresponding to  smaller and faster  steps converges
toward the  sliding path.  Algorithm  \ref{alg:walk} shows pseudo-code
that takes  a sliding path $p$  as input and  returns a collision-free
walk trajectory.

\begin{algorithm}[h]
\caption{FindDynamicTrajectory(Path $p$)}
\label{alg:walk}
\begin{algorithmic}
\STATE $Footprints \leftarrow \text{ComputeFootprints}(p)$

\STATE $StackOfTasks$.initialize()

\STATE $StackOfTasks$.addFootprintTask($Footprints$)

\STATE $StackOfTasks$.addWaistTask()

\STATE $StackOfTasks$.addConfigurationTask($p$)

\STATE $DynamicTrajectory \leftarrow \text{Animate}(StackOfTasks)$

\IF{(CheckForCollisions($DynamicTrajectory$) = Colliding)}

\STATE $(p_1,p_2) \leftarrow \text{CutInHalf}(p)$

\STATE $DT_1 \leftarrow \text{FindDynamicTrajectory}(p_1)$

\STATE $DT_2 \leftarrow \text{FindDynamicTrajectory}(p_2)$

\RETURN $\text{Concatenate}(DT_1,DT_2)$

\ELSE

\RETURN $DynamicTrajectory$

\ENDIF
\end{algorithmic}
\end{algorithm}

\section{Experimental Results}

\label{sec:exp}

The motion planning algorithms presented in this paper have been implemented
using KineoWorks\texttrademark \cite{laumond2006kcs}. The planning times have been measured
on an Intel Core~2~Duo 2.13~GHz PC with 2~GB of RAM. Evaluation of the
randomized algorithm has been conducted by executing 50 trials on each
of the first two problems. We present the average results of these trials.

\subsection{Passing between two chairs}

The environment shown in Fig. \ref{fig:couv} and \ref{fig:sliding} was
presented in \cite{el2011path}. The  authors solved it by using a bounding
box method, leading the robot to walk sideways between the two chairs.
Our method generated a locomotion  trajectory in which the robot walks
forward, which could be required if the robot has to use vision during
locomotion for  example. The first planning stage  required 29.6~s
on  average.   The  animation  of   the  sliding  path   presented  in
Fig.   \ref{fig:sliding}    used   66.5~s  of  computation   time.

\begin{figure}[h!]
\centering
\includegraphics[width=0.7\linewidth]{pics/chairs/waist-trajectory.png}

\caption{Horizontal trajectory of the robot CoM during
  locomotion. When the robot is close to obstacles, the amplitude of
  the oscillations decreases.}
\label{fig:chairs-waist}
\end{figure}

Fig.  \ref{fig:chairs-waist} shows  the horizontal  trajectory  of the
robot CoM  during  locomotion. The amplitude of the oscillations  decreases when passing
between the  chairs.  This motion has  been validated on  a real HRP-2
platform. Extension 2 shows a video of the experiment.

\subsection{Walking among floating obstacles}

\begin{figure}[h!]

\centering

\includegraphics[width=0.24\linewidth]{pics/objects-cloud/perspective-1.png}
\includegraphics[width=0.24\linewidth]{pics/objects-cloud/perspective-2.png}
\includegraphics[width=0.24\linewidth]{pics/objects-cloud/perspective-3.png}
\includegraphics[width=0.24\linewidth]{pics/objects-cloud/perspective-4.png}
\\ 
\vskip 0.08cm
\includegraphics[width=0.24\linewidth]{pics/objects-cloud/perspective-5.png}
\includegraphics[width=0.24\linewidth]{pics/objects-cloud/perspective-6.png}
\includegraphics[width=0.24\linewidth]{pics/objects-cloud/perspective-7.png}
\includegraphics[width=0.24\linewidth]{pics/objects-cloud/perspective-8.png}

\caption{Solution path for a cluttered environment, the robot walks
  among floating obstacles.}
\label{fig:cluttered}
\end{figure}

In the environment shown in Fig. \ref{fig:cluttered}, the robot
has to find a way among floating obstacles. In this
environment neither bounding box nor footstep planning strategies
could find a collision-free walk trajectory.
The first planning stage required
184.3~s on average, and the animation of the trajectory presented in 
Fig.~\ref{fig:cluttered} used 339.5~s of computation time. Fig.~\ref{fig:cluttered-waist} 
shows the robot CoM trajectory during locomotion.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.7\linewidth]{pics/objects-cloud/waist-trajectory.png}

  \caption{Horizontal trajectory of the robot CoM during
    locomotion.}
  \label{fig:cluttered-waist} 
\end{figure}

\subsection{'Put the ball on a shelf'}

In the problem shown in Fig. \ref{fig:shelf} the robot has to put a ball on a
shelf, in a constrained apartment environment. The final configuration is defined 
implicitly as a desired hand position. We generated automatically goal configurations 
solving the task, as described in Section~\ref{sec:wb} . Then, we 
applied our planner to generate a whole-body walk motion that solved the hand reaching
task.

The solution sliding path is constrained between the table on the right and the lamp
on the left. This passage is too narrow for the robot nominal walk parameters. 
When executing the walk motion resulting from our algorithm, the robot left hand
is only a few centimeters away from the lamp.

Fig.~\ref{fig:shelf} shows the resulting walk motion and Fig.~\ref{fig:shelf-waist} 
shows the robot CoM trajectory during locomotion. Extension 3 presents a 
comprehensive video of this problem, including the motion execution on
the real robot HRP-2.

\begin{figure}[h]
\centering
\includegraphics[width=0.24\linewidth]{pics/shelves/trajectory-1.png}
\includegraphics[width=0.24\linewidth]{pics/shelves/trajectory-2.png}
\includegraphics[width=0.24\linewidth]{pics/shelves/trajectory-3.png}
\includegraphics[width=0.24\linewidth]{pics/shelves/trajectory-4.png}
\\ 
\vskip 0.1cm
\includegraphics[width=0.24\linewidth]{pics/shelves/trajectory-5.png}
\includegraphics[width=0.24\linewidth]{pics/shelves/trajectory-6.png}
\includegraphics[width=0.24\linewidth]{pics/shelves/trajectory-7.png}
\includegraphics[width=0.24\linewidth]{pics/shelves/trajectory-8.png}

\caption{Solution path for a hand reaching problem in an
  appartment. The goal is implicitly defined as an inverse kinematics
  task.} 
\label{fig:shelf}
\end{figure}

\begin{figure}[h]
\centering

\includegraphics[width=0.5\linewidth]{pics/shelves/waist-trajectory.png}

\caption{Horizontal trajectory of the robot CoM during
    locomotion.}
\label{fig:shelf-waist}
\end{figure}

\section{Limits and Discussions}
\label{sec:limits}

This section lists several limitations of the current methods, and discusses
potential future work to overcome them.

\subsection{Stepping over obstacles}

Because of the kinematic constraints we apply at the planning stage, we are not
able yet to plan motions where the robot steps over obstacles, while this is an 
important feature of humanoid robots. Nevertheless, because we compute collision queries on
an exact model of the robot, our method is able to generate paths where obstacles
pass between the feet of the robot.
In future work, we plan to develop mixed
methods, where collision avoidance at the leg level can be solved by footstep
planning techniques, while whole-body collision-avoidance can be solved by the
algorithm presented in this paper.

\subsection{Environment representation}

The presented experiment setup assumes perfect knowledge of the environment. This can
be guaranteed during experiments by using calibrated objects and motion capture systems.
This allows us to focus on complex motion planning problems. The perception problem,
interesting as it is, is completely decoupled from the planning problem in our
presentation.

\subsection{Trajectory following}

The presented setup also assumes perfect execution of the plan. It can be 
critical here, since
non-nominal stepping may cause to drift away from the planned trajectory. Future 
experiments will  include trajectory following during plan execution.

\section{Conclusion}

In this paper, we have presented a simple algorithm for constrained motion planning
and used it within a novel, well-grounded strategy for humanoid whole-body
manipulation planning including locomotion. The locomotion algorithm is based on a formal
small-space controllability property of humanoid robots. An important point is that
this strategy only holds for dynamic walking robots, and not for quasi-static walking ones.
We have used our motion planner on different challenging examples, and validated the
generated motions on a real platform. We have discussed the limits and potential extensions
of our method, and we plan to address them in future work.

\bibliographystyle{agsm}

\bibliography{bibli}

\end{document}

### 2012-06-14 stc1.tex
Add auxiliary files

\definecolor{cc6c6c6}{RGB}{198,198,198}
\definecolor{cffffff}{RGB}{255,255,255}

\begin{tikzpicture}[y=0.4pt, x=0.4pt,yscale=-1, inner sep=0pt, outer sep=0pt]
% path5273
\path[shift={(-0.42857142,4.7142857)},draw=black,miter limit=4.00,line
  width=0.800pt]
  (365.7143,269.5050)arc(0:180:97)arc(-180:0:97) --
  cycle;

% path5275
\path[shift={(-6.0609153,-3.535534)},draw=black,fill=cc6c6c6,miter
  limit=4.00,line width=0.800pt]
  (323.7539,278.0803)arc(-0:180:51)arc(-180:0:51) --
  cycle;

% path5315
\path[draw=black,line join=miter,line cap=butt,miter limit=4.00,line
  width=1.200pt] (265.6701,275.0498) .. controls (265.6701,275.0498) and
  (165.7228,265.2948) .. (178.2919,306.8696) .. controls (185.4208,330.4496) and
  (285.8732,311.9204) .. (285.8732,311.9204) -- (285.8732,311.9204);

% path4761
\path[draw=black,fill=cffffff,miter limit=4.00,line width=0.400pt]
  (278.8021,274.5447)arc(0:180:12)arc(-180:0:12) --
  cycle;

% path4761-4
\path[shift={(17.677674,33.840124)},draw=black,fill=black,miter limit=4.00,line
  width=0.400pt]
  (278.8021,274.5447)arc(-0:180:12)arc(-180:0:12) --
  cycle;

% text5307
\path[fill=black] (288,334.92981) node[right] (text5307) {$q'$};

% text5311
\path[fill=black] (280,286.9537) node[above right] (text5311) {$q$};

% path5319
\path[draw=black,fill=black,line join=miter,line cap=butt,line width=0.800pt]
  (201.0204,312.4254) -- (220.7183,318.9914) -- (200.5153,323.0320) --
  (201.0204,312.4254) -- cycle;

% path2829
\path[draw=black,line join=miter,line cap=butt,line width=0.800pt,<->]
  (273.5714,264.5050) -- (320.0,191.6479);

% path2831
\path[draw=black,line join=miter,line cap=butt,line width=0.800pt,<->]
  (254.2857,270.2193) -- (226.4286,245.2193);

% text2833
\path[fill=black] (302.14285,231.6479) node[above right] (text2833) {$\epsilon$};

% text2837
\path[fill=black] (242.14285,258.07648) node[above right] (text2837) {$\eta$};

\end{tikzpicture}

### 2012-06-15 humanoid.tex
Add a picture to the small space controllability proof. Minor corrections

\begin{tikzpicture}[y=0.55pt, x=0.55pt,yscale=-1, inner sep=0pt, outer sep=0pt]
  \path[cm={{0.91292191,0.40813428,-0.40813428,0.91292191,(116.6153,-98.710192)}},draw=black,fill=black,miter
  limit=4.00,fill opacity=0.621,line width=2.400pt]
  (373.7564,122.0117)arc(0.000:180.000:58.841385 and
  23.739)arc(-180.000:0.000:58.841385 and 23.739) -- cycle;
\path[cm={{0.91292191,0.40813428,-0.40813428,0.91292191,(-6.875857,3.3152152)}},draw=black,fill=black,miter
  limit=4.00,fill opacity=0.621,line width=2.400pt]
  (373.7564,122.0117)arc(0.000:180.000:58.841385 and
  23.739)arc(-180.000:0.000:58.841385 and 23.739) -- cycle;
\path[shift={(26.616754,-41.712121)},draw=black,fill=black,miter limit=4.00,fill
  opacity=0.621,line width=2.400pt] (275.7716,94.7376)arc(0.000:180.000:8.838835
  and 8.586)arc(-180.000:0.000:8.838835 and 8.586) -- cycle;
\path[draw=black,dash pattern=on 1.60pt off 0.80pt,line join=miter,line
  cap=butt,miter limit=4.00,line width=0.800pt] (187.8884,281.1107) --
(402.9370,93.8920) node [right=5] {$y$-axis};
\path[draw=black,line join=miter,line cap=butt,line width=0.870pt]
  (284.9572,56.9063) -- (202.5850,114.7252) -- (254.4490,172.5441) --
  (232.6622,243.4361);
\path[draw=black,line join=miter,line cap=butt,line width=0.722pt]
  (302.9315,55.9207) -- (371.2797,74.1552) -- (344.6998,94.3611) --
  (353.6560,136.7942);
\path[fill=black] (245,42.362179) node[above right] (text3828) {CoM};
\path[draw=black,dash pattern=on 1.60pt off 0.80pt,line join=miter,line
  cap=butt,miter limit=4.00,line width=0.800pt] (292.5714,63.0765) --
  (292.5714,219.2525);
\path[fill=black] (270,136.6479) node[above right] (text3842) {$z_c$};
\path[draw=black,line join=miter,line cap=butt,line width=0.800pt, style=<->]
(299.6429,281.2908) -- (361.7857,227.3622); 
\node(L) at (335,260) [right] {$L$};
\path[draw=black,line join=miter,line cap=butt,line width=0.800pt,style=<->]
  (339.2857,155.9336) -- (372.5000,127.0050);
\node (l0) at (365,145) {$l$};

\end{tikzpicture}

### 2012-06-15 paper.tex
Add Nishiwaki footstep reference
Add a picture to the small space controllability proof. Minor corrections

@@
 Obstacle avoidance can be taken into account with similar methods. To
-do so, one has to include the obstacles as other task constraints to
-be satisfied. A recent  contribution on this subject is \cite{kanehiro2008lca}.
+do so, one has to include the obstacles as  constraints to
+satisfy, see for example \cite{kanehiro2008lca}.
 These methods are prone to fall into local minima, global motion 
@@
 Simplifying  the environment  consists in  considering obstacles  at a
-footstep   level.   \cite{chestnutt2005footstep,kuffner2005motion}
-use   an  
-A$^{*}$   algorithm  to   find
+footstep   level only.   \cite{kuffner2001footstep,chestnutt2005footstep,kuffner2005motion}
+use   an  A$^{*}$   algorithm  to   find
 collision-free   footsteps.    In   \cite{perrinbiped},   the   authors   compute
@@
 constrained sampling-based motion 
-planner and  computes a collision-free path  for a statically stable robot on the
-ground. Another  contribution of this  paper is the formal  proof that
+planner and  computes a collision-free  statically stable path  for a robot fixed or sliding 
+on the ground. Another  contribution of this  paper is the formal  proof that
 this first path can always be  approximated by a dynamically balanced, collision-free
@@
 The problem solved here
-differs from classic configuration approaches in two ways:
+differs from classic configuration space approaches in two ways:
 \begin{enumerate}
@@
 the use of randomized motion planning 
-on complex manipulation problems. The humanoid robot HRP-2, faces shelves. It has to: (i) grasp
+on complex manipulation problems. The humanoid robot HRP-2 faces shelves. It has to: (i) grasp
 a ball lying on a shelf, (ii) put it on a higher shelf, (iii) come back to a natural
@@
 accessibility space of this robot is not wide enough to accomplish this task further 
-away from the shelves, which could have ease the motion. One possibility to extend the
+away from the shelves, which could have eased the motion. One possibility to extend the
 robot accessibility space is to automatically change the support polygon to solve the task.
@@
 $q$ and radius $\eta$ are reachable by trajectories included in the
-ball of center $q$ and radius $\epsilon$.
+ball of center $q$ and radius $\epsilon$. Fig.~\ref{fig:ssc1} shows an illustration
+of this property.
+
+\begin{figure}[h]
+  \centering
+  \input{stc1.tex}
+
+  \caption{The small-space controllability local property:  any configuration $q'$ 
+    at distance less than
+    $\eta$ is reachable from $q$ by an admissible trajectory included in
+    a ball of size $\epsilon$.}
+  \label{fig:ssc1}
+\end{figure}
 
@@
 
-Fig. \ref{fig:stc} shows an example of collision-free 
+Fig. \ref{fig:ssc2} shows an example of collision-free 
 path approximation by admissible collision-free sub-trajectories. The fact 
@@
 
-  \begin{minipage}{3cm}
-    \input{stc1.tex}
-  \end{minipage}
-  \begin{minipage}{4cm}
   \input{stc2.tex}
-  \end{minipage}
-
-  \caption{Small-space controllability in motion planning. On the left,
-    the local property: any configuration $q'$ at distance less than
-    $\eta$ is reachable from $q$ by an admissible trajectory included in
-    a ball of size $\epsilon$. On the right, a collision-free path from
+
+  \caption{Small-space controllability in motion planning. 
+    A collision-free path from
     $q_1$ to $q_2$ is approximated by collision-free and admissible
@@
   }
-  \label{fig:stc}
+  \label{fig:ssc2}
 \end{figure}
@@
 
-\begin{proof}
+\begin{proof}[Proof of Theorem \ref{thm:ssc}]
   Let $\mathcal{C}$ be the configuration space of a small-space controllable robot, and 
@@
   $\mathcal{C}_{free}$. The small-space controllability property states that for every $x$,
-  there exist $\eta_x$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
+  there exist $\eta_x > 0$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
   from $\tau(x)$ by a trajectory included in $B(\tau(x),\epsilon_x)$.
@@
 \begin{theorem}
+\label{thm:humanoid-ssc}
 A quasi-static walking robot is not small-space controllable. A dynamic walking robot is.
@@
 
-\begin{proof}
+\begin{proof}[Proof of Theorem \ref{thm:humanoid-ssc}]
 
 The first claim is straightforward. Let the robot be in a configuration $q$ where the two
-feet are separated from a positive distance. Let us assume that the CoM is at a horizontal
-distance $L>0$ from the left foot (if the CoM is over the left foot, we can consider similarly 
+feet are separated from a positive distance. Let  $L>0$ be the positive horizontal distance
+between the CoM and the left foot (if the CoM is over the left foot, we can consider similarly 
 the right foot). For any $\epsilon < L$, any valid trajectory starting from $q$, included in
@@
 the trajectory. Thus, the set of accessible configurations from $q$ by staying inside 
-$B(q,\epsilon)$ does not form a neighborhood of $q$. This shows that the robot is 
-not small-space controllable.
+$B(q,\epsilon)$ does not form a neighborhood of $q$, since it does not include any configuration
+where the right foot has moved. This shows that the robot is not small-space controllable.
 
@@
 we will keep the $x$ coordinates  of the CoM and ZMP constant equal to
-$0$. We have made the hypothesis that the feet were of positive surface,
+$0$. By hypothesis, the feet are of positive surface,
 let $l>0$ be such that the length of the section of a foot along the 
-$y$-axis is greater than $l$.
+$y$-axis is greater than $l$. Fig.~\ref{fig:simple-humanoid} summarizes the notations used
+in the following.
+
+\begin{figure}[h]
+  \centering
+  \input{humanoid.tex}
+
+  \caption{Simplified model of a legged robot. The CoM is at $(0,0,z_c)$, the two feet 
+    are flat on the ground, aligned with the $y$-axis, at a horizontal distance $L$ 
+    from the CoM.}
+  \label{fig:simple-humanoid}
+\end{figure}
 
@@
 apply a  scaling factor between  the amplitude of the  oscillations of
-the CoM and of the ZMP. For $\omega >0$, if the CoM follows a trajectory given
-by  $y(t) = \epsilon \sin(\omega t)$,  Eq. (\ref{eq:walk-zmp}) gives:
+the CoM and of the ZMP. For $\omega >0$, assuming the CoM follows the trajectory
+$y(t) = \epsilon \sin(\omega t)$,  Eq. (\ref{eq:walk-zmp}) gives:
 \[
@@
 amplitude  of  the oscillations  of  $y$  is  multiplied by  a  factor
-$(1+\left(\frac{\omega}{\omega_0}\right)^2)$.  If we choose  $\omega =
-\omega_0 \sqrt{\frac{L}{\epsilon} -1}$,  $p_y$ oscillates between $-L$
-and    $L$.   At    time   $t_l^{(n)}    =    n\frac{2\pi}{\omega}   +
+$(1+\left(\frac{\omega}{\omega_0}\right)^2)$.  Choosing  $\omega =
+\omega_0 \sqrt{\frac{L}{\epsilon} -1}$ makes  $p_y$ oscillates between $-L$
+and    $L$, while $y$ oscillate between $-\epsilon$ and $\epsilon$.   
+At    time   $t_l^{(n)}    =    n\frac{2\pi}{\omega}   +
 \frac{\pi/2}{\omega}$, the  ZMP is located  at the center of  the left
@@
 that $\dot{f}(0) = 0$. When $t\geq T$, the robot is in the permanent
-state described above  and can successively move its feet  inside small neighborhoods.
+state described above  and can successively move either of its 
+feet  inside small neighborhoods.
 The  last point to check
@@
 We have thus exhibited a continuous control scheme that allows to move any of the feet
-while keeping the CoM inside an arbitrarily small neighborhood, this concludes
+in any direction, while keeping the CoM inside an arbitrarily small neighborhood. This concludes
 the proof.
@@
 controller -- i.e. how much the configuration changes when the input of an inverse
-kinematic task changes--, would be needed to generalize the proof. We have not
-proven such properties. The  proof presented above still gave the intuition for the
+kinematics task changes--, would be needed to generalize the proof. We have not
+proven such properties for the inverse kinematics controller we use. 
+The  proof presented above still gives the intuition for the
 algorithm presented in the following.
@@
 the beginning and end of the locomotion. In the actual implementation,
-we chose  to generate  CoM motions with  a ZMP preview  controller, as
+we have chosen  to generate  CoM motions with  a ZMP preview  controller, as
 presented in \cite{kajita2003biped}.  We  have observed experimentally that the
 amplitude of  CoM trajectories decreases  when the frequency  of steps
-increases.
+increases. Our current ZMP preview controller relies on the cart-table simplified model.
+To make this approximation valid, we fix the height of the robot CoM during walk,
+as well as the vertical orientation of the robot waist. These geometric constraints 
+are also applied when planning a statically stable path. Note that this is
+due to our current implementation of ZMP preview controller, and does not affect
+the generality of the small-space controllability result presented above.
 
@@
 and collision-free trajectories. However, the theorem depends on a control law
-that will generate trajectories with unbounded velocities for the CoM, when the input
+that generates trajectories with unbounded velocities for the CoM, when the input
 path is close to obstacles. The humanoid robot
@@
 dynamically stable  walk trajectory has been inspired  by the previous
-small-space controllability  proof. Given a statically  stable path $p$,
+small-space controllability  proof.
+Given a statically  stable path $p$,
 we  start  by placing  footsteps  corresponding  to  the nominal  walk
@@
 and a preview controller outputs  a corresponding CoM trajectory, using
-the cart-table model.
-The stack of tasks applied to
-the robot is - in decreasing priority order:
+the cart-table model. The stack of tasks applied to
+the robot to generate a dynamic walk motion is -- in decreasing priority order:
 
@@
 
-\item Position and orientation of  feet,
+\item Positions and orientations of  feet,
 

### 2012-06-15 paper.tex
Fix minor typos, modify some mathematical terms.

@@
 When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
-difficulty  is to  take into  account stability  constraints,  i.e. to
+is to  take into  account stability  constraints,  i.e. to
 generate  random  configurations   on  zero  volume  submanifolds  of
@@
   set of inverse kinematic tasks;
-\item the goal is also defined implicitely, by another set of inverse kinematic tasks.
+\item the goal is also defined implicitly, by another set of inverse kinematic tasks.
 \end{enumerate}
@@
   uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
-  This will be detailed in the following of the section.
+  This will be detailed in the following section.
 \end{itemize}
@@
 The motion presented in Fig.~\ref{fig:wb-shelves} looks very complicated.
-The reasons are that the environment is complex and HRP-2 kinematic chain is highly constrained.
+The reasons are that the environment is complex and HRP-2 kinematic tree is highly constrained.
 The motion has been optimized based on these constraints. Note that the
@@
 away from the shelves, which could have eased the motion. One possibility to extend the
-robot accessibility space is to automatically change the support polygon to solve the task.
+robot accessibility space is to automatically change the support polygon durint the motion to solve the task.
 This is an ongoing research field, \cite{OussamaKanoun06102010} proposes some 
@@
 The proof is based on ideas from control theory, in particular small-space 
-controllability. It also uses the fact that balanced criteria for dynamic walk are different
+controllability. It also uses the fact that balance criteria for dynamic walk are different
 from the ones for static stability. An important point is that our result and method do not hold
@@
 trajectory  going  from  $q_1$ to  $q_2$.  It  is  
-\textit{small-space  controllable} if for  any configuration  $q$, 
-for any $\epsilon >0$, there
-exist $\eta >0$ such that all the configurations contained in the ball of center
+\textit{small-space  controllable} if for all configurations  $q$, 
+for all $\epsilon >0$, there
+exists $\eta >0$ such that all the configurations contained in the ball of center
 $q$ and radius $\eta$ are reachable by trajectories included in the
@@
   consider in-contact configurations as colliding, so $\mathcal{C}_{free}$ is an open set.
-  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. For every $x \in [0,1]$,
+  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. For all $x \in [0,1]$,
   $\tau(x) \in \mathcal{C}_{free}$, so there exists $\epsilon_x$ such that the open ball 
   $B(\tau(x),\epsilon_x)$ of center $\tau(x)$ and radius $\epsilon_x$ is included in 
-  $\mathcal{C}_{free}$. The small-space controllability property states that for every $x$,
-  there exist $\eta_x > 0$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
+  $\mathcal{C}_{free}$. The small-space controllability property states that for all $x$,
+  there exists $\eta_x > 0$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
   from $\tau(x)$ by a trajectory included in $B(\tau(x),\epsilon_x)$.
@@
 In the control theory literature, the property used is usually \textit{small-time controllability}, 
-which states that for  any configuration  $q$, for  any time
+which states that for  all configurations  $q$, for  all times
 $T>0$, the set of configurations accessible from $q$ in time less than
@@
 This section  discusses the small-space controllability of walking robots. To clarify
-the presentation, we consider a simplified model of legged robot consisting of two feet
+the presentation, we consider a simplified model of a legged robot consisting of two feet
 of null mass and a punctual mass free to move in three dimensions.
@@
 conditions for a dynamic walking robot are that the ZMP should
-always be in the robot support polygon, and on foot can move iff the ZMP is over the other 
+always be in the robot support polygon, and one foot can move iff the ZMP is over the other 
 foot. For a precise description of dynamically balanced walking conditions, the reader can refer
@@
 The first claim is straightforward. Let the robot be in a configuration $q$ where the two
-feet are separated from a positive distance. Let  $L>0$ be the positive horizontal distance
+feet are separated by a positive distance. Let  $L>0$ be the positive horizontal distance
 between the CoM and the left foot (if the CoM is over the left foot, we can consider similarly 
-the right foot). For any $\epsilon < L$, any valid trajectory starting from $q$, included in
+the right foot). For all $\epsilon < L$, any valid trajectory starting from $q$, included in
 the ball of center $q$ and radius $\epsilon$, is such that the CoM is never over the left foot.
@@
 polygon, it is possible to move it in a quasi-static way inside a neighborhood of its current position that projects itself over the support polygon.
-It is thus sufficient and necessary to prove that for any $\epsilon >0$, it is possible to move
+It is thus sufficient and necessary to prove that for all $\epsilon >0$, it is possible to move
 the feet while keeping the CoM inside a neighborhood of size $\epsilon$. Let such $\epsilon >0$

### 2012-06-15 paper.tex
Bibliography correction

@@
 planning and walk pattern generators based on the Zero-Momentum
-Point (ZMP) formalism. This section summarizes the related work for each of them.
+Point (ZMP) formalism. This section summarizes the literature related to each of 
+these fields.
 
@@
 \cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
-high-dimensional   planning  problems.
+high-dimensional   planning  problems, see \cite{Lav06} and \cite{choset2005prm} for comprehensive
+overviews.
 When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
@@
 use   an  A$^{*}$   algorithm  to   find
-collision-free   footsteps.    In   \cite{perrinbiped},   the   authors   compute
-collision-free motions  for the legs by using  an RRT$^{*}$ algorithm.
+collision-free   footsteps.    In   \cite{perrin2012fast},   the   authors   compute
+dynamic walking motions avoiding collisions at the leg level
+by using  an RRT algorithm.
 
@@
 a growth of algorithmic complexity, particularly costly for high-dimensional
-systems such as humanoid robots. In our work, we show that it is not always
+systems such as humanoid robots. To the authors knowledge such techniques have not been
+used on humanoid robotic platforms so far.
+In our work, we show that it is not always
 necessary to take into account dynamics while planning, even though the final
-output of our algorithm is a dynamically stable walk motion.
+output of our algorithm is a dynamically stable walk motion. This result was first
+presented in \cite{dalibard2011small}.
 
@@
 apply a  scaling factor between  the amplitude of the  oscillations of
-the CoM and of the ZMP. For $\omega >0$, assuming the CoM follows the trajectory
+the CoM and of the ZMP. The faster the CoM oscillates, the bigger is the amplitude
+of the ZMP oscillations. Following is a formalization of this  idea.
+
+For $\omega >0$, assuming the CoM follows the trajectory
 $y(t) = \epsilon \sin(\omega t)$,  Eq. (\ref{eq:walk-zmp}) gives:
@@
 amplitude of  CoM trajectories decreases  when the frequency  of steps
-increases. Our current ZMP preview controller relies on the cart-table simplified model.
+increases. Our current ZMP preview controller relies on the cart-table model approximation.
 To make this approximation valid, we fix the height of the robot CoM during walk,
 as well as the vertical orientation of the robot waist. These geometric constraints 
-are also applied when planning a statically stable path. Note that this is
-due to our current implementation of ZMP preview controller, and does not affect
+are also applied when planning statically stable paths, to ensure that the paths
+can be approximated by dynamic walk trajectories. Note that this is
+due to our current  ZMP preview controller implementation, and does not affect
 the generality of the small-space controllability result presented above.
@@
 small-space controllability  proof.
-Given a statically  stable path $p$,
+Given a statically  stable path $p$ respecting the cart-table model approximation constraints,
 we  start  by placing  footsteps  corresponding  to  the nominal  walk
 pattern  of the  robot. Given the footsteps, we  compute a ZMP trajectory, 
-and a preview controller outputs  a corresponding CoM trajectory, using
-the cart-table model. The stack of tasks applied to
+and a preview controller outputs  a corresponding CoM trajectory. 
+The stack of tasks applied to
 the robot to generate a dynamic walk motion is -- in decreasing priority order:
@@
 
-\item Upper-body configuration task towards corresponding
-  configuration  of $p$.
+\item Configuration task towards corresponding
+  configuration  in $p$.
 
@@
 
-The environment shown in Fig. \ref{fig:couv} and \ref{fig:sliding} was
-presented in \cite{el2011path}. The  authors solved it by using a bounding
+The environment shown in Fig. \ref{fig:couv} and \ref{fig:sliding} has been
+first presented in \cite{el2011path}. It was solved by using a bounding
 box method, leading the robot to walk sideways between the two chairs.
-Our method generated a locomotion  trajectory in which the robot walks
-forward, which could be required if the robot has to use vision during
-locomotion for  example. The first planning stage  required 29.6~s
+Our method generates a locomotion  trajectory in which the robot walks
+forward, which may be required if the robot has to use vision during
+locomotion. The first planning stage  requires 29.6~s
 on  average.   The  animation  of   the  sliding  path   presented  in
-Fig.   \ref{fig:sliding}    used   66.5~s  of  computation   time.
+Fig.   \ref{fig:sliding}   uses  66.5~s  of  computation   time.
 
@@
 could find a collision-free walk trajectory.
-The first planning stage required
+The first planning stage requires
 184.3~s on average, and the animation of the trajectory presented in 
-Fig.~\ref{fig:cluttered} used 339.5~s of computation time. Fig.~\ref{fig:cluttered-waist} 
+Fig.~\ref{fig:cluttered} uses 339.5~s of computation time. Fig.~\ref{fig:cluttered-waist} 
 shows the robot CoM trajectory during locomotion.
@@
 shelf, in a constrained apartment environment. The final configuration is defined 
-implicitly as a desired hand position. We generated automatically goal configurations 
-solving the task, as described in Section~\ref{sec:wb} . Then, we 
-applied our planner to generate a whole-body walk motion that solved the hand reaching
+implicitly as a desired hand position. We have generated automatically goal configurations 
+solving the task, as described in Section~\ref{sec:wb} . Then, we have
+applied our planner to generate a whole-body walk motion that solves the hand reaching
 task.

### 2012-06-15 paper.tex
Some preliminary modifications
  1. fix a few typos,
  2. describe numerical constraint solver,
  3. Use pictures of HRP-2 to illustrate random goal configuration generation.

@@
 planning is needed to overcome this limitation.
+\cite{TouGieGoe2007} Propose a motion generation method where tasks follow
+trajectories define by cubic B-splines. The whole-body motion is optimized
+with respect to the control point positions. This method can take into account
+collision with simple obstacles.
 
@@
 complexity  of a  humanoid  robotics  kinematic tree  is  to use  the
-so-called "cart-table" simplified model \cite{kajita2003biped}. Based on such a
+so-called "cart-table" simplified model \cite{kajita2003biped}. Based this
 model,  planning a trajectory  for the  ZMP is  reduced to  planning a
@@
 Collision-free  locomotion   trajectories  are  usually   obtained  by
-simplifying the model of the  robot or of its environment. By reducing 
-a  humanoid robot to a bounding volume, that includes its swaying motions,
+simplifying the model of the  robot or of the environment. By reducing 
+a  humanoid robot to a bounding volume that wraps the swaying motion,
 one can  use a simple  planar motion planner  on this bounding  volume and
@@
 this first path can always be  approximated by a dynamically balanced, collision-free
-walking  trajectory in a second stage. We have implemented this well-grounded 
-method and  used it on a model of  HRP-2 robot.  The results have been  
-validated on a real platform.
+walking  trajectory. We have implemented this well-grounded 
+method.  The results have been validated on a real HRP-2 robot.
 
@@
 The problem solved here
-differs from classic configuration space approaches in two ways:
+differs from classic approaches in two ways:
 \begin{enumerate}
 \item the set of valid configurations is defined implicitly, as the ones satisfying a given 
-  set of inverse kinematic tasks;
-\item the goal is also defined implicitly, by another set of inverse kinematic tasks.
+  set of inverse kinematics and balance tasks;
+\item the goal is also defined implicitely, by additional inverse kinematic tasks.
 \end{enumerate}
-In this presentation, we use a prioritized inverse kinematic solver, or local planner,
- as a black box. We will not describe the details of its implementation. They are presented in 
-\cite{yoshida2006tds}, and are out of scope of the present work.
-
 During global planning, we will consider several types of
@@
 
-The architecture of the constrained planning algorithm presented here follows
-the one of randomized tree expansion strategies, introduced in both \cite{HsuLat99c} and
-\cite{kuffner00rrtconnect}. Next subsection will recall the structure of the RRT algorithm,
+Our algorithm generalize randomized tree expansion strategies, introduced in both \cite{HsuLat99c} and
+\cite{kuffner00rrtconnect} to constrained motion planning problems. Next subsection will recall the structure of the RRT algorithm,
 a popular randomized motion planning algorithm.
@@
 The classic RRT algorithm, as presented in  \cite{kuffner00rrtconnect}, grows 
-a random tree inside a robot 
+a random tree inside the robot 
 collision-free configuration space 
@@
 \item Shoot a random configuration in $\mathcal{C}$ with uniform distribution.
-\item Call the local planner on this configuration, with the static
+\item Call the \textit{constraint solver} (Algorithm~\ref{algo:newton}) on this configuration, with the static
   stability constraints and the goal task.
-\item Check that all of those tasks are achieved.
-\item Check for collisions.
+\item If success, check for collisions.
 \end{enumerate}
+
+\begin{algorithm}
+\caption{constraint-solver($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
+\label{algo:newton}
+\begin{algorithmic}
+\STATE $\alpha=.1$, $i=0$, $\alpha_{max} = .95$
+\WHILE{$\|f(q)\| > \varepsilon$ and $i\leq 20$}
+\STATE {\color{red} // $\left(.\right)^{+}$ denotes the Moore-Penrose pseudo-inverse}
+\STATE $q \leftarrow$ $q - \alpha \left(\frac{\partial f}{\partial q}(q)\right)^{+} f(q)$
+\STATE $i$ $\leftarrow$ $i+1$
+\STATE {\color{red}// Make $\alpha$ tend toward $\alpha_{max}$}
+\STATE $\alpha \rightarrow \alpha_{max} - .8*(\alpha_{max} - \alpha)$
+\ENDWHILE
+\IF {$\|f(q)\| \leq \varepsilon$}
+\STATE return success
+\ELSE
+\STATE return failure
+\ENDIF
+\end{algorithmic}
+\end{algorithm}
 
@@
 \begin{figure}[h]
-\centering
-\includegraphics[width=\linewidth]{pics/goal.png}
+\centerline {
+\includegraphics[width=.24\linewidth]{pics/goal-config/goal0.png}
+\includegraphics[width=.24\linewidth]{pics/goal-config/goal1.png}
+\includegraphics[width=.24\linewidth]{pics/goal-config/goal2.png}
+\includegraphics[width=.24\linewidth]{pics/goal-config/goal3.png}
+}
 
@@
   configurations are stable and collision-free, and the right hand of the character
-  reaches the purple box.}
+  reaches the orange ball. Note that the ball is also an obstacle.}
 \label{fig:goal}

### 2012-06-18 paper.tex
Add abstract and introduction. Fix typos. Change title.

@@
 
-\title{Whole-Body Manipulation Planning for Humanoid Robots}
+\title{Whole-Body Motion Planning for Humanoid Robots}
 
@@
 
+This paper presents a general method for planning collision-free whole-body walking motions
+for humanoid robots. First, we present an randomized algorithm for constrained motion
+planning, that is used to generate collision-free statically stable paths solving
+manipulation tasks. Then, we prove that these paths can always be approximated by collision-free
+dynamically balanced trajectories. The proof is based on humanoid robot
+\textit{small-space controllability}, and leads to a sound algorithm presented
+in the paper. We have  applied and evaluated our method
+on several  problems where whole-body planning and  walk are needed,
+and the results have been validated  on a real HRP-2 robot.
+
 \end{abstract}
@@
 
-Redundancy, walk
-
-Local planning
-
-different levels of complexity
+During the last twenty years, impressive progress has been achieved in humanoid
+robot hardware and control. This leads to a rising need for software and algorithms
+improving the usability and autonomy of those robots. One important area of research
+focuses on the development of robust and general motion generation techniques for safe
+and autonomous operation in human environments, such as offices or homes.
+
+Motion planning for humanoid robots is challenging for several reasons. First,
+the computational complexity of classic motion planning algorithms is exponential
+in the number of Degrees of Freedom (DoFs) of the considered system, which is
+high for humanoid kinematic trees. Second, a humanoid robot is an under-actuated system:
+the DoFs that control the position and orientation of the whole robot in space
+are not directly controlled, they derive from the articular DoFs of the robot legs.
+Those latter should be controlled with care to guarantee dynamically balanced motions,
+for manipulation or navigation.
+
+When planning collision free motions for humanoid robots, different representations
+of the robot and its environment can be used. The choice of the level of details of
+the representation indicates the difficulty of the considered problem. The simplest
+option consists in considering the robot as a navigating 2D shape, and computing 
+obstacle avoidance in a planar model of the world. Another possibility is to 
+compute only collision-free footsteps. 
+In complex and difficult environments, such as the one presented in Fig.~\ref{fig:couv},
+it can be necessary to consider exact 3D models of a humanoid robot and its environment.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.6\linewidth]{pics/chairs/couv.png}
+\caption{The robot HRP-2 passing between two chairs. In this kind of
+  environment whole-body collision avoidance is needed during
+  locomotion.} 
+\label{fig:couv}
+\end{figure}
+
+There are two ways to use motion planners to generate dynamically balanced robotic motions.
+The more general one is to plan in a robot dynamic space. By taking into account both robot
+configuration and velocity, motions that satisfy dynamic balance constraints can be generated
+at a planning phase. When planning motion for humanoid robots, this is a particularly costly
+approach, as the size of the space to explore is augmented with the robot velocity and footstep
+positions. The other way is to first plan  a geometric path that can be approximated by a
+dynamic trajectory in a second step. The approach we present in this paper falls into the second 
+category. Some feasible dynamic motions are inherently impossible to compute with this kind
+of approach. For example, jumping motions cannot be generated by a purely geometric planner.
+
+In this paper, we present a planning algorithm that considers exact models of a humanoid
+robot and its environment. It is used to solve navigation and manipulation problems. Our planner 
+generates collision-free, statically stable 
+geometric paths. We also present a constructive formal proof that these paths can be approximated by
+collision-free, dynamically balanced walk trajectories. Our planner is designed for indoor
+environments, where the floor is horizontal and flat. Because we do not compute explicitely footstep
+positions at the planning stage, the presented planner is unable to plan motions in which the
+robot steps over obstacles. This limitation is discussed in the paper.
 
@@
 
-Our algorithm generalize randomized tree expansion strategies, introduced in both \cite{HsuLat99c} and
+Our algorithm generalizes randomized tree expansion strategies, introduced in both \cite{HsuLat99c} and
 \cite{kuffner00rrtconnect} to constrained motion planning problems. Next subsection will recall the structure of the RRT algorithm,
@@
 not small-time, if the trajectories generated by its controller are arbitrarily long.
-The important property
-in terms of motion planning application is the small-space controllability, as 
+The important property, regarding
+motion planning application, is small-space controllability, as 
 Theorem~\ref{thm:ssc} shows. In the following, we show that legged robots
@@
 
-This section  discusses the small-space controllability of walking robots. To clarify
+This section  discusses walking robot small-space controllability. To clarify
 the presentation, we consider a simplified model of a legged robot consisting of two feet
@@
 Without loss of generality, let us assume that the robot is in a configuration
-in which the CoM is at the horizontal position $(0,0)$, the center of the feet are 
-aligned with the $y$-axis and the horizontal distance between the CoM and either of the centers
-of the feet is $L$. To achieve dynamically balanced walking, we aim at making
+in which the CoM is at the horizontal position $(0,0)$, the foot centers  are 
+aligned with the $y$-axis and the horizontal distance between the CoM and either of the foot centers
+is $L$. To achieve dynamically balanced walking, we aim at making
 $p_y(t)$ oscillate  between $-L$ and $L$. To move the ZMP  under a
@@
 $(1+\left(\frac{\omega}{\omega_0}\right)^2)$.  Choosing  $\omega =
-\omega_0 \sqrt{\frac{L}{\epsilon} -1}$ makes  $p_y$ oscillates between $-L$
-and    $L$, while $y$ oscillate between $-\epsilon$ and $\epsilon$.   
+\omega_0 \sqrt{\frac{L}{\epsilon} -1}$ makes  $p_y$ oscillate between $-L$
+and    $L$, while $y$ oscillates between $-\epsilon$ and $\epsilon$.   
 At    time   $t_l^{(n)}    =    n\frac{2\pi}{\omega}   +
@@
 
-The environment shown in Fig. \ref{fig:couv} and \ref{fig:sliding} has been
-first presented in \cite{el2011path}. It was solved by using a bounding
+The environment shown in Fig. \ref{fig:couv} and \ref{fig:sliding} was presented
+in \cite{el2011path}. There, the motion planning problem is solved with a bounding
 box method, leading the robot to walk sideways between the two chairs.

### 2012-06-18 paper.tex
Change to more explicit title

@@
 
-\title{Whole-Body Motion Planning for Humanoid Robots}
+\title{Locomotion and Whole-Body Motion Planning for Humanoid Robots: an Integrated Approach}
 
@@
 This paper presents a general method for planning collision-free whole-body walking motions
-for humanoid robots. First, we present an randomized algorithm for constrained motion
+for humanoid robots. First, we present a randomized algorithm for constrained motion
 planning, that is used to generate collision-free statically stable paths solving
@@
 dynamically balanced trajectories. The proof is based on humanoid robot
-\textit{small-space controllability}, and leads to a sound algorithm presented
+\textit{small-space controllability} and leads to a sound algorithm presented
 in the paper. We have  applied and evaluated our method

### 2012-06-19 paper.tex
Change description of constrained rrt to comply with implementation in hpp-constrained

@@
 \usepackage{algorithm}
+
+\usepackage{url}
 
@@
 \STATE $q_{near} \leftarrow $ Nearest$(q_{rand},\mathcal{T})$
-\STATE Extend$(q_{near},q_{rand})$
+\STATE Extend$(\mathcal{T},q_{near},q_{rand})$
 \ENDFOR
@@
 
-  \begin{tikzpicture}[x=0.5cm,y=0.5cm]
+  \begin{tikzpicture}[x=0.45cm,y=0.45cm]
 
@@
 
-    \node [draw,circle,inner sep=2pt] (qn1) at (4.8,2) {};
-    \node [draw,circle,inner sep=2pt] (qn2) at (6.2,2) {};
-    \node [draw,circle,inner sep=2pt] (qn3) at (7.6,2) {};
-
-    \draw (qn) -- (qn1) -- (qn2) -- (qn3) ;
-
-    \path [<->,thick,red] (3.4,1.5) edge  node [below,black]
-          {$\varepsilon$}  (4.8,1.5) ;
-          \path [<->,thick,red] (4.8,1.5) edge  node [below,black]
-                {$\varepsilon$}  (6.2,1.5) ;
-                \path [<->,thick,red] (6.2,1.5) edge  node [below,black]
-                      {$\varepsilon$} (7.6,1.5) ;
-
-                      \node [rectangle] (newconf) at (6.2,3.5) {\begin{small}new configurations\end{small}};
-
-                      \draw [->,blue] (newconf) -- (qn1.north) ;
-                      \draw [->,blue] (newconf) -- (qn2.north) ;
-                      \draw [->,blue] (newconf) -- (qn3.north) ;
-
-                      \node [draw,circle,inner sep=2pt] (qrand) at (13,2) {};
-                      \node [below right] at (qrand) {$q_{rand}$};
-
-                      \draw [dashed,thin] (qn3) -- (qrand) ;
+    \node [draw,circle,inner sep=2pt] (qn2) at (7.6,2) {};
+    \node [above left] at (qn2) {$q_{new}$};
+
+    \draw (qn) -- (qn2) ;
+
+    \path[draw=black,line join=miter,line cap=butt,line width=0.800pt, fill=gray]
+    (8,2) .. controls (8,4) and (10,6) ..
+    (12,5) .. controls (14,4) and (14,4) ..
+    (15,2) .. controls (15,0) and (14,0) ..
+    (13,0) .. controls (11,1) and (8,1) ..
+    (8,2) -- cycle;
+
+    \node [draw,circle,inner sep=2pt,fill=white] (qrand) at (13,2) {};
+    \node [below] at (qrand) {$q_{rand}$};
+
+    \node at (10.5,4) {Obstacle};
+
+    \draw [dashed,thin] (qn2) -- (qrand) ;
 
@@
 
-  \caption{One step of extension of the RRT algorithm} 
+  \caption{One step of extension of the RRT algorithm. The algorithm tries to add the longest possible edge
+  from $q_{near}$ towards $q_{rand}$, while avoiding collisions.} 
   \label{fig:rrt-extend}
@@
 
-The equivalent random  extension on a constrained manifold $\mathcal{M}$
+The equivalent random  extension on a constrained manifold $\mathcal{M}$, 
+defined by the constraint function $f$,
 starts from a valid configuration $q_{near} \in \mathcal{M}$, and extends the tree
 towards a random configuration $q_{rand}$, while keeping the constraints
-satisfied. To do so, we add a task to the local planner, whose value
-is the distance to  $q_{rand}$ in $\mathcal{C}$. Algorithm~\ref{alg:constrained}
-shows the pseudo-code of a constrained random extension from  $q_{near}$ towards
-$q_{rand}$ while satisfying the constraints defining $\mathcal{M}$: $Constraints$.
-The configuration space task used to explore $\mathcal{C}$ is 
-noted $ConfigurationTask$.   This
-task is added with the lowest priority. Each new configuration $q_{new}$ generated by the
-local planner  is 
-checked for constraint satisfaction and collision 
-avoidance. If the new configuration is valid, each new edge is checked for collision
-avoidance, as in a classic RRT algorithm.
+satisfied. Extension attempts orthogonal to $\mathcal{M}$ are useless,
+as newly added edges have to be included  in $\mathcal{M}$. To extend in directions
+that follow the directions of $\mathcal{M}$, we rely on the framework of 
+jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents 
+the adaptation of the classic extend function, and Fig.~\ref{fig:gikrrt} illustrates this
+extension. The idea is to first project $q_{rand}$ on the tangent
+space to $\mathcal{M}$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
+$q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
+$q_{rand}''$  be the result of a call to $\text{constraint-solver}(q_{rand}',f,\epsilon)$.  It is 
+the projection
+of $q_{rand}'$ on $\mathcal{M}$. Instead of extending the tree from $q_{near}$ towards 
+$q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
+$q_{rand}''$\footnote{The algorithm presented here gives a precise idea of the method, without 
+  focusing on technical details. Readers interested in the actual implementation can
+  refer to \url{https://github.com/laas/hpp-constrained} and 
+  \url{https://github.com/laas/hpp-constrained-planner}, where the open-source 
+  code corresponding to this algorithm is made available.}.
+
+\begin{figure}[h]
+\centering
+\begin{minipage}[c]{0.6\linewidth}
+\begin{tikzpicture}[y=0.55pt, x=0.55pt,yscale=-1, inner sep=0pt, outer sep=0pt]
+\definecolor{dg}{rgb}{0,0.3,0}
+\path[draw=black,line join=miter,line cap=butt,line width=0.800pt, fill=gray]
+  (231.3249,143.2249) .. controls (178.0814,153.1889) and (178.5527,172.4247) ..
+  (180.8173,200.8036) .. controls (183.2959,231.8645) and (195.6610,188.6038) ..
+  (256.5787,230.0980) .. controls (295.8993,256.8813) and (346.4296,308.3911) ..
+  (295.9747,178.5802) .. controls (275.5517,126.0356) and (304.8066,129.4735) ..
+  (231.3249,143.2249) -- cycle;
+\node at (230,180) {Obstacle};
+\path[dashed,draw=dg,line join=miter,line cap=butt,line width=1.5pt]
+  (88.8934,277.5752) .. controls (196.3490,276.5363) and (532.8532,213.7494) ..
+(434.3656,81.6056) node [above = 0.1cm, color=dg, text width=1.8cm] {\small{Constrained manifold $\mathcal{M}$}};
+\node [draw,circle,inner sep=2pt,fill = gray] (qn) at (97,277) {};
+\node [above = 0.15cm] at (qn) {$q_{near}$};
+\node [draw,circle,inner sep=2pt] (qr1) at (367,99) {};
+\node [left = 0.2cm] at (qr1) {$q_{rand}$};
+\draw [dashed,thin] (qn) -- (450,277);
+\node [draw,circle,inner sep=2pt] (qr2) at (367,277) {};
+\node [below right = 0.1cm] at (qr2) {$q_{rand}'$};
+\node [draw,circle,inner sep=2pt,fill = gray] (qr3) at (345,225) {};
+\node [above = 0.22cm] at (qr3) {$q_{rand}''$} ;
+\draw [dashed,thin] (qr1) -- (qr2) -- (qr3);
+\node [draw,circle,inner sep=2pt,fill = gray] (qnew) at (270,250);
+\node [above = 0.1cm, left = 0.2cm] at (qnew) {$q_{new}$};
+%\draw [thick] (qn) -- (qnew);
+
+\end{tikzpicture}
+\end{minipage}
+\begin{minipage}[c]{0.3\linewidth}
+\begin{tikzpicture}[x=0.61cm,y=0.61cm]
+   \definecolor{dg}{rgb}{0,0.3,0}
+      %Stack of tasks:
+      \draw [color=black, thick,fill = white] (0,1) rectangle (2,2);
+      \draw [color=black, thick,fill = white] (0,2) rectangle (2,3);
+      \draw [color=black, thick,fill = white] (0,3) rectangle (2,5);
+      \draw [color=black, thick,fill = white] (0,5) rectangle (2,6);
+      \draw [color=black, thick,fill = white] (0,6) rectangle (5,7);
+      \draw [color=black, thick,fill = white] (1,1) rectangle (5,6);
+      \draw [thick] (1,1) -- (5,1) -- (5,6);
+      \draw (1,2) -- (5,2) ;
+      \node [text width = 5cm,text centered] at (2.5,6.5) {\large Stack
+        of tasks};
+      \begin{scriptsize}
+        \node at (0.5,5.5) {1};
+        \node at (0.5,4) {$\vdots$}; 
+        \node at (0.5,2.5) {$n$};
+        \node at (0.5,1.5) {$n$+1};
+
+        \node [text width = 2cm,color=dg] at (3.2,4) {\small Constraints defining
+          $\mathcal{M}$};
+        \node [text width = 3cm,text centered] at (3,1.5)
+              { Configuration task towards $q_{rand}$};    
+      \end{scriptsize}
+
+\end{tikzpicture}
+\end{minipage}
+
+\caption{One step of constrained extension, and the stack of tasks used
+  to generate $q_{rand}'$ with a prioritized inverse kinematics solver, 
+  starting from $q_{near}$:
+  first, the constraints defining $\mathcal{M}$, and at a lower priority
+  a configuration task towards $q_{rand}$.  $q_{rand}'$ is then projected onto
+  $\mathcal{M}$, and a classic RRT extension tries to go as far as possible
+  from $q_{near}$ towards $q_{rand}''$.}
+\label{fig:gikrrt}
+\end{figure}
 
 \begin{algorithm}[h]
-  \caption{Constrained-Extend($\mathcal{T},q_{near},q_{rand},Constraints$)}
+  \caption{Constrained-Extend($\mathcal{T},q_{near},q_{rand},f,\epsilon$)}
   \label{alg:constrained}
   \begin{algorithmic}
-    \STATE $Tasks.$Initialize$(Constraints)$
-    \STATE $ConfigurationTask$.SetTarget($q_{rand}$)
-    \STATE $Tasks.$Add$(ConfigurationTask)$
-    \STATE $q_{current} \leftarrow q_{near}$
-    \STATE $State \leftarrow$ Progressing
-    \WHILE{$State =  $ Progressing}
-    \STATE $q_{new} \leftarrow $ LocalPlannerPerformOneStep($Tasks,q_{current}$)
-    \IF{$q_{new} \neq q_{current}$ \\ 
-      and CollisionCheck($q_{new}$) = OK\\ 
-      and CollisionCheckEdge($q_{current},q_{new}$) = OK \\
-      and ConstraintsCheck($q_{new},Constraints$) = OK }
-    \STATE $\mathcal{T}$.AddNode($q_{new}$)
-    \STATE $\mathcal{T}$.AddEdge($q_{current},q_{new}$)
-    \STATE $q_{current} \leftarrow q_{new}$
-    \IF{$q_{target} = q_{rand}$}
-    \STATE $State \leftarrow$ Reached
+    \STATE {\color{red} // $I$ denotes the identity matrix.}
+    \STATE $q_{rand}' \leftarrow q_{near} +   
+    \left(I -  
+    \left(\frac{\partial f}{\partial q}(q_{near})\right)^{+}
+    \frac{\partial f}{\partial q}(q_{near})\right) (q_{rand} - q_{near})$
+    \IF {$\text{constraint-solver}(q_{rand}',f,\epsilon) = success$} 
+    \STATE RRT::Extend($\mathcal{T},q_{near},q_{rand}'$)
     \ENDIF
-    \ELSE 
-    \STATE $State \leftarrow$ Trapped
-    \ENDIF
-    \ENDWHILE
-
   \end{algorithmic}
 \end{algorithm}
-
-Fig. \ref{fig:gikrrt} shows one step of constrained
-extension, and the prioritized stack of tasks and constraints solved by the local planner.
-
-\begin{figure}[h]
-\centering
-  \begin{tikzpicture}[x=0.61cm,y=0.61cm]
-    \definecolor{dg}{rgb}{0,0.3,0}
-     \draw [dashed,color=dg,thick,text width=2cm] (1,1)
-      arc(139:90:10.61) node [above right] {\small{Constrained manifold $\mathcal{M}$}} ; 
-      \node [draw,circle,inner sep=2pt,fill=gray] (qnear) at (2,2) {};
-      \node [left] (qn0) at (qnear) {$q_{near}$} ;
-      \node [draw,circle,inner sep=2pt] (q1) at (3,3) {};
-      \node [draw,circle,inner sep=2pt] (q2) at (4,4) {};
-      \node [draw,circle,inner sep=2pt] (q3) at (5,5) {};
-      \node [draw,circle,inner sep=2pt] (qrand) at (7,7) {};
-      \node [below right] (qr0) at (qrand) {$q_{rand}$} ;
-      \draw [dashed,thin] (qnear) -- (q1) -- (q2) -- (q3) -- (qrand) ;
-
-    \node [draw,circle,inner sep=2pt,fill=gray] (qc1) at (3.11,2.83) {};
-      \draw (qnear) -- (qc1);
-      \node [draw,circle,inner sep=2pt,fill=gray] (qc2) at (4.26,3.49) {};
-      \draw (qc1) -- (qc2);
-     \node [draw,circle,inner sep=2pt,fill=gray] (qc3) at (5.37,3.97) {};
-      \draw (qc2) -- (qc3);
-      \node [rectangle] (newconf) at (6,2) {\small{new configurations}}; 
-      \draw [->,blue] (newconf) -- (qc1) ;
-      \draw [->,blue] (newconf) -- (qc2) ;
-      \draw [->,blue] (newconf) -- (qc3) ;
-
-      \draw [color=black, thick,fill = white] (15,1) rectangle (17,2);
-      \draw [color=black, thick,fill = white] (15,2) rectangle (17,3);
-      \draw [color=black, thick,fill = white] (15,3) rectangle (17,5);
-      \draw [color=black, thick,fill = white] (15,5) rectangle (17,6);
-      \draw [color=black, thick,fill = white] (15,6) rectangle (20,7);
-      \draw [color=black, thick,fill = white] (16,1) rectangle (20,6);
-      \draw [thick] (16,1) -- (20,1) -- (20,6);
-      \draw (16,2) -- (20,2) ;
-      \node [text width = 5cm,text centered] at (17.5,6.5) {\large Stack
-        of tasks};
-      \begin{scriptsize}
-        \node at (15.5,5.5) {1};
-        \node at (15.5,4) {$\vdots$}; 
-        \node at (15.5,2.5) {$n$};
-        \node at (15.5,1.5) {$n$+1};
-
-        \node [text width = 2cm,color=dg] at (18.2,4) {\small Constraints defining
-          $\mathcal{M}$};
-        \node [text width = 3cm,text centered] at (18,1.5)
-              { Configuration task towards $q_{rand}$};    
-      \end{scriptsize}
-
-\end{tikzpicture}
-
-\caption{One step of constrained extension, and the stack of tasks used
-  to generate new configurations with a local planner starting from $q_{near}$:
-  first, the constraints defining $\mathcal{M}$, and at a lower priority
-  a configuration task towards $q_{rand}$.}
-\label{fig:gikrrt}
-\end{figure}
 
@@
 walk trajectory  corresponding to  smaller and faster  steps converges
-toward the  sliding path.  Algorithm  \ref{alg:walk} shows pseudo-code
+toward the  sliding path.  Algorithm ~\ref{alg:walk} shows pseudo-code
 that takes  a sliding path $p$  as input and  returns a collision-free
-walk trajectory.
+walk trajectory\footnote{The actual implementation of this algorithm is
+part of an open-source package, available at
+\url{https://github.com/laas/hpp-wholebody-step-planner}.}.
 

### 2012-06-20 paper.tex
Add precision about the constrained rrt method
Add snapshots of the 'College de France' experiment. Minor corrections.

@@
 
-There are two ways to use motion planners to generate dynamically balanced robotic motions.
+There are two main ways of using motion planners to generate dynamically balanced robotic motions.
 The more general one is to plan in a robot dynamic space. By taking into account both robot
@@
 $q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
-$q_{rand}''$\footnote{The algorithm presented here gives a precise idea of the method, without 
+$q_{rand}''$\footnote{This presentation attempts to give a precise idea of the algorithm, without 
   focusing on technical details. Readers interested in the actual implementation can
   refer to \url{https://github.com/laas/hpp-constrained} and 
-  \url{https://github.com/laas/hpp-constrained-planner}, where the open-source 
-  code corresponding to this algorithm is made available.}.
+  \url{https://github.com/laas/hpp-constrained-planner}, where the corresponding open-source 
+  code is available.}. While extending the tree,
+the configurations along the new edge are automatically projected onto $\mathcal{M}$. These projections 
+are not very costly if the edge is close to the constrained manifold.
 
@@
 comprehensive video of this problem, including the motion execution on
-the real robot HRP-2.
+the real robot HRP-2 on stage. Fig.~\ref{fig:shelf-cdf} shows some snapshots 
+taken from Extension 3.
 
@@
 
-\includegraphics[width=0.5\linewidth]{pics/shelves/waist-trajectory.png}
+\includegraphics[width=0.4\linewidth]{pics/shelves/waist-trajectory.png}
 
@@
 \label{fig:shelf-waist}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+
+\includegraphics[width=0.19\linewidth]{pics/shelves-cdf/1.jpg}
+\includegraphics[width=0.19\linewidth]{pics/shelves-cdf/2.jpg}
+\includegraphics[width=0.19\linewidth]{pics/shelves-cdf/3.jpg}
+\includegraphics[width=0.19\linewidth]{pics/shelves-cdf/4.jpg}
+\includegraphics[width=0.19\linewidth]{pics/shelves-cdf/5.jpg}
+
+\caption{Execution of the walking trajectory by HRP-2 on stage. The robot first goes
+  to the shelves to release the ball, then comes back to a rest position.}
+
+\label{fig:shelf-cdf}
 \end{figure}

### 2012-06-20 paper.tex
Few corrections from Jean-Paul

@@
 
-\title{Locomotion and Whole-Body Motion Planning for Humanoid Robots: an Integrated Approach}
+\title{Dynamic Walking and Whole-Body Motion Planning for Humanoid Robots: an Integrated Approach}
 
@@
 planning, that is used to generate collision-free statically stable paths solving
-manipulation tasks. Then, we prove that these paths can always be approximated by collision-free
-dynamically balanced trajectories. The proof is based on humanoid robot
-\textit{small-space controllability} and leads to a sound algorithm presented
-in the paper. We have  applied and evaluated our method
-on several  problems where whole-body planning and  walk are needed,
+manipulation tasks. Then, we show that dynamic walking makes humanoid robots 
+small-space controllable. Such a property allows to easily transform 
+collision-free statically stable paths into 
+collision-free dynamically balanced trajectories. It leads to a sound 
+algorithm which has been applied and evaluated on several
+problems where whole-body planning and  walk are needed,
 and the results have been validated  on a real HRP-2 robot.
@@
 
-\subsection{Walking Robot Controllability}
+\subsection{Dynamic Walking Makes Humanoid Robots Small-Space Controllable}
 \label{sec:humanoid-ssc}

### 2012-06-20 paper.tex
Add experimental results for grasping problem.

@@
 \draw [dashed,thin] (qr1) -- (qr2) -- (qr3);
-\node [draw,circle,inner sep=2pt,fill = gray] (qnew) at (270,250);
+\node [draw,circle,inner sep=2pt,fill = gray] (qnew) at (270,250) {};
 \node [above = 0.1cm, left = 0.2cm] at (qnew) {$q_{new}$};
@@
 
-An experimental study, presented in  \cite{dalibard09}, showed that
-in complex environments, randomized motion planners were faster than
-collision avoiding local planners. Another benefit of using global motion
-planning compared to local gradient descent is the ability to avoid local 
-minima.
-
 We present in Fig.~\ref{fig:wb-shelves} an illustration of 
@@
 defined by tasks constraining the robot feet and CoM. The goal of the first problem
-is defined by a hand configuration task (position and orientation), and a gaze task
-(the robot has to see the ball). The goal of the second problem is defined by a 
-different hand configuration and the same gaze task,
-and finally, the goal of the third problem is the rest configuration.
+is defined by a hand configuration task (3 positions and 2 orientations), and a gaze task
+(the robot has to look at the ball). The goal of the second problem is defined by a 
+different hand configuration and the gaze task.
+Finally, the goal of the third problem is the rest configuration.
+
+For the two reaching motions, we generate 8 random goal configurations (Section~\ref{sec:goal-sampling}).
+
+The result of RRT is optimized in two steps (see~\cite{dalibard09} for details):
+\begin{enumerate}
+\item the path is extended in order to move the final posture closer to the rest posture,
+\item the path is optimized by random short-cuts.
+\end{enumerate}
+
 Extension 1 presents a video of this motion.
@@
 
-The motion presented in Fig.~\ref{fig:wb-shelves} looks very complicated.
-The reasons are that the environment is complex and HRP-2 kinematic tree is highly constrained.
-The motion has been optimized based on these constraints. Note that the
-accessibility space of this robot is not wide enough to accomplish this task further 
-away from the shelves, which could have eased the motion. One possibility to extend the
-robot accessibility space is to automatically change the support polygon durint the motion to solve the task.
-This is an ongoing research field, \cite{OussamaKanoun06102010} proposes some 
-local solutions. This has not been integrated in our global planning algorithm yet.
+We have run this set of motion planning problems 20 times.
+The results are compiled in Table~\ref{table:reaching}.
+We can notice that optimization dominates the total time of computation.
+This suggests that improvement can be easily achieved by using numerical optimization techniques.
+This is on-going work.
+\begin{table}
+\begin{tabular}{lrrr}
+& min & max & average \\
+number of nodes & 43.00 & 481.00 & 102.70 \\
+goal generation time (seconds) \hspace {1cm} & 1.00 & 1.56 & 1.22 \\
+planning time (seconds)& 67.36 & 376.84 & 134.28 \\
+optimization time (seconds) & 204.22 & 671.26 & 521.47 \\
+total times (seconds) & 468.23 & 864.68 & 656.97
+\end{tabular}
+\caption {Experimental results on 20 runs: 3 motion planning problems and 2 goal generations.}
+\label{table:reaching}
+\end{table}
+
+Among all calls to \texttt{constraint-solver}, the average number of iterations is 6.5 and the success rate is above 95 percent.
 

### 2012-06-20 paper.tex
Algo 2 returns a configuration and Algo 3 uses q'' instead of q'.
Add a column in table to make result look less miserable.

@@
 planning is needed to overcome this limitation.
-\cite{TouGieGoe2007} Propose a motion generation method where tasks follow
+\cite{TouGieGoe2007} propose a motion generation method where tasks follow
 trajectories define by cubic B-splines. The whole-body motion is optimized
@@
 \IF {$\|f(q)\| \leq \varepsilon$}
-\STATE return success
+\STATE return $q$
 \ELSE
@@
     \frac{\partial f}{\partial q}(q_{near})\right) (q_{rand} - q_{near})$
-    \IF {$\text{constraint-solver}(q_{rand}',f,\epsilon) = success$} 
-    \STATE RRT::Extend($\mathcal{T},q_{near},q_{rand}'$)
+    \IF {$\text{constraint-solver}(q_{rand}',f,\epsilon) = q_{rand}''$} 
+    \STATE RRT::Extend($\mathcal{T},q_{near},q_{rand}''$)
     \ENDIF
@@
 \begin{table}
-\begin{tabular}{lrrr}
-& min & max & average \\
+\begin{tabular}{lrrrr}
+& min & max & average & average \\ 
+&&&& per problem \\
 number of nodes & 43.00 & 481.00 & 102.70 \\
-goal generation time (seconds) \hspace {1cm} & 1.00 & 1.56 & 1.22 \\
-planning time (seconds)& 67.36 & 376.84 & 134.28 \\
-optimization time (seconds) & 204.22 & 671.26 & 521.47 \\
-total times (seconds) & 468.23 & 864.68 & 656.97
+goal generation time \hspace {1.cm} & 1.00 & 1.56 & 1.22\\
+planning time& 67.36 & 376.84 & 134.28 & 44.76\\
+optimization time & 204.22 & 671.26 & 521.47 & 173.82 \\
+total times & 468.23 & 864.68 & 656.97 &
 \end{tabular}
-\caption {Experimental results on 20 runs: 3 motion planning problems and 2 goal generations.}
+\caption {Experimental results on 20 runs: 3 motion planning problems and 2 goal generations. Times are expressed in seconds.}
 \label{table:reaching}

### 2012-06-20 script/benchmark.py
Add scripts used to extract experimental results

@@ -1,0 +1,60 @@
+pid=2853
+success = parseLineInLog (prefix='INFO:/home/florent/devel/hpp/src/hpp-constrained/src/config-projector.cc:151: Projection succeeded.', pid=pid)
+fails = parseLineInLog (prefix='INFO:/home/florent/devel/hpp/src/hpp-constrained/src/config-projector.cc:148: Projection failed', pid=pid)
+p = 'INFO:/home/florent/devel/hpp/src/hpp-constrained/src/config-projector.cc:136: '
+cumulative = []
+cumulative.append (parseLineInLog (prefix = p + '0.1', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.27', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.406', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.5148', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.60184', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.671472', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.727178', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.771742', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.807394', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.835915', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.858732', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.876986', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.891588', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.903271', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.912617', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.920093', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.926075', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.93086', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.934688', pid=pid))
+cumulative.append (parseLineInLog (prefix = p + '0.93775', pid=pid))
+
+cumulative.append (0)
+iteration = map (lambda x : x[0] - x [1], zip (cumulative, cumulative [1:]))
+
[…]

### 2012-06-20 script/grasping.py

@@ -1,0 +1,113 @@
+import sys, os
+import hpp.corbaserver
+import hpp.corbaserver.constrained
+from hpp_corbaserver.hpp import Configuration
+from tools import parseConfigInLog, parseTimeInBenchmark, parseLineInLog
+
+cl = hpp.corbaserver.Client(['robot', 'obstacle', 'problem'])
+clw = hpp.corbaserver.constrained.client.Client()
+
+def setHand (right, alpha, lock):
+    if right:
+        rank = 28
+    else:
+        rank = 40
+    q = cl.robot.getCurrentConfig (0)
+    q[rank:rank+6] = [alpha, -alpha, alpha, -alpha, alpha, -alpha]
+    cl.robot.setCurrentConfig (0,q)
+    if lock:
+        for dof in range (rank, rank+6):
+            cl.robot.setDofLocked (0, dof, True, q [dof])
+
+def openHand (right, lock):
+    setHand (right, .75, lock)
+
+def closeHand (right, lock):
+    setHand (right, .38, lock)
+
+env_dir = os.getenv('HOME')+'/devel/hpp/src/hpp-environment-data/share/hpp-environment/2012ijrr-shelves'
+
+def playPath (robot, path):
+    length = cl.problem.pathLength (robot, path)
+    for i in range (201):
+        l = i*length/200.
+        cfg = cl.problem.configAtDistance (robot, path, l)
+        cl.robot.setCurrentConfig (robot, cfg)
+
+cl.robot.loadHrp2Model(.05)
+half_sitting = cl.robot.getCurrentConfig(0)
+cl.problem.initializeProblem()
+cl.problem.parseFile(env_dir + "/env.kxml")
+
+nodes = []
+edges = []
+for i in range (20):
+    # Reaching motion starting from half-sitting
+    # Set right hand locked open
+    openHand (True, True)
+    # Set left hand lock closed
+    closeHand (False, True)
+    q = cl.robot.getCurrentConfig (0)
+    q_init = q [::]
+    cl.problem.setInitialConfig(0, q)
+    if clw.problem.setTarget (0.62, 0.07, 1.02) != 0:
+        raise RuntimeError ("Failed to set target.")
+    clw.problem.generateGoalConfigurations (0, 8)
+    if cl.problem.solve () != 0:
+        raise RuntimeError ("Failed to reach the object")
+    pathId = 1+6*i
[…]

### 2012-06-20 script/tools.py

@@ -1,0 +1,52 @@
+import os
+import datetime
+
+def playPath (cl, robot, path):
+    length = cl.problem.pathLength (robot, path)
+    for i in range (101):
+        l = i*length/100.
+        cfg = cl.problem.configAtDistance (robot, path, l)
+        cl.robot.setCurrentConfig (robot, cfg)
+
+def isfloat (string) :
+    try :
+        float (string)
+    except ValueError as exc :
+        return False
+    return True
+    
+def parseConfigInLog (pid, prefix):
+    listConfig = []
+    devel_dir = os.getenv ('DEVEL_DIR')
+    with open (devel_dir + '/stable/var/log/hpp/journal.' + str(pid) + '.log',
+               'r') as f:
+        for line in f:
+            if line[:len(prefix)] == prefix:
+                configString = line [len(prefix):].strip(' ')
+                config = map (float, filter (isfloat, configString.split (' ')))
+                listConfig.append (config)
+    return listConfig
+
+def parseLineInLog (pid, prefix):
+    nbLines = 0
+    devel_dir = os.getenv ('DEVEL_DIR')
+    with open (devel_dir + '/stable/var/log/hpp/journal.' + str(pid) + '.log',
+               'r') as f:
+        for line in f:
+            if line[:len(prefix)] == prefix:
+                nbLines += 1
+    return nbLines
+
+def parseTimeInBenchmark (pid, prefix):
+    listTime = []
+    devel_dir = os.getenv ('DEVEL_DIR')
+    with open (devel_dir + '/stable/var/log/hpp/benchmark.' + str(pid) + '.log',
+               'r') as f:
+        for line in f:
+            if line[:len(prefix)] == prefix:
+                timeString = line [len(prefix):].strip (' ')
+                t = timeString.split (':')
+                listTime.append (datetime.timedelta (hours = int (t [0]),
+                                                     minutes = int (t [1]),
+                                                     seconds = float (t [2])))
+    return listTime

### 2012-06-20 paper.tex
Do not speak about optimization.

@@
 
-The result of RRT is optimized in two steps (see~\cite{dalibard09} for details):
-\begin{enumerate}
-\item the path is extended in order to move the final posture closer to the rest posture,
-\item the path is optimized by random short-cuts.
-\end{enumerate}
-
-Extension 1 presents a video of this motion.
+Extension 1 presents a video of this motion after an optimization step.
 
@@
 
-We have run this set of motion planning problems 20 times.
-The results are compiled in Table~\ref{table:reaching}.
-We can notice that optimization dominates the total time of computation.
-This suggests that improvement can be easily achieved by using numerical optimization techniques.
-This is on-going work.
+We have run this set of motion planning problems 20 times. The results are compiled in Table~\ref{table:reaching}.
 \begin{table}
@@
 planning time& 67.36 & 376.84 & 134.28 & 44.76\\
-optimization time & 204.22 & 671.26 & 521.47 & 173.82 \\
-total times & 468.23 & 864.68 & 656.97 &
 \end{tabular}

### 2012-06-20 paper.tex
constraint-solver -> \texttt{constraint-solver}
Minor rephrasing

@@
 \begin{algorithm}
-\caption{constraint-solver($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
+\caption{\texttt{constraint-solver}($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
 \label{algo:newton}
@@
 $q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
-$q_{rand}''$  be the result of a call to $\text{constraint-solver}(q_{rand}',f,\epsilon)$.  It is 
+$q_{rand}''$  be the result of a call to $\texttt{constraint-solver}(q_{rand}',f,\epsilon)$.  It is 
 the projection
@@
     \frac{\partial f}{\partial q}(q_{near})\right) (q_{rand} - q_{near})$
-    \IF {$\text{constraint-solver}(q_{rand}',f,\epsilon) = q_{rand}''$} 
+    \IF {$\texttt{constraint-solver}(q_{rand}',f,\epsilon) = q_{rand}''$} 
     \STATE RRT::Extend($\mathcal{T},q_{near},q_{rand}''$)
@@
 
-This result has been long known and used in motion planning, in particular in planning
-for non-holonomic systems. A detailed proof of it can be found in
+This result has been long known and used in motion planning, in particular
+for non-holonomic systems. A detailed proof can be found in
 \cite{taix-94}. We will present a sketch of the proof to give an intuition about the 
@@
 $T$ forms a  neighborhood of $q$. When accelerations and velocities are bounded,
-small-time controllability implies small-space controllability, which is why 
+small-time controllability implies small-space controllability. This is why 
 a lot of motion planning previous work only refers to the sufficient small-time controllability
@@
 the presentation, we consider a simplified model of a legged robot consisting of two feet
-of null mass and a punctual mass free to move in three dimensions.
+of zero mass and a point mass free to move in three dimensions.
 We do not consider 

### 2012-11-27 paper.tex
Add Acknowledgment section
Fix typos (preliminary pass).

@@
 collision-free, dynamically balanced walk trajectories. Our planner is designed for indoor
-environments, where the floor is horizontal and flat. Because we do not compute explicitely footstep
+environments, where the floor is horizontal and flat. Because we do not compute explicitly footstep
 positions at the planning stage, the presented planner is unable to plan motions in which the
@@
 The problem of inverse kinematics for a humanoid robot, or any articulated
-structure, is to compute a joint motion to achieve an end-effector pose. As the
+structure, is to compute a joint position to achieve an end-effector pose. As the
 robots we deal with are redundant, it is natural to take advantage of
@@
 different priorities. This problem has been widely studied in robotics
-planning and control literature, and many jacobian-based solutions have been
+planning and control literature, and many Jacobian-based solutions have been
 proposed, among which 
@@
 satisfy, see for example \cite{kanehiro2008lca}.
-These methods are prone to fall into local minima, global motion 
+These methods are prone to fall into local minima, thus global motion
 planning is needed to overcome this limitation.
 \cite{TouGieGoe2007} propose a motion generation method where tasks follow
-trajectories define by cubic B-splines. The whole-body motion is optimized
+trajectories defined by cubic B-splines. The whole-body motion is optimized
 with respect to the control point positions. This method can take into account
-collision with simple obstacles.
+collisions with simple obstacles.
 
@@
 the last few years, \cite{Berenson15032011} presents a exhaustive survey
-of jacobian-based methods. Other recent contributions \cite{porta2012randomized}
+of Jacobian-based methods. Other recent contributions \cite{porta2012randomized}
 present sophisticated constrained planning techniques based on higher-dimensional
@@
 complexity  of a  humanoid  robotics  kinematic tree  is  to use  the
-so-called "cart-table" simplified model \cite{kajita2003biped}. Based this
+so-called "cart-table" simplified model \cite{kajita2003biped}. Based on this
 model,  planning a trajectory  for the  ZMP is  reduced to  planning a
@@
 trajectory  of  the CoM  and  footstep  positions, inverse  kinematics
-solvers can animate  the whole set of degrees of freemdom (DoFs) of the  
+solvers can animate  the whole set of degrees of freedom (DoFs) of the
 robot to generate a dynamically stable walk trajectory.
@@
   set of inverse kinematics and balance tasks;
-\item the goal is also defined implicitely, by additional inverse kinematic tasks.
+\item the goal is also defined implicitly, by additional inverse kinematic tasks.
 \end{enumerate}
@@
 \begin{itemize}
-\item  Static stability: the center of mass of the robot stays at
-  the vertical of the support polygon center, the two feet have fixed position and
+\item Static stability: the center of mass of the robot stays above
+  the support polygon center, the two feet have fixed position and
   orientation.
@@
 that follow the directions of $\mathcal{M}$, we rely on the framework of 
-jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents 
+Jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents
 the adaptation of the classic extend function, and Fig.~\ref{fig:gikrrt} illustrates this
@@
 
-The balanced walking conditions for a quasi-static walking robot are that the punctual
+The balanced walking conditions for a quasi-static walking robot are that the point
 mass, or CoM, should always be over the support polygon (the convex hull of the two feet), and 
@@
 
-The model of a walking robot with a punctual
+The model of a walking robot with a point
 mass at a fixed height is known in the literature as the cart-table model \cite{kajita2003biped}.
@@
 \paragraph{Generalization to a complete model:} More work is needed to generalize 
-the previous proof to any legged robot model. First, because we use a model where
-the mass is punctual, we did not have to consider the angular-momentum of the robot
+the previous proof to any legged robot model. First, because we use a model with
+a point mass, we did not have to consider the angular-momentum of the robot
 around the CoM, while it accounts for dynamic balance. Moreover, when considering
@@
 \caption{Solution path for a hand reaching problem in an
-  appartment. The goal is implicitly defined as an inverse kinematics
+  apartment. The goal is implicitly defined as an inverse kinematics
   task.} 
@@
 
+\section{Acknowledgments}
+
+This work was supported by the French FUI Project ROMEO and the European Project ECHORD-231143.
+The authors would like to thank Thomas Moulard and Olivier Stasse for their valuable help during experiments.
+
 \bibliographystyle{agsm}

### 2012-11-27 script/benchmark_analyzer.py
Add script, data, and eps plots for rrt benchmarks.

@@ -1,0 +1,142 @@
+#!/usr/bin/python
+
+import numpy as np
+import scipy as sp
+import pylab as pl
+import matplotlib.pyplot as plt
+from matplotlib.patches import Polygon
+
+# Load benchmark data.
+oc_rrt = np.genfromtxt("objects-cloud/bench-rrt.log")
+oc_ipp = np.genfromtxt("objects-cloud/bench-ipp.log")
+sh_rrt = np.genfromtxt("shelf/bench-rrt.log")
+sh_ipp = np.genfromtxt("shelf/bench-rrt.log")
+# lt_rrt = np.genfromtxt("bench-lt-rrt.log")
+# pca_rrt = np.genfromtxt("bench-pca-rrt.log")
+# pca_ipp = np.genfromtxt("bench-pca-ipp.log")
+# lt_ipp = np.genfromtxt("bench-lt-ipp.log")
+# pca_lt_rrt = np.genfromtxt("bench-pca-lt-rrt.log")
+# pca_lt_ipp = np.genfromtxt("bench-pca-lt-ipp.log")
+
+rrt = [('Floating obstacles RRT', oc_rrt),
+       ('Floating obstacles IPP-RRT', oc_ipp),
+       ('Shelf RRT', sh_ipp),
+       ('Shelf IPP-RRT', sh_ipp)]
+       # ('Local-Trees RRT', lt_rrt),
+       # ('Local-Trees IPP-RRT', lt_ipp),
+       # ('PCA-RRT', pca_rrt),
+       # ('PCA-IPP-RRT', pca_ipp),
+       # ('PCA-LT-RRT', pca_lt_rrt),
+       # ('PCA-LT-IPP-RRT', pca_lt_ipp)]
+
+# Compute mean and standard deviation over rows (axis 0 of array).
+m_rrt = []
+s_rrt = []
+min_rrt = []
+max_rrt = []
+for i in range(len(rrt)):
+    m_rrt.append(sp.mean(rrt[i][1], 0))
+    s_rrt.append(sp.std(rrt[i][1], 0))
+    min_rrt.append(sp.amin(rrt[i][1], 0))
+    max_rrt.append(sp.amax(rrt[i][1], 0))
+
+# Build node, iterations, and time vectors.
+m_it = np.zeros(len(rrt))
+s_it = np.zeros(len(rrt))
+min_it = np.zeros(len(rrt))
+max_it = np.zeros(len(rrt))
+m_t = np.zeros(len(rrt))
[…]

### 2012-11-27 paper.tex
Fix some syntax and grammar errors.

@@
   \footnote{
-    This paper summarizes and and extends previous work that appeared in the 9th and 11th
+    This paper summarizes and extends previous work that appeared in the 9th and 11th
     IEEE-RAS International Conference on Humanoid Robots, 2009 and 2011.
@@
 small-space controllable. Such a property allows to easily transform 
-collision-free statically stable paths into 
+collision-free statically balanced paths into 
 collision-free dynamically balanced trajectories. It leads to a sound 
@@
 research: prioritized inverse kinematics, randomized  whole-body  motion
-planning and walk pattern generators based on the Zero-Momentum
+planning and walk pattern generators based on the Zero-Moment
 Point (ZMP) formalism. This section summarizes the literature related to each of 
@@
 When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
-is to  take into  account stability  constraints,  i.e. to
+is to  take into  account balance  constraints,  i.e. to
 generate  random  configurations   on  zero  volume  submanifolds  of
 $\mathcal{C}$. This problem has been investigated with success during
-the last few years, \cite{Berenson15032011} presents a exhaustive survey
+the last few years, \cite{Berenson15032011} presents an exhaustive survey
 of Jacobian-based methods. Other recent contributions \cite{porta2012randomized}
-present sophisticated constrained planning techniques based on higher-dimensional
+present sophisticated constrained motion planning techniques based on higher-dimensional
 continuation. Section \ref{sec:wb} presents a simple adaptation
@@
 Another  field of  humanoid  robotics research  is  the generation  of
-dynamically stable  walk patterns. Since  the introduction of  the ZMP
+dynamically balanced walk patterns. Since  the introduction of  the ZMP
 formalism  \cite{vukobratovic1969contribution},  several  methods  have  been  proposed  to
 generate  walking  motions efficiently.   One  way  to  deal with  the
-complexity  of a  humanoid  robotics  kinematic tree  is  to use  the
+complexity  of a  humanoid  robot  kinematic tree  is  to use  the
 so-called "cart-table" simplified model \cite{kajita2003biped}. Based on this
@@
 trajectory  for  the Center  of  Mass (CoM)  of  the  robot.  Given  a
-trajectory  of  the CoM  and  footstep  positions, inverse  kinematics
-solvers can animate  the whole set of degrees of freedom (DoFs) of the
+trajectory  of  the CoM  and  footprint  positions, inverse  kinematics
+solvers can animate  the whole set of DoFs of the
 robot to generate a dynamically stable walk trajectory.
@@
 Collision-free  locomotion   trajectories  are  usually   obtained  by
-simplifying the model of the  robot or of the environment. By reducing 
+simplifying the model of either the  robot or the environment. By reducing 
 a  humanoid robot to a bounding volume that wraps the swaying motion,
@@
 can be seen as a general way 
-to consider quasi-static multi-step planning. They are not directly applicable to
-humanoid dynamically stable locomotion.
+to consider quasi-static multi-step planning. They are not however directly applicable to
+humanoid dynamically balanced locomotion.
 Other recent contributions to the field of locomotion planning  include algorithms 
@@
 necessary to take into account dynamics while planning, even though the final
-output of our algorithm is a dynamically stable walk motion. This result was first
+output of our algorithm is a dynamically balanced walk motion. This result was first
 presented in \cite{dalibard2011small}.
@@
 constrained sampling-based motion 
-planner and  computes a collision-free  statically stable path  for a robot fixed or sliding 
+planner and  computes a collision-free  statically balanced path  for a robot which can be fixed or sliding 
 on the ground. Another  contribution of this  paper is the formal  proof that
@@
 walking  trajectory. We have implemented this well-grounded 
-method.  The results have been validated on a real HRP-2 robot.
-
-\section{Randomized Motion Planning on Constrained Manifolds}
+method, and the results have been validated on the HRP-2 robot.
+
+\section{Randomized Motion Planning on Constraint Manifolds}
 \label{sec:wb}
+
+FIXME: Avoid the mixup due to the use of task and constraints?
 
@@
   set of inverse kinematics and balance tasks;
-\item the goal is also defined implicitly, by additional inverse kinematic tasks.
+\item the goal is also defined implicitly, by additional inverse kinematics tasks.
 \end{enumerate}
@@
 \begin{itemize}
-\item Static stability: the center of mass of the robot stays above
+\item Static balance: the CoM of the robot stays above
   the support polygon center, the two feet have fixed position and
@@
 \item End-effector position and orientation: the goals of some problems presented
-  in the experimental section of this paper are defined as robot hand configuration,
-  or gaze direction.
+  in the experimental section of this paper are defined as a specific robot hand pose,
+  or a gaze direction.
 \item Configuration task: our adaptation of randomized motion planning algorithms 
@@
 $\mathcal{C}_{free}$. Each iteration of the algorithm attempts to extend the tree
-by adding new vertices in the direction of a randomly-selected configuration
+by adding new vertices in the direction of a randomly selected configuration
 $q_{rand}$. Algorithm~\ref{algRRT} shows the pseudo-code of the RRT algorithm.
@@
 Section~\ref{sec:goal-sampling}  describes the goal manifold sampling, and 
-Section~\ref{sec:extension} the adaptation or RRTs random extensions to constrained planning.
+Section~\ref{sec:extension} the adaptation or RRT random extensions to constrained motion planning.
 
@@
 \item Call the \textit{constraint solver} (Algorithm~\ref{algo:newton}) on this configuration, with the static
-  stability constraints and the goal task.
+  balance constraints and the goal task.
 \item If success, check for collisions.
@@
 
-Fig.~\ref{fig:goal} shows different random configurations respecting stability 
+Fig.~\ref{fig:goal} shows different random configurations enforcing banlance 
 constraints and solving a reaching task for an animated character.
@@
 \caption{Random goal configurations solving a reaching task. All the
-  configurations are stable and collision-free, and the right hand of the character
+  configurations are balanced and collision-free, and the right hand of the character
   reaches the orange ball. Note that the ball is also an obstacle.}
@@
 a ball lying on a shelf, (ii) put it on a higher shelf, (iii) come back to a natural
-rest configuration. The statically stable manifold on which the planning takes place is
+rest configuration. The statical balance manifold on which the planning takes place is
 defined by tasks constraining the robot feet and CoM. The goal of the first problem
+
+FIXME: what follows is not clear, rephrase to say that the hand
+position is fixed and that it can rotate along the vertial axis.
+
 is defined by a hand configuration task (3 positions and 2 orientations), and a gaze task
@@
 \caption{HRP-2 grabs a ball on a shelf, puts it on another shelf, and comes back to 
-  a rest position. Static stability constraints are respected along the path, and 
+  a rest position. Static balance constraints are enforced along the path, and 
   the intermediary goals consisting in grasping and displacing the ball are defined
-  implicitly as inverse kinematic tasks.}
+  implicitly as inverse kinematics tasks.}
 \label{fig:wb-shelves}
@@
 \end{tabular}
-\caption {Experimental results on 20 runs: 3 motion planning problems and 2 goal generations. Times are expressed in seconds.}
+\caption {Experimental results on 20 runs: 3 motion planning problems and 2 goal generations. Time is expressed in seconds.}
 \label{table:reaching}
@@
 The previous section has presented a simple algorithm that solves manipulation planning
-problems on a given constrained manifold of $\mathcal{C}$.
+problems on a given constraint manifold of $\mathcal{C}$.
 
 If we use this algorithm with static 
-stability constraints without fixing globally the robot foot positions, it generates 
-statically stable paths for a robot sliding on the ground. Fig~\ref{fig:sliding} shows
+balance constraints without fixing globally the robot foot positions, it generates 
+statically balanced paths for a robot sliding on the ground. Fig~\ref{fig:sliding} shows
 an example of a whole-body collision-free path for a robot passing between two chairs.
@@
 
-  \caption{Collision-free  statically stable path  for a  humanoid  robot  sliding on  the
+  \caption{Collision-free  statically balanced path  for a  humanoid  robot  sliding on  the
     ground.}
@@
 
-This section presents a \textit{constructive} proof that any such statically stable, 
+This section presents a \textit{constructive} proof that any such statically balanced, 
 collision-free path for a legged robot sliding
-on the ground can be approximated by a collision-free dynamic walk trajectory.
+on the ground can be approximated by a  dynamically balanced, collision-free walk trajectory.
 The proof is based on ideas from control theory, in particular small-space 
-controllability. It also uses the fact that balance criteria for dynamic walk are different
-from the ones for static stability. An important point is that our result and method do not hold
+controllability. It also uses the fact that balance criteria for dynamic walking are different
+from the ones for static balance. An important point is that our result and method do not hold
 for a biped robot that walks in a quasi-static way.
@@
 Section~\ref{sec:ssc} recalls the definition of small-space controllability and its
-use in motion planning. Section~\ref{sec:humanoid-ssc} proves that a legged robot
-dynamically walking is small-space controllable, while a legged robot quasi-statically
-walking is not. Section~\ref{sec:ssc-application} shows how this property is used to
-approximate collision-free statically stable paths by dynamic walk trajectories.
+use in motion planning. Section~\ref{sec:humanoid-ssc} proves that a
+dynamically walking legged robot is small-space controllable, while a quasi-statically
+walking legged robot is not. Section~\ref{sec:ssc-application} shows how this property is used to
+approximate collision-free statically balanced paths by dynamic walking trajectories.
 
@@
   \caption{The small-space controllability local property:  any configuration $q'$ 
-    at distance less than
+    at a distance less than
     $\eta$ is reachable from $q$ by an admissible trajectory included in
@@
 
-The main  consequence of  that property  
+The main  consequence of  this property  
 in  motion planning  is the following theorem, that shows how planning for
 dynamic systems is reduced to geometric planning
-thanks to small-space controllability:
+thanks to the small-space controllability property:
 
@@
 Fig. \ref{fig:ssc2} shows an example of collision-free 
-path approximation by admissible collision-free sub-trajectories. The fact 
-that this algorithm  converges is guaranteed by the small-space
+path approximation by admissible collision-free sub-trajectories. The convergence 
+of this algorithm is guaranteed by the small-space
 controllability property.
@@
   consider in-contact configurations as colliding, so $\mathcal{C}_{free}$ is an open set.
-  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. For all $x \in [0,1]$,
-  $\tau(x) \in \mathcal{C}_{free}$, so there exists $\epsilon_x$ such that the open ball 
+  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. Thus for all $x \in [0,1]$,
+  $\tau(x) \in \mathcal{C}_{free}$, there exists $\epsilon_x$ such that the open ball 
   $B(\tau(x),\epsilon_x)$ of center $\tau(x)$ and radius $\epsilon_x$ is included in 
@@
 
-This section  discusses walking robot small-space controllability. To clarify
+This section  discusses a walking robot small-space controllability. To clarify
 the presentation, we consider a simplified model of a legged robot consisting of two feet
@@
 and the feet are assumed to have a positive surface. For our presentation, it is not 
-necessary to consider the feet height, so the configuration space of the robot is:
+necessary to consider the foot height, so the configuration space of the robot is:
 \[
@@
 \label{thm:humanoid-ssc}
-A quasi-static walking robot is not small-space controllable. A dynamic walking robot is.
+A quasi-statically walking robot is not small-space controllable. A dynamically walking robot is.
 \end{theorem}
@@
 the ball of center $q$ and radius $\epsilon$, is such that the CoM is never over the left foot.
-Given the quasi-static walk conditions, the right foot of the robot is fixed along 
+Given the quasi-static walking conditions, the right foot of the robot is fixed along 
 the trajectory. Thus, the set of accessible configurations from $q$ by staying inside 
@@
 
-Let us now consider a dynamic walking robot. Starting from any valid static configuration, 
+Let us now consider a dynamically walking robot. Starting from any valid static configuration, 
 the CoM can move vertically without affecting balance, so there is no need to 
@@
 we will keep the $x$ coordinates  of the CoM and ZMP constant equal to
-$0$. By hypothesis, the feet are of positive surface,
+$0$. By hypothesis, the feet have a positive surface,
 let $l>0$ be such that the length of the section of a foot along the 
@@
 $\dot{f}(0) = 0$, $f(T) =  1$, $\dot{f}(T) = 0$ and $\ddot{f}(T)
-=  0$.  We can explicitly  construct  such  an $f$   with  a degree  4
-spline.   We   also   request   that   for  all   $t   \in   [0,T]$,
+=  0$.  We can explicitly  construct  such  an $f$   with  a
+spline of degree  4.   We   also   request   that   for  all   $t   \in   [0,T]$,
 $|2\epsilon\dot{f}(t)\frac{\omega}{\omega_0^2}|   \leq  \frac{l}{4}$
@@
 example  of CoM  motion  on the  $y$  axis and  the corresponding  ZMP
-motion. Once in permanent walk in-place state, the robot can come back
+motion. Once in permanent in-place walking state, the robot can come back
 to  a static  state by  applying a  symmetric transient  state  used to
@@
 the previous proof to any legged robot model. First, because we use a model with
-a point mass, we did not have to consider the angular-momentum of the robot
+a point mass, we did not have to consider the angular momentum of the robot
 around the CoM, while it accounts for dynamic balance. Moreover, when considering
 an actual kinematic and dynamic model,
-the positions of feet and CoM are not controlled degrees of freedom, but inputs 
+the positions of feet and CoM are not controlled DoFs, but inputs 
 for inverse kinematic tasks. Continuity and regularity results about the robot
@@
 hardware may be a limitation to such trajectories. To prevent the generated CoM 
-oscillations from being too fast, one has to require that the statically stable 
+oscillations from being too fast, one has to require that the statically balanced 
 path is included
@@
 Given a statically  stable path $p$ respecting the cart-table model approximation constraints,
-we  start  by placing  footsteps  corresponding  to  the nominal  walk
-pattern  of the  robot. Given the footsteps, we  compute a ZMP trajectory, 
+we  start  by placing  footprints  corresponding  to  the nominal  walk
+pattern  of the  robot. Given the footprints, we  compute a ZMP trajectory, 
 and a preview controller outputs  a corresponding CoM trajectory. 
@@
 solving the task, as described in Section~\ref{sec:wb} . Then, we have
-applied our planner to generate a whole-body walk motion that solves the hand reaching
+applied our planner to generate a whole-body walking motion that solves the hand reaching
 task.

### 2012-11-28 script/hpp-wholebody-step-planner/benchmark_analyzer.py
Update script to plot chairs benchmarks.

@@ -18,10 +18,10 @@ sh_ipp = np.genfromtxt("shelf/bench-rrt.log")
 # pca_lt_rrt = np.genfromtxt("bench-pca-lt-rrt.log")
 # pca_lt_ipp = np.genfromtxt("bench-pca-lt-ipp.log")
 
-rrt = [('Floating obstacles RRT', oc_rrt),
-       ('Floating obstacles IPP-RRT', oc_ipp),
-       ('Shelf RRT', sh_ipp),
-       ('Shelf IPP-RRT', sh_ipp)]
+rrt = [('Floating obstacles\n RRT', oc_rrt),
+       ('Floating obstacles\n IPP-RRT', oc_ipp),
+       ('Shelf\n RRT', sh_ipp),
+       ('Shelf\n IPP-RRT', sh_ipp)]
        # ('Local-Trees RRT', lt_rrt),
        # ('Local-Trees IPP-RRT', lt_ipp),
        # ('PCA-RRT', pca_rrt),

### 2012-11-29 paper.tex
Add modified plots and insert them in document.

@@
 \usepackage[pdftex]{graphicx}
+\usepackage{epstopdf}
 \usepackage{tikz}
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.8\linewidth]{plots/rrt-it.eps}
+\caption{Number of RRT iterations $it$ for the floating objects and the
+  shelf scenarios, using two variants of RRT. Mean $\overline{it}$,
+  standard deviation $\sigma_{it}$, minimum and maximum values are
+  represented.}
+\label{fig:rrt-it}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.8\linewidth]{plots/rrt-t.eps}
+\caption{RRT computation time $t$ for the floating objects and the shelf
+  scenarios, using two variants of RRT. Mean $\overline{t}$, standard
+  deviation $\sigma_{t}$, minimum and maximum values are represented.}
+\label{fig:rrt-t}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.8\linewidth]{plots/rrt-n.eps}
+\caption{Number of tree nodes $n$ for the floating objects and the
+  shelf scenarios, using two variants of RRT. Mean $\overline{n}$,
+  standard deviation $\sigma_{n}$, minimum and maximum values are
+  represented.}
+\label{fig:rrt-n}
+\end{figure}
+
 \section{Limits and Discussions}

### 2012-11-29 script/hpp-wholebody-step-planner/benchmark_analyzer.py
Update plot script.

@@ -7,27 +7,21 @@ import matplotlib.pyplot as plt
 from matplotlib.patches import Polygon
 
 # Load benchmark data.
+# ch_rrt = np.genfromtxt("chairs/bench-rrt.log")
+# ch_ipp = np.genfromtxt("chairs/bench-ipp.log")
 oc_rrt = np.genfromtxt("objects-cloud/bench-rrt.log")
 oc_ipp = np.genfromtxt("objects-cloud/bench-ipp.log")
 sh_rrt = np.genfromtxt("shelf/bench-rrt.log")
-sh_ipp = np.genfromtxt("shelf/bench-rrt.log")
-# lt_rrt = np.genfromtxt("bench-lt-rrt.log")
-# pca_rrt = np.genfromtxt("bench-pca-rrt.log")
-# pca_ipp = np.genfromtxt("bench-pca-ipp.log")
-# lt_ipp = np.genfromtxt("bench-lt-ipp.log")
-# pca_lt_rrt = np.genfromtxt("bench-pca-lt-rrt.log")
-# pca_lt_ipp = np.genfromtxt("bench-pca-lt-ipp.log")
+sh_ipp = np.genfromtxt("shelf/bench-ipp.log")
 
-rrt = [('Floating obstacles\n RRT', oc_rrt),
-       ('Floating obstacles\n IPP-RRT', oc_ipp),
-       ('Shelf\n RRT', sh_ipp),
-       ('Shelf\n IPP-RRT', sh_ipp)]
-       # ('Local-Trees RRT', lt_rrt),
-       # ('Local-Trees IPP-RRT', lt_ipp),
-       # ('PCA-RRT', pca_rrt),
-       # ('PCA-IPP-RRT', pca_ipp),
-       # ('PCA-LT-RRT', pca_lt_rrt),
-       # ('PCA-LT-IPP-RRT', pca_lt_ipp)]
+rrt = [
+    # ('Chairs\n RRT', ch_rrt),
+    # ('Chairs\n IPP-RRT', ch_ipp),
+    ('Floating\n RRT', oc_rrt),
+    ('Floating\n IPP-RRT', oc_ipp),
+    ('Shelf\n RRT', sh_rrt),
+    ('Shelf\n IPP-RRT', sh_ipp)
+    ]
 
 # Compute mean and standard deviation over rows (axis 0 of array).
 m_rrt = []
@@ -82,14 +76,22 @@ data_t = data_t.transpose()
 data_n = (np.array) (data_n)
[…]

### 2012-12-04 paper.tex
Use balance term insteand of stability for paths and trajectories.

@@
 for humanoid robots. First, we present a randomized algorithm for constrained motion
-planning, that is used to generate collision-free statically stable paths solving
+planning, that is used to generate collision-free statically balanced paths solving
 manipulation tasks. Then, we show that dynamic walking makes humanoid robots 
@@
 
-In this paper, we present a planning algorithm that considers exact models of a humanoid
+\textbf{In this paper, we present a planning algorithm that considers exact models of a humanoid
 robot and its environment. It is used to solve navigation and manipulation problems. Our planner 
-generates collision-free, statically stable 
-geometric paths. We also present a constructive formal proof that these paths can be approximated by
-collision-free, dynamically balanced walk trajectories. Our planner is designed for indoor
+generates collision-free, statically balanced
+geometric paths, i.e. paths such that for any path configurations, the robot is balanced on its
+feet with zero speed. We also present a constructive formal proof that these paths can be
+approximated by collision-free, dynamically balanced walk trajectories, i.e. trajectories such
+that for any trajectory instant, the robot is balanced on at least one foot.} Our planner is designed for indoor
 environments, where the floor is horizontal and flat. Because we do not compute explicitly footstep
@@
 solvers can animate  the whole set of DoFs of the
-robot to generate a dynamically stable walk trajectory.
+robot to generate a dynamically balanced walk trajectory.
 
@@
 
-\section{From Statically Stable Paths to Dynamic Walk Trajectories}
+\section{From Statically Balanced Paths to Dynamic Walk Trajectories}
 \label{sec:wb-step}
@@
 as well as the vertical orientation of the robot waist. These geometric constraints 
-are also applied when planning statically stable paths, to ensure that the paths
+are also applied when planning statically balanced paths, to ensure that the paths
 can be approximated by dynamic walk trajectories. Note that this is
@@
 
-\subsection{Application: Dynamic Approximation of a Statically Stable Sliding Path}
+\subsection{Application: Dynamic Approximation of a Statically Balanced Sliding Path}
 \label{sec:ssc-application}
 
-The  algorithm   that  animates  a  statically  stable   path  into  a
-dynamically stable  walk trajectory has been inspired  by the previous
+The  algorithm   that  animates  a  statically  balanced   path  into  a
+dynamically balanced  walk trajectory has been inspired  by the previous
 small-space controllability  proof.
-Given a statically  stable path $p$ respecting the cart-table model approximation constraints,
+Given a statically  balanced path $p$ respecting the cart-table model approximation constraints,
 we  start  by placing  footprints  corresponding  to  the nominal  walk
@@
 
-Tasks (1)  and (2) generate a  dynamically stable motion  by using the
+Tasks (1)  and (2) generate a  dynamically balanced motion  by using the
 simplified cart-table model  and the ZMP criterion. Tasks  (3) and (4)

### 2012-12-04 paper.tex
Add details about benchmark tests.

@@
 
-The motion planning algorithms presented in this paper have been implemented
-using KineoWorks\texttrademark \cite{laumond2006kcs}. The planning times have been measured
-on an Intel Core~2~Duo 2.13~GHz PC with 2~GB of RAM. Evaluation of the
-randomized algorithm has been conducted by executing 50 trials on each
-of the first two problems. We present the average results of these trials.
+\textbf{The motion planning algorithms presented in this paper have been
+implemented using KineoWorks\texttrademark \cite{laumond2006kcs}. The
+planning times have been measured on an Intel Core~2~Duo 2.13~GHz PC
+with 2~GB of RAM. Evaluation of the randomized algorithm has been
+conducted by executing 500 trials on each scenario using two flavors of
+RRT: the classic RRT and IPP-RRT \cite{FERR04A}. We present the
+results in Fig. \ref{fig:rrt-it}, \ref{fig:rrt-t}, \ref{fig:rrt-n}, as
+well as in Extension 4 for the raw data.}
 

### 2012-12-04 paper.tex
Respond to review concerning perfectly known environments.

@@
 approximated by collision-free, dynamically balanced walk trajectories, i.e. trajectories such
-that for any trajectory instant, the robot is balanced on at least one foot.} Our planner is designed for indoor
-environments, where the floor is horizontal and flat. Because we do not compute explicitly footstep
+that for any trajectory instant, the robot is balanced on at least one foot. Our planner is designed for perfectly modeled indoor
+environments, where the floor is horizontal and flat. Also, because we do not compute explicitly footstep
 positions at the planning stage, the presented planner is unable to plan motions in which the
-robot steps over obstacles. This limitation is discussed in the paper.
+robot steps over obstacles. These limitations are discussed in the paper.}
 
@@
 
-The presented experiment setup assumes perfect knowledge of the environment. This can
-be guaranteed during experiments by using calibrated objects and motion capture systems.
-This allows us to focus on complex motion planning problems. The perception problem,
-interesting as it is, is completely decoupled from the planning problem in our
-presentation.
+\textbf{The presented experiment setup assumes perfect knowledge of
+  the environment. This can be guaranteed during experiments by using
+  calibrated objects and motion capture systems. This indeed allows us
+  to focus on complex motion planning problems. The perception
+  problem, interesting as it is, is thus completely decoupled from the
+  planning problem in our presentation.}
+
+\textbf{The constrained motion planner and the small-space
+  controllability property can luckily still be used with different
+  environment representations that might more suited with perception
+  systems and their inherent measurement errors, using for instance
+  point clouds. In the future, we would like to achieve dynamic walk
+  planning using a Kinect sensor that would be mounted on the humanoid
+  robot. This would relax the a priori environment knowledge
+  constraint and bring us one step closer to reactive planning in
+  unknown environments.}
 
@@
 
-The presented setup also assumes perfect execution of the plan. It can be 
-critical here, since
-non-nominal stepping may cause to drift away from the planned trajectory. Future 
-experiments will  include trajectory following during plan execution.
+The presented setup also assumes perfect execution of the plan. It can
+be critical here, since non-nominal stepping may cause the robot to
+drift away from the planned trajectory. Future experiments will
+include trajectory following during plan execution.
 

### 2012-12-04 paper.tex
Respond to review concerning perfectly known environments.

@@
 
-Simplifying  the environment  consists in  considering obstacles  at a
-footstep   level only.   \cite{kuffner2001footstep,chestnutt2005footstep,kuffner2005motion}
-use   an  A$^{*}$   algorithm  to   find
-collision-free   footsteps.    In   \cite{perrin2012fast},   the   authors   compute
-dynamic walking motions avoiding collisions at the leg level
-by using  an RRT algorithm.
+\textbf{One possible simplification of the environment consists in
+  considering obstacles at a footstep level only.
+  \cite{kuffner2001footstep,chestnutt2005footstep,kuffner2005motion}
+  use an A$^{*}$ algorithm to find collision-free footsteps.}  In
+\cite{perrin2012fast}, the authors compute dynamic walking motions
+avoiding collisions at the leg level by using an RRT algorithm.
 

### 2012-12-04 paper.tex
Adress reviewer issue concerning the definition of the conifguration space.

@@
 \newtheorem{theorem}{Theorem}
+\renewcommand{\labelitemi}{$-$}
+\newcommand\goalmanifold{\mathcal{M}_{g}}
 
@@
 used on humanoid robotic platforms so far.
-In our work, we show that it is not always
-necessary to take into account dynamics while planning, even though the final
-output of our algorithm is a dynamically balanced walk motion. This result was first
-presented in \cite{dalibard2011small}.
+
+\textbf{In this work, we show that, under some assumptions, it is
+  sufficient to plan in the configuration space in a first step, even
+  though the final output of our algorithm is guaranteed to be a
+  dynamically balanced walk motion. This result was first presented in
+  \cite{dalibard2011small}.}
 
@@
 
-The main contribution of this work is a  whole-body motion planner for humanoid robots
-that computes collision-free walking trajectories, based on
-exact models of both robot and environment. It is used to solve manipulation tasks
-that may require walking. The first stage of our algorithm uses a
-constrained sampling-based motion 
-planner and  computes a collision-free  statically balanced path  for a robot which can be fixed or sliding 
-on the ground. Another  contribution of this  paper is the formal  proof that
-this first path can always be  approximated by a dynamically balanced, collision-free
-walking  trajectory. We have implemented this well-grounded 
-method, and the results have been validated on the HRP-2 robot.
+The main contribution of this work is a whole-body motion planner for
+humanoid robots that computes collision-free walking trajectories,
+based on exact models of both robot and environment. It is used to
+solve manipulation tasks that may require walking. The first stage of
+our algorithm uses a sampling-based constrained motion planner and
+computes a collision-free statically balanced path for a robot which
+can be fixed or sliding on the ground.
+
+Another contribution of this paper is the formal proof that this first
+path can always be approximated by a dynamically balanced,
+collision-free walking trajectory. We have implemented this
+well-grounded method, and the results have been validated on the HRP-2
+robot.
 
@@
 
-FIXME: Avoid the mixup due to the use of task and constraints?
-
-This section presents an algorithm for constrained motion planning. 
-The problem solved here
-differs from classic approaches in two ways:
+\textbf{This section presents an algorithm for constrained motion
+  planning in the configuration space $\mathcal{C}$. In the particular
+  case of a humanoid robot which is an under-actuated system,
+  $\mathcal{C}=\mathcal{Q} \times SE(3)$, where $\mathcal{Q}$
+  represents the kinematic tree actuators and $SE(3)$ corresponds to a
+  fictitious floating joint attached to the root of the tree. If the
+  robot has $n$ actuated DoFs, then $dim(\mathcal{Q})=n$ and
+  $dim(\mathcal{C})=n+6$. A configuration $q$ of $\mathcal{C}$ is said
+  to be valid iff, besides being collision-free, it lies on a specific
+  submanifold $\mathcal{M}$ of $\mathcal{C}$. $\mathcal{M}$ is called
+  the planning submanifold.}
+
+The problem solved here differs from classic approaches in two ways:
 \begin{enumerate}
-\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
-  set of inverse kinematics and balance tasks;
-\item the goal is also defined implicitly, by additional inverse kinematics tasks.
+\item the set of valid configurations is defined implicitly, as the
+  set of collision-free configurations satisfying a given set of
+  inverse kinematics balance constraints;
+\item the goal submanifold $\goalmanifold$ is also defined implicitly,
+  by additional inverse kinematics constraints.
 \end{enumerate}

### 2012-12-04 paper.tex
Address review concerning fast dynamic walking motion.

@@
 
+\textbf{Relying on the cart-table model approximation means that the
+  angular momentum induced by arm movements for instance can lead to
+  non dynamically balanced walking motion. We thus implement the ZMP
+  filtering stage proposed in \cite{kajita2003biped} to compute the
+  exact ZMP and take into account the full dynamics of the robot and
+  generate feasible trajectories.}
+
 \paragraph{Speed of CoM:} The theoretical result presented in this section implies

### 2012-12-04 paper.tex
Respond review wrt constraints and tasks (partial).

@@
 \item Shoot a random configuration in $\mathcal{C}$ with uniform distribution.
-\item Call the \textit{constraint solver} (Algorithm~\ref{algo:newton}) on this configuration, with the static
-  balance constraints and the goal task.
+\item \textbf{Call the \textit{constraint solver}
+  (Algorithm~\ref{algo:newton}) on this configuration, with the
+  intersection of the planning and goal manifolds $\mathcal{M} \cap
+  \goalmanifold$.}
 \item If success, check for collisions.
@@
 
-Fig.~\ref{fig:goal} shows different random configurations enforcing banlance 
-constraints and solving a reaching task for an animated character.
+\textbf{Fig.~\ref{fig:goal} shows resulting random configurations
+  enforcing balance ($\mathcal{M}$) and reaching ($\goalmanifold$)
+  constraints for the HRP-2 robot.}
 

### 2012-12-04 paper.tex
Respond to review wrt one-stage Newton-Rapshon, udd reference to Heine-Borel theorem.

--- 2012-12-04 paper.tex
@@
 \renewcommand{\labelitemi}{$-$}
+\newcommand\manifold{\mathcal{M}}
 \newcommand\goalmanifold{\mathcal{M}_{g}}
@@
   to be valid iff, besides being collision-free, it lies on a specific
-  submanifold $\mathcal{M}$ of $\mathcal{C}$. $\mathcal{M}$ is called
+  submanifold $\manifold$ of $\mathcal{C}$. $\manifold$ is called
   the planning submanifold.}
@@
 
-Section~\ref{sec:goal-sampling}  describes the goal manifold sampling, and 
-Section~\ref{sec:extension} the adaptation or RRT random extensions to constrained motion planning.
-
-\subsection{Goal Manifold Sampling}
-\label{sec:goal-sampling}
-
-The way we generate a goal configuration is the following:
-\begin{enumerate}
-\item Shoot a random configuration in $\mathcal{C}$ with uniform distribution.
-\item Call the \textit{constraint solver} (Algorithm~\ref{algo:newton}) on this configuration, with the static
-  balance constraints and the goal task.
-\item If success, check for collisions.
-\end{enumerate}
+Section~\ref{sec:constraint-solver} describes a constraint solver,
+Section~\ref{sec:goal-sampling} the goal manifold sampling, and
+Section~\ref{sec:extension} the adaptation or RRT random extensions to
+constrained motion planning.
+
+\subsection{Multiple Constraints Solver}
+\label{sec:constraint-solver}
+
+\textbf{We show here how a multiple constraints solver works: its
+  purpose is to find the root $q$ of a non-linear $C^1$ function
+  $f(q)$ with a tolerance of $\varepsilon$. If we want to find a
+  configuration on a manifold $\manifold$, $f(q)$ can be defined as a
+  vector function that contains the concatenation of all constraints
+  defining $\manifold$. Note that as the intersection of two or more
+  manifolds is also a manifold, this multi-constraint solver allows us
+  also to generate configurations that lie at the intersection of
+  multiple manifolds.}
+
+\textbf{In Algorithm~\ref{algo:newton} a Newton-Raphson method
+  \cite{bonnans2006numerical}: starting from an initial value of $q$,
+  $q$ is updated iteratively by $- \alpha \left(\frac{\partial
+    f}{\partial q}(q)\right)^{+} f(q)$, where $\alpha$ denotes a gain
+  and $\left(\frac{\partial f}{\partial q}(q)\right)^{+}$ denotes the
+  Moore-Penrose pseudo-inverse of the Jacobian of $f(q)$. The use of
+  an adaptive gain $\alpha$ that increases iteratively from an initial
+  value $\alpha$ to a maximum value $\alpha_{max}$ has proven to avoid
+  overshoots and accelerate convergence. The update rule relies on a
+  real factor $w \in [0,1]$; the greater $w$ is, the faster $\alpha$
+  will reach $\alpha_{max}$. Obviously, the solver convergence depends
+  of the initial value of $q$, and a bad initialization can lead to
+  either slow convergence or failure. A cutoff number of iterations
+  $it_{max}$ is hence introduced to bypass these cases.}
+
+\textbf{In this work, we observed that values of
+  $\varepsilon=10^{-6}$, $\alpha=0.1$, $\alpha_{max}=0.95$ and $w=0.8$
+  lead to good behavior, i.e. fast convergence and low failure rate.}
 
@@
 \begin{algorithmic}
-\STATE $\alpha=.1$, $i=0$, $\alpha_{max} = .95$
-\WHILE{$\|f(q)\| > \varepsilon$ and $i\leq 20$}
+\STATE $i=0$
+\WHILE{$\|f(q)\| > \varepsilon$ and $i\leq it_{max}$}
 \STATE {\color{red} // $\left(.\right)^{+}$ denotes the Moore-Penrose pseudo-inverse}
@@
 \STATE {\color{red}// Make $\alpha$ tend toward $\alpha_{max}$}
-\STATE $\alpha \rightarrow \alpha_{max} - .8*(\alpha_{max} - \alpha)$
+\STATE $\alpha \rightarrow \alpha_{max} - w(\alpha_{max} - \alpha)$
 \ENDWHILE
@@
 
-Fig.~\ref{fig:goal} shows different random configurations enforcing banlance 
-constraints and solving a reaching task for an animated character.
+\subsection{Goal Manifold Sampling}
+\label{sec:goal-sampling}
+
+The way we generate a goal configuration is the following:
+\begin{enumerate}
+\item \textbf{Shoot a random configuration $q_{rand}$ in $\mathcal{C}$ with
+  uniform distribution.}
+\item \textbf{Call \textit{constraint-solver}
+  (Algorithm~\ref{algo:newton}) on $q_{rand}$, with the
+  intersection of the planning and goal manifolds $\manifold \cap
+  \goalmanifold$.}
+\item If success, check for collisions.
+\end{enumerate}
+
+\textbf{Fig.~\ref{fig:goal} shows resulting random configurations
+  enforcing balance ($\manifold$) and reaching ($\goalmanifold$)
+  constraints for the HRP-2 robot.}
 
@@
 
-The equivalent random  extension on a constrained manifold $\mathcal{M}$, 
+The equivalent random  extension on a constrained manifold $\manifold$, 
 defined by the constraint function $f$,
-starts from a valid configuration $q_{near} \in \mathcal{M}$, and extends the tree
+starts from a valid configuration $q_{near} \in \manifold$, and extends the tree
 towards a random configuration $q_{rand}$, while keeping the constraints
-satisfied. Extension attempts orthogonal to $\mathcal{M}$ are useless,
-as newly added edges have to be included  in $\mathcal{M}$. To extend in directions
-that follow the directions of $\mathcal{M}$, we rely on the framework of 
+satisfied. Extension attempts orthogonal to $\manifold$ are useless,
+as newly added edges have to be included  in $\manifold$. To extend in directions
+that follow the directions of $\manifold$, we rely on the framework of 
 Jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents
@@
 extension. The idea is to first project $q_{rand}$ on the tangent
-space to $\mathcal{M}$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
+space to $\manifold$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
 $q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
@@
 the projection
-of $q_{rand}'$ on $\mathcal{M}$. Instead of extending the tree from $q_{near}$ towards 
+of $q_{rand}'$ on $\manifold$. Instead of extending the tree from $q_{near}$ towards 
 $q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
@@
   code is available.}. While extending the tree,
-the configurations along the new edge are automatically projected onto $\mathcal{M}$. These projections 
+the configurations along the new edge are automatically projected onto $\manifold$. These projections 
 are not very costly if the edge is close to the constrained manifold.
@@
   (88.8934,277.5752) .. controls (196.3490,276.5363) and (532.8532,213.7494) ..
-(434.3656,81.6056) node [above = 0.1cm, color=dg, text width=1.8cm] {\small{Constrained manifold $\mathcal{M}$}};
+(434.3656,81.6056) node [above = 0.1cm, color=dg, text width=1.8cm] {\small{Constrained manifold $\manifold$}};
 \node [draw,circle,inner sep=2pt,fill = gray] (qn) at (97,277) {};
@@
         \node [text width = 2cm,color=dg] at (3.2,4) {\small Constraints defining
-          $\mathcal{M}$};
+          $\manifold$};
         \node [text width = 3cm,text centered] at (3,1.5)
@@
   starting from $q_{near}$:
-  first, the constraints defining $\mathcal{M}$, and at a lower priority
+  first, the constraints defining $\manifold$, and at a lower priority
   a configuration task towards $q_{rand}$.  $q_{rand}'$ is then projected onto
-  $\mathcal{M}$, and a classic RRT extension tries to go as far as possible
+  $\manifold$, and a classic RRT extension tries to go as far as possible
   from $q_{near}$ towards $q_{rand}''$.}
@@
   The set of open balls $\left( B(\tau(x),\eta_x) \right)_{x\in [0,1]}$ forms an open cover
-  of $\tau([0,1])$ which is compact. The Heine-Borel theorem states that there exists a
-  finite subcover $\left( B(\tau(x_i),\eta_{x_i}) \right)_{i\in \{ 1,\dots ,n \}}$ of $\tau([0,1])$. To this
+  of $\tau([0,1])$ which is compact. \textbf{The Heine-Borel theorem \cite{fitzpatrick2006advanced} states that there exists a
+  finite subcover $\left( B(\tau(x_i),\eta_{x_i}) \right)_{i\in \{ 1,\dots ,n \}}$ of $\tau([0,1])$}. To this
   finite subcover corresponds a finite number of feasible trajectories, going from $\tau(0)$ to  

### 2012-12-04 paper.tex
Add modifications wrt previous commit.

@@
   also to generate configurations that lie at the intersection of
-  multiple manifolds.}
+  several manifolds.}
 
@@
 \item \textbf{Call \textit{constraint-solver}
-  (Algorithm~\ref{algo:newton}) on $q_{rand}$, with the
-  intersection of the planning and goal manifolds $\manifold \cap
+  (Algorithm~\ref{algo:newton}) on $q_{rand}$, with $f(q)$ defined by
+  the intersection of the planning and goal manifolds $\manifold \cap
   \goalmanifold$.}

### 2012-12-04 paper.tex
Respond to reviewer wrt table comment.

@@
 \begin{table}
-\begin{tabular}{lrrrr}
+\begin{tabular}{l|r|r|r|r|}
+\cline{2-5}
 & min & max & average & average \\ 
 &&&& per problem \\
-number of nodes & 43.00 & 481.00 & 102.70 \\
-goal generation time \hspace {1.cm} & 1.00 & 1.56 & 1.22\\
-planning time& 67.36 & 376.84 & 134.28 & 44.76\\
+\hline
+\multicolumn{1}{|l|}{number of nodes} & 43.00 & 481.00 & 102.70 \\
+\cline{1-4}
+\multicolumn{1}{|l|}{goal generation time (s)} & 1.00 & 1.56 & 1.22\\
+\hline
+\multicolumn{1}{|l|}{planning time (s)} & 67.36 & 376.84 & 134.28 & 44.76\\
+\hline
 \end{tabular}
-\caption {Experimental results on 20 runs: 3 motion planning problems and 2 goal generations. Time is expressed in seconds.}
+\caption {\textbf{Experimental results on 20 runs: Each run consists
+    of 3 motion planning problems and 2 goal generations for the three
+    phases. Time is expressed in seconds.}}
 \label{table:reaching}

### 2012-12-04 paper.tex
Improve text and respond to reviewer wrt success rate.

@@
   configurations are balanced and collision-free, and the right hand of the character
-  reaches the orange ball. Note that the ball is also an obstacle.}
+  reaches the orange ball.}
 \label{fig:goal}
@@
 
-We present in Fig.~\ref{fig:wb-shelves} an illustration of 
-the use of randomized motion planning 
-on complex manipulation problems. The humanoid robot HRP-2 faces shelves. It has to: (i) grasp
-a ball lying on a shelf, (ii) put it on a higher shelf, (iii) come back to a natural
-rest configuration. The statical balance manifold on which the planning takes place is
-defined by tasks constraining the robot feet and CoM. The goal of the first problem
-
-FIXME: what follows is not clear, rephrase to say that the hand
-position is fixed and that it can rotate along the vertial axis.
-
-is defined by a hand configuration task (3 positions and 2 orientations), and a gaze task
-(the robot has to look at the ball). The goal of the second problem is defined by a 
-different hand configuration and the gaze task.
-Finally, the goal of the third problem is the rest configuration.
-
-For the two reaching motions, we generate 8 random goal configurations (Section~\ref{sec:goal-sampling}).
-
-Extension 1 presents a video of this motion after an optimization step.
+\textbf{We present in Fig.~\ref{fig:wb-shelves} an illustration of the
+  use of randomized motion planning on complex manipulation
+  problems. The humanoid robot HRP-2 faces shelves. It has to: (i)
+  grasp a ball lying on a shelf, (ii) put it on a higher shelf, (iii)
+  come back to a natural rest configuration. We can hence define three
+  separate constrained motion planning problems where the planning
+  manifold $\manifold$ is the static balance manifold defined in
+  \ref{sec:wb}; the goal manifold of problem (i) is defined by a hand
+  pose constraint (the hand must be horizontal and its position has to
+  coincide with the ball initial position), and a gaze constraint (the
+  robot has to look at the ball in its initial position). Similarly,
+  the goal manifold of problem (ii) is defined by hand and gaze
+  constraints that correspond to the position of the ball on the
+  higher shelf. Finally, we define the rest configuration as the
+  single goal configuration for problem (iii).}
+
+\textbf{Note that for phases (i) and (iii) the ball is also considered
+  as an obstacle. This is necessary to prevent the robot grasping hand
+  from colliding with the ball during the approach (respectively
+  retraction) phase.}
+
+\textbf{For the two reaching motions in (i) and (ii), we first
+  generate 8 random goal configurations
+  (Section~\ref{sec:goal-sampling}), then we solve the three
+  constrained motion planning problems separately. As randomized
+  motion planning algorithms produce, a classic shortcut method is
+  used to optimize and shorten the paths. Extension 1 presents a
+  video of the concatenated motion.}
+
+\textbf{We have run this set of motion planning problems 20 times;
+  results are compiled in Table~\ref{table:reaching}. We have also
+  measured the performance of \texttt{constraint-solver} (Algorithm
+  \ref{algo:newton}) when used to project configurations on
+  $\manifold$; the average number of iterations is 6.5 per call, and
+  the success rate, i.e. the ratio of the number of successfully
+  projected configurations over the total number of calls, is above 95
+  percent.}
 
@@
 
-We have run this set of motion planning problems 20 times. The results are compiled in Table~\ref{table:reaching}.
 \begin{table}
@@
 \end{table}
-
-Among all calls to \texttt{constraint-solver}, the average number of iterations is 6.5 and the success rate is above 95 percent.
 

### 2012-12-04 paper.tex
Add modification to previous commit.

@@
   the intermediary goals consisting in grasping and displacing the ball are defined
-  implicitly as inverse kinematics tasks.}
+  implicitly as inverse kinematics constraints.}
 \label{fig:wb-shelves}

### 2012-12-04 paper.tex
Respond to reviewer wrt unfeasible sliding path.

@@
 
-The previous section has presented a simple algorithm that solves manipulation planning
-problems on a given constraint manifold of $\mathcal{C}$.
-
-If we use this algorithm with static 
-balance constraints without fixing globally the robot foot positions, it generates 
-statically balanced paths for a robot sliding on the ground. Fig~\ref{fig:sliding} shows
-an example of a whole-body collision-free path for a robot passing between two chairs.
-Such paths are physically unfeasible. They are, however, easier to generate than feasible
-dynamic trajectories because only geometric constraints are considered at planning time.
+The previous section has presented a simple algorithm that solves
+manipulation planning problems on a given constraint manifold
+$\manifold$ of $\mathcal{C}$.
+
+If we use this algorithm with static balance constraints without
+fixing globally the robot foot positions, it generates statically
+balanced paths for a robot \textit{sliding} on the
+ground. \textbf{Fig~\ref{fig:sliding} shows an example of a whole-body
+  collision-free path for a robot passing between two chairs. Since in
+  reality a legged robot cannot slide on a regular floor, such paths
+  are physically unfeasible.They are, however, easier to generate than
+  feasible dynamic walking trajectories because only geometric
+  constraints are considered at planning time.}
 

### 2012-12-04 stc1.tex
Increase small-space figure size and add balls labels.

@@
 \definecolor{cffffff}{RGB}{255,255,255}
+\definecolor{darkGrey}{RGB}{75,75,75}
 
-\begin{tikzpicture}[y=0.4pt, x=0.4pt,yscale=-1, inner sep=0pt, outer sep=0pt]
+\begin{tikzpicture}[y=0.7pt, x=0.7pt,yscale=-1, inner sep=0pt, outer sep=0pt]
 % path5273
@@
 
+% Ball epsilon q
+\path[fill=black] (330,200) node[above right] (text2834) {$\mathcal{B}_{\epsilon}(q)$};
+
 % text2837
@@
 
+% Ball eta q
+\path[fill=black] (320,258) node[above right, color=darkGrey] (text2838) {$\mathcal{B}_{\eta}(q)$};
+
 \end{tikzpicture}

### 2012-12-04 paper.tex
Respond to review wrt experiment details.

@@
 
+\textbf{Our whole-body motion planner generates a robot configuration
+  trajectory that is sampled at a 200~Hz rate and stored in a
+  file. This file can then be used to play the trajectory in open-loop
+  on the HRP-2 robot, which is position-controlled. Scenarios in
+  Sections \ref{sec:chairs} and \ref{sec:shelf} were both successfully
+  executed.}
+
 \subsection{Passing between two chairs}
+\label{sec:chairs}
 
@@
 \subsection{'Put the ball on a shelf'}
+\label{sec:shelf}
 

### 2012-12-05 paper.tex
Respond to reviewer wrt equality constraints.
Respond to reviewer wrt uniform sampling.

@@
 \end{enumerate}
-During global planning, we will consider several types of
-tasks and constraints for various reasons:
+\textbf{During global planning, we will consider several types of
+equality constraints for various reasons:}
 \begin{itemize}
@@
   projected configurations over the total number of calls, is above 95
-  percent.}
+  percent. This success rate, high as it is, could be further improved
+  by sampling a better initial configuration of $\mathcal{C}$, for
+  example by introducing a heuristic bias towards statically balanced
+  configurations. Nevertheless we choose to sample $\mathcal{C}$
+  uniformly-randomly for the sake of genericity.}
 

### 2012-12-05 paper.tex
Add multimedia extensions in appendix.

@@
 
+\usepackage{float}
+\restylefloat{table}
 \usepackage{array} 
@@
 
+\appendix
+\section{Index to Multimedia Extensions}
+
+\begin{table}[H]
+\begin{tabular}{c c m{0.7\linewidth}}
+Extension & Type & Description \\
+\hline
+1 & Video & Example of a constrained motion planning result executed
+on HRP-2. \\
+2 & Video & Chairs scenario: solution of whole-body planning executed
+on HRP-2. \\
+3 & Video & Shelf scenario: solution of whole-body planning executed
+in simulation and on HRP-2. \\
+4 & Data & Raw data of sliding motion planning benchmarks for all
+scenarios. \\
+\end{tabular}
+\end{table}
+
 \bibliographystyle{agsm}

### 2012-12-05 paper.tex
Update table format.
Move benchmark plots to appendix.

@@
 \usepackage{float}
-\restylefloat{table}
 \usepackage{array} 
@@
 
-\textbf{The motion planning algorithms presented in this paper have been
-implemented using KineoWorks\texttrademark \cite{laumond2006kcs}. The
-planning times have been measured on an Intel Core~2~Duo 2.13~GHz PC
-with 2~GB of RAM. Evaluation of the randomized algorithm has been
-conducted by executing 500 trials on each scenario using two flavors of
-RRT: the classic RRT and IPP-RRT \cite{FERR04A}. We present the
-results in Fig. \ref{fig:rrt-it}, \ref{fig:rrt-t}, \ref{fig:rrt-n}, as
-well as in Extension 4 for the raw data.}
+\textbf{The motion planning algorithms presented in this paper have
+  been implemented using KineoWorks\texttrademark
+  \cite{laumond2006kcs}. The planning times have been measured on an
+  Intel Core~2~Duo 2.13~GHz PC with 2~GB of RAM. Evaluation of the
+  randomized algorithm has been conducted by executing 500 trials on
+  each scenario using two flavors of RRT: the classic RRT and IPP-RRT
+  \cite{FERR04A}. We present the results in Appendix \ref{app:bench},
+  Fig. \ref{fig:rrt-it}, \ref{fig:rrt-t}, \ref{fig:rrt-n}, as well as
+  in Extension 4 for the raw data.}
 
@@
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.8\linewidth]{plots/rrt-it.eps}
-\caption{Number of RRT iterations $it$ for the floating objects and the
-  shelf scenarios, using two variants of RRT. Mean $\overline{it}$,
-  standard deviation $\sigma_{it}$, minimum and maximum values are
-  represented.}
-\label{fig:rrt-it}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.8\linewidth]{plots/rrt-t.eps}
-\caption{RRT computation time $t$ for the floating objects and the shelf
-  scenarios, using two variants of RRT. Mean $\overline{t}$, standard
-  deviation $\sigma_{t}$, minimum and maximum values are represented.}
-\label{fig:rrt-t}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.8\linewidth]{plots/rrt-n.eps}
-\caption{Number of tree nodes $n$ for the floating objects and the
-  shelf scenarios, using two variants of RRT. Mean $\overline{n}$,
-  standard deviation $\sigma_{n}$, minimum and maximum values are
-  represented.}
-\label{fig:rrt-n}
-\end{figure}
-
 \section{Limits and Discussions}
@@
 \appendix
+\section{Sliding Motion Planning Benchmarks}
+\label{app:bench}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.8\linewidth]{plots/rrt-it.eps}
+\caption{Number of RRT iterations $it$ for the floating objects and the
+  shelf scenarios, using two variants of RRT. Mean $\overline{it}$,
+  standard deviation $\sigma_{it}$, minimum and maximum values are
+  represented.}
+\label{fig:rrt-it}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.8\linewidth]{plots/rrt-t.eps}
+\caption{RRT computation time $t$ for the floating objects and the shelf
+  scenarios, using two variants of RRT. Mean $\overline{t}$, standard
+  deviation $\sigma_{t}$, minimum and maximum values are represented.}
+\label{fig:rrt-t}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.8\linewidth]{plots/rrt-n.eps}
+\caption{Number of tree nodes $n$ for the floating objects and the
+  shelf scenarios, using two variants of RRT. Mean $\overline{n}$,
+  standard deviation $\sigma_{n}$, minimum and maximum values are
+  represented.}
+\label{fig:rrt-n}
+\end{figure}
+
 \section{Index to Multimedia Extensions}
@@
 \begin{tabular}{c c m{0.7\linewidth}}
+\hline
 Extension & Type & Description \\
@@
 scenarios. \\
+\hline
 \end{tabular}

### 2012-12-05 paper.tex
Minor revision.

@@
 
-Section~\ref{sec:constraint-solver} describes a constraint solver,
+Section~\ref{sec:SolveConstraints} describes a constraint solver,
 Section~\ref{sec:goal-sampling} the goal manifold sampling, and
@@
 
-\subsection{Multiple Constraints Solver}
-\label{sec:constraint-solver}
+\subsection{Constraint Solver}
+\label{sec:SolveConstraints}
 
@@
 
-\textbf{In Algorithm~\ref{algo:newton} a Newton-Raphson method
+\textbf{Algorithm~\ref{algo:newton} implements a Newton-Raphson method
   \cite{bonnans2006numerical}: starting from an initial value of $q$,
@@
   Moore-Penrose pseudo-inverse of the Jacobian of $f(q)$. The use of
-  an adaptive gain $\alpha$ that increases iteratively from an initial
-  value $\alpha$ to a maximum value $\alpha_{max}$ has proven to avoid
-  overshoots and accelerate convergence. The update rule relies on a
-  real factor $w \in [0,1]$; the greater $w$ is, the faster $\alpha$
-  will reach $\alpha_{max}$. Obviously, the solver convergence depends
-  of the initial value of $q$, and a bad initialization can lead to
-  either slow convergence or failure. A cutoff number of iterations
-  $it_{max}$ is hence introduced to bypass these cases.}
-
-\textbf{In this work, we observed that values of
-  $\varepsilon=10^{-6}$, $\alpha=0.1$, $\alpha_{max}=0.95$ and $w=0.8$
-  lead to good behavior, i.e. fast convergence and low failure rate.}
+  an adaptive gain $\alpha$, which increases iteratively from an
+  initial value $\alpha$ to a maximum value $\alpha_{max}$, allows the
+  overshoot avoidance and convergence acceleration. The update rule
+  relies on a real factor $w \in [0,1]$; the greater $w$ is, the
+  faster $\alpha$ will reach $\alpha_{max}$. Obviously, the solver
+  convergence depends on the initial value of $q$, and a bad
+  initialization can lead to either slow convergence or failure. A
+  cutoff number of iterations $it_{max}$ is hence introduced to bypass
+  these cases.}
+
+\textbf{We observed that values of $\varepsilon=10^{-6}$,
+  $\alpha=0.1$, $\alpha_{max}=0.95$ and $w=0.8$ lead to good behavior,
+  i.e. fast convergence and low failure rate. These values are kept
+  constant for all scenarios in this work.}
 
 \begin{algorithm}
-\caption{\texttt{constraint-solver}($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
+\caption{\texttt{SolveConstraints}($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
 \label{algo:newton}
@@
   uniform distribution.}
-\item \textbf{Call \textit{constraint-solver}
-  (Algorithm~\ref{algo:newton}) on $q_{rand}$, with $f(q)$ defined by
-  the intersection of the planning and goal manifolds $\manifold \cap
-  \goalmanifold$.}
+\item \textbf{Call SolveConstraints (Algorithm~\ref{algo:newton}) on
+  $q_{rand}$, with $f(q)$ defined by the intersection of the planning
+  and goal manifolds $\manifold \cap \goalmanifold$.}
 \item If success, check for collisions.
@@
 \textbf{Fig.~\ref{fig:goal} shows resulting random configurations
-  enforcing balance ($\manifold$) and reaching ($\goalmanifold$)
-  constraints for the HRP-2 robot.}
+  which satisfy both balance ($\manifold$) and reaching
+  ($\goalmanifold$) constraints for the HRP-2 robot.}
 
@@
 $q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
-$q_{rand}''$  be the result of a call to $\texttt{constraint-solver}(q_{rand}',f,\epsilon)$.  It is 
+$q_{rand}''$  be the result of a call to $\texttt{SolveConstraints}(q_{rand}',f,\epsilon)$.  It is 
 the projection
@@
 \begin{algorithm}[h]
-  \caption{Constrained-Extend($\mathcal{T},q_{near},q_{rand},f,\epsilon$)}
+  \caption{ConstrainedExtend($\mathcal{T},q_{near},q_{rand},f,\epsilon$)}
   \label{alg:constrained}
@@
     \frac{\partial f}{\partial q}(q_{near})\right) (q_{rand} - q_{near})$
-    \IF {$\texttt{constraint-solver}(q_{rand}',f,\epsilon) = q_{rand}''$} 
+    \IF {$\texttt{SolveConstraints}(q_{rand}',f,\epsilon) = q_{rand}''$}
     \STATE RRT::Extend($\mathcal{T},q_{near},q_{rand}''$)
@@
   results are compiled in Table~\ref{table:reaching}. We have also
-  measured the performance of \texttt{constraint-solver} (Algorithm
+  measured the performance of \texttt{SolveConstraints} (Algorithm
   \ref{algo:newton}) when used to project configurations on

### 2012-12-05 paper.tex
Minor revision.

@@
 \begin{algorithm}
-\caption{\texttt{SolveConstraints}($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
+\caption{\texttt{SolveConstraints}($q$, f, $\varepsilon$): find $q$
+  such that $f(q) = 0$}
 \label{algo:newton}
@@
   uniform distribution.}
-\item \textbf{Call SolveConstraints (Algorithm~\ref{algo:newton}) on
+\item \textbf{Call \texttt{SolveConstraints} (Algorithm~\ref{algo:newton}) on
   $q_{rand}$, with $f(q)$ defined by the intersection of the planning
@@
 \begin{algorithm}[h]
-  \caption{ConstrainedExtend($\mathcal{T},q_{near},q_{rand},f,\epsilon$)}
+  \caption{\texttt{ConstrainedExtend}($\mathcal{T},q_{near},q_{rand},f,\epsilon$)}
   \label{alg:constrained}

### 2012-12-05 paper.tex
Respond to review wrt stack of tasks.

@@
 The equivalent random extension on a constrained manifold $\manifold$,
-defined by the constraint function $f$,
-starts from a valid configuration $q_{near} \in \manifold$, and extends the tree
-towards a random configuration $q_{rand}$, while keeping the constraints
+defined by the constraint function $f$, starts from a valid
+configuration $q_{near} \in \manifold$, and extends the tree towards a
+random configuration $q_{rand}$, while keeping the constraints
 satisfied. Extension attempts orthogonal to $\manifold$ are useless,
-as newly added edges have to be included  in $\manifold$. To extend in directions
-that follow the directions of $\manifold$, we rely on the framework of 
-Jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents
-the adaptation of the classic extend function, and Fig.~\ref{fig:gikrrt} illustrates this
-extension. The idea is to first project $q_{rand}$ on the tangent
-space to $\manifold$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
-$q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
-$q_{rand}''$  be the result of a call to $\texttt{SolveConstraints}(q_{rand}',f,\epsilon)$.  It is 
-the projection
-of $q_{rand}'$ on $\manifold$. Instead of extending the tree from $q_{near}$ towards 
-$q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
-$q_{rand}''$\footnote{This presentation attempts to give a precise idea of the algorithm, without 
-  focusing on technical details. Readers interested in the actual implementation can
-  refer to \url{https://github.com/laas/hpp-constrained} and 
-  \url{https://github.com/laas/hpp-constrained-planner}, where the corresponding open-source 
-  code is available.}. While extending the tree,
-the configurations along the new edge are automatically projected onto $\manifold$. These projections 
-are not very costly if the edge is close to the constrained manifold.
+as newly added edges have to be included in $\manifold$. To extend in
+directions that follow the directions of $\manifold$, we rely on
+Jacobian-based inverse kinematics. Algorithm~\ref{alg:constrained}
+presents the adaptation of the classic extend function, and
+Fig.~\ref{fig:gikrrt} illustrates this extension. The idea is to first
+project $q_{rand}$ on the tangent space to $\manifold$ at
+$q_{near}$. Let us call the projected configuration $q_{rand}'$. Let
+$q_{rand}''$ be the result of a call to
+$\texttt{SolveConstraints}(q_{rand}',f,\epsilon)$.  It is the
+projection of $q_{rand}'$ on $\manifold$. Instead of extending the
+tree from $q_{near}$ towards $q_{rand}$, the algorithm tries to extend
+from $q_{near}$ towards $q_{rand}''$ while remaining on
+$\manifold$ \footnote{This presentation attempts to give a precise
+  idea of the algorithm, without focusing on technical
+  details. Readers interested in the actual implementation can refer
+  to \url{https://github.com/laas/hpp-constrained} and
+  \url{https://github.com/laas/hpp-constrained-planner}, where the
+  corresponding open-source code is available.}. While extending the
+tree, the configurations along the new edge are automatically
+projected onto $\manifold$. These projections are not very costly if
+the edge is close to the constrained manifold.
 
@@
 \begin{minipage}[c]{0.3\linewidth}
-\begin{tikzpicture}[x=0.61cm,y=0.61cm]
-   \definecolor{dg}{rgb}{0,0.3,0}
-      %Stack of tasks:
-      \draw [color=black, thick,fill = white] (0,1) rectangle (2,2);
-      \draw [color=black, thick,fill = white] (0,2) rectangle (2,3);
-      \draw [color=black, thick,fill = white] (0,3) rectangle (2,5);
-      \draw [color=black, thick,fill = white] (0,5) rectangle (2,6);
-      \draw [color=black, thick,fill = white] (0,6) rectangle (5,7);
-      \draw [color=black, thick,fill = white] (1,1) rectangle (5,6);
-      \draw [thick] (1,1) -- (5,1) -- (5,6);
-      \draw (1,2) -- (5,2) ;
-      \node [text width = 5cm,text centered] at (2.5,6.5) {\large Stack
-        of tasks};
-      \begin{scriptsize}
-        \node at (0.5,5.5) {1};
-        \node at (0.5,4) {$\vdots$}; 
-        \node at (0.5,2.5) {$n$};
-        \node at (0.5,1.5) {$n$+1};
-
-        \node [text width = 2cm,color=dg] at (3.2,4) {\small Constraints defining
-          $\manifold$};
-        \node [text width = 3cm,text centered] at (3,1.5)
-              { Configuration task towards $q_{rand}$};    
-      \end{scriptsize}
-
-\end{tikzpicture}
+%% \begin{tikzpicture}[x=0.61cm,y=0.61cm]
+%%    \definecolor{dg}{rgb}{0,0.3,0}
+%%       %Stack of tasks:
+%%       \draw [color=black, thick,fill = white] (0,1) rectangle (2,2);
+%%       \draw [color=black, thick,fill = white] (0,2) rectangle (2,3);
+%%       \draw [color=black, thick,fill = white] (0,3) rectangle (2,5);
+%%       \draw [color=black, thick,fill = white] (0,5) rectangle (2,6);
+%%       \draw [color=black, thick,fill = white] (0,6) rectangle (5,7);
+%%       \draw [color=black, thick,fill = white] (1,1) rectangle (5,6);
+%%       \draw [thick] (1,1) -- (5,1) -- (5,6);
+%%       \draw (1,2) -- (5,2) ;
+%%       \node [text width = 5cm,text centered] at (2.5,6.5) {\large Stack
+%%         of tasks};
+%%       \begin{scriptsize}
+%%         \node at (0.5,5.5) {1};
+%%         \node at (0.5,4) {$\vdots$}; 
+%%         \node at (0.5,2.5) {$n$};
+%%         \node at (0.5,1.5) {$n$+1};
+
+%%         \node [text width = 2cm,color=dg] at (3.2,4) {\small Constraints defining
+%%           $\manifold$};
+%%         \node [text width = 3cm,text centered] at (3,1.5)
+%%               { Configuration task towards $q_{rand}$};    
+%%       \end{scriptsize}
+
+%% \end{tikzpicture}
 \end{minipage}
 
-\caption{One step of constrained extension, and the stack of tasks used
-  to generate $q_{rand}'$ with a prioritized inverse kinematics solver, 
-  starting from $q_{near}$:
-  first, the constraints defining $\manifold$, and at a lower priority
-  a configuration task towards $q_{rand}$.  $q_{rand}'$ is then projected onto
-  $\manifold$, and a classic RRT extension tries to go as far as possible
-  from $q_{near}$ towards $q_{rand}''$.}
+\caption{\textbf{One step of constrained extension illustrating
+    Algorithm \ref{alg:constrained}: $q_{rand}$ is first projected on
+    the tangent space of $\manifold$. $q_{rand}'$ is then projected
+    onto $\manifold$, and a classic RRT extension tries to go as far
+    as possible from $q_{near}$ towards $q_{rand}''$ while remaining
+    on $\manifold$. $q_{new}$ is then returned.}}
 \label{fig:gikrrt}
@@
 dynamically balanced walk trajectory has been inspired by the previous
-small-space controllability  proof.
-Given a statically  balanced path $p$ respecting the cart-table model approximation constraints,
+small-space controllability proof. \textbf{Given a statically balanced
+  path $p$ verifying the cart-table model approximation constraints,
   we start by placing footprints corresponding to the nominal walk
-pattern  of the  robot. Given the footprints, we  compute a ZMP trajectory, 
-and a preview controller outputs  a corresponding CoM trajectory. 
-The stack of tasks applied to
-the robot to generate a dynamic walk motion is -- in decreasing priority order:
+  pattern of the robot. Given the footprints, we compute a ZMP
+  trajectory, derive foot trajectories, and a preview controller
+  outputs a corresponding CoM trajectory. Classic numerical
+  Jacobian-based prioritized inverse kinematics methods prove to be
+  very useful to generate a dynamic walking trajectory while trying to
+  accomplish secondary tasks, such as following a reference
+  configuration trajectory. We use the same framework which was
+  developped in \cite{yoshida2006tds}.}
+
+The stack of tasks applied to the robot to generate a dynamic walking
+motion is -- in decreasing priority order:
 
@@
 
-\STATE $DynamicTrajectory \leftarrow \text{Animate}(StackOfTasks)$
+\STATE $DynamicTrajectory \leftarrow
+\text{ComputeWalkTrajectory}(StackOfTasks)$
 

### 2012-12-05 paper.tex
Minor revision.

@@
   defining $\manifold$. Note that as the intersection of two or more
-  manifolds is also a manifold, this multi-constraint solver allows us
-  also to generate configurations that lie at the intersection of
-  several manifolds.}
+  manifolds is also a manifold, this constraint solver allows us also
+  to generate configurations that lie at the intersection of several
+  manifolds.}
 
@@
   configuration trajectory. We use the same framework which was
-  developped in \cite{yoshida2006tds}.}
+  developed in \cite{yoshida2006tds}.}
 

### 2012-12-05 paper.tex
Update and add computation times.

@@
 forward, which may be required if the robot has to use vision during
-locomotion. The first planning stage  requires 29.6~s
+locomotion. The first planning stage  requires 1~s
 on  average.   The  animation  of   the  sliding  path   presented  in
@@
 The first planning stage requires
-184.3~s on average, and the animation of the trajectory presented in 
+53~s on average, and the animation of the trajectory presented in 
 Fig.~\ref{fig:cluttered} uses 339.5~s of computation time. Fig.~\ref{fig:cluttered-waist} 
@@
 
-Fig.~\ref{fig:shelf} shows the resulting walk motion and Fig.~\ref{fig:shelf-waist} 
-shows the robot CoM trajectory during locomotion. Extension 3 presents a 
-comprehensive video of this problem, including the motion execution on
-the real robot HRP-2 on stage. Fig.~\ref{fig:shelf-cdf} shows some snapshots 
-taken from Extension 3.
+\textbf{The first planning stage requires 15~s on average, and the
+  animation of the resulting walk motion presented in
+  Fig.~\ref{fig:shelf} requires around 190~s of computation time.}
+Fig.~\ref{fig:shelf-waist} shows the robot CoM trajectory during
+locomotion. Extension 3 presents a comprehensive video of this
+problem, including the motion execution on the real robot HRP-2 on
+stage. Fig.~\ref{fig:shelf-cdf} shows some snapshots taken from
+Extension 3.
 

### 2012-12-05 paper.tex
Add RRT* reference
Minor changes following reviewer comments

@@
 the representation indicates the difficulty of the considered problem. The simplest
-option consists in considering the robot as a navigating 2D shape, and computing 
+option consists in considering the robot as a navigating 2D shape \cite{pettre20032}, and computing 
 obstacle avoidance in a planar model of the world. Another possibility is to 
-compute only collision-free footsteps. 
+compute only collision-free footsteps  \cite{kuffner2001footstep}. 
 In complex and difficult environments, such as the one presented in Fig.~\ref{fig:couv},
@@
 There are two main ways of using motion planners to generate dynamically balanced robotic motions.
-The more general one is to plan in a robot dynamic space. By taking into account both robot
+The more general one is to plan in a robot dynamic space, see for example \cite{shkolnik2011bounding}. 
+By taking into account both robot
 configuration and velocity, motions that satisfy dynamic balance constraints can be generated
@@
 positions. The other way is to first plan  a geometric path that can be approximated by a
-dynamic trajectory in a second step. The approach we present in this paper falls into the second 
+dynamic trajectory in a second step \cite{yoshida-humanoids05}. The approach we present in this paper 
+falls into the second 
 category. Some feasible dynamic motions are inherently impossible to compute with this kind
@@
 planning is needed to overcome this limitation. 
+{\bf Note that when local methods find solutions, these are usually smoother
+and may look more natural. The choice of using global motion planners
+is justified by the need for complete algorithms.}
 \cite{TouGieGoe2007} propose a motion generation method where tasks follow
@@
 a popular randomized motion planning algorithm. 
+{\bf Our method could be applied to any randomized sampling planning method, such
+as RRT\up{*} \cite{Karaman01062011}, which is slower than RRT but generates optimal
+paths.}
 
@@
 
+{\bf \cite{berenson2011task} presents a formal proof that projection-based 
+constrained random motion planning on a fixed dimension manifold is 
+probabilisticaly complete. This proof applies to our algorithm. }
+
 \begin{figure}[h]

### 2012-12-05 paper.tex
Fix typo

@@
 {\bf Our method could be applied to any randomized sampling planning method, such
-as RRT\up{*} \cite{Karaman01062011}, which is slower than RRT but generates optimal
+as RRT\textsuperscript{*} \cite{Karaman01062011}, which is slower than RRT but generates optimal
 paths.}

### 2012-12-06 paper.tex
Remove reference to colors in plot.

@@
 
-\caption{CoM motion (in plain red) along $y$ axis.  The CoM stays in the interval
+\caption{CoM motion (solid line) along $y$ axis.  The CoM stays in the interval
   $[-\epsilon,\epsilon]$ while during  permanent state ($t \geq T$),
-  the ZMP (dashed blue) oscillates between the centers of the feet, which allows
+  the ZMP (dashed line) oscillates between the centers of the feet, which allows
   in-place walk.}

### 2012-12-06 paper.tex
Update CNRS affiliation.

@@
   Jean-Paul Laumond
-  \footnote{The authors are with CNRS ; LAAS ; 7 avenue du colonel Roche, F-31077
-    Toulouse Cedex 4, France and Universit\'e de Toulouse ; UPS, INSA, INP, ISAE ;
-    UT1, UTM, LAAS ; F-31077 Toulouse
-    Cedex 4, France. S\'ebastien Dalibard is now with Nanyang Technological University, Singapore.}
+  \footnote{The authors are with CNRS, LAAS, 7 avenue du colonel
+    Roche, F-31400 Toulouse Cedex 4, France and Univ de Toulouse,
+    LAAS, F-31400 Toulouse Cedex 4, France. S\'ebastien Dalibard is
+    now with Nanyang Technological University, Singapore.}
   \footnote{

### 2012-12-06 paper.tex
Fix grammar.

@@
   filtering stage proposed in \cite{kajita2003biped} to compute the
-  exact ZMP and take into account the full dynamics of the robot and
+  exact ZMP, take into account the full dynamics of the robot and
   generate feasible trajectories.}

### 2012-12-06 paper.tex
Fix confusion between submanifolds and manifols.

@@
   submanifold $\manifold$ of $\mathcal{C}$. $\manifold$ is called
-  the planning submanifold.}
+  the planning manifold.}
 
@@
   inverse kinematics balance constraints;
-\item the goal submanifold $\goalmanifold$ is also defined implicitly,
+\item the goal manifold $\goalmanifold$ is also defined implicitly,
   by additional inverse kinematics constraints.
@@
 a tree from the goal. To make use of the idea of growing multiple trees, we first
-randomly sample the goal submanifold and generate several goal configurations. Then, 
+randomly sample the goal manifold and generate several goal configurations. Then, 
 we grow random

### 2012-12-06 paper.tex
Fix typos pointed out by Florent.

@@
 
-\textbf{In this paper, we present a planning algorithm that considers exact models of a humanoid
-robot and its environment. It is used to solve navigation and manipulation problems. Our planner 
-generates collision-free, statically balanced
-geometric paths, i.e. paths such that for any path configurations, the robot is balanced on its
-feet with zero speed. We also present a constructive formal proof that these paths can be
-approximated by collision-free, dynamically balanced walk trajectories, i.e. trajectories such
-that for any trajectory instant, the robot is balanced on at least one foot. Our planner is designed for perfectly modeled indoor
-environments, where the floor is horizontal and flat. Also, because we do not compute explicitly footstep
-positions at the planning stage, the presented planner is unable to plan motions in which the
-robot steps over obstacles. These limitations are discussed in the paper.}
+\textbf{In this paper, we present a planning algorithm that considers
+  exact models of a humanoid robot and its environment. It is used to
+  solve navigation and manipulation problems. Our planner generates
+  collision-free, statically balanced geometric paths, i.e. paths such
+  that for any configuration along the path, the robot is balanced on
+  its feet with zero speed. We also present a constructive formal
+  proof that these paths can be approximated by collision-free,
+  dynamically balanced walk trajectories, i.e. trajectories satisfying
+  the Newton-Euler equations. Our planner is designed for perfectly
+  modeled indoor environments, where the floor is horizontal and
+  flat. Also, because we do not compute explicitly footstep positions
+  at the planning stage, the presented planner is unable to plan
+  motions in which the robot steps over obstacles. These limitations
+  are discussed in the paper.}
 
@@
   $\mathcal{C}=\mathcal{Q} \times SE(3)$, where $\mathcal{Q}$
-  represents the kinematic tree actuators and $SE(3)$ corresponds to a
-  fictitious floating joint attached to the root of the tree. If the
-  robot has $n$ actuated DoFs, then $dim(\mathcal{Q})=n$ and
+  represents the kinematic tree actuators and $SE(3)$ represents the
+  position in space of the root of the tree. If the robot has $n$
+  actuated DoFs, then $dim(\mathcal{Q})=n$ and
   $dim(\mathcal{C})=n+6$. A configuration $q$ of $\mathcal{C}$ is said
   to be valid iff, besides being collision-free, it lies on a specific
-  submanifold $\manifold$ of $\mathcal{C}$. $\manifold$ is called
-  the planning manifold.}
+  submanifold $\manifold$ of $\mathcal{C}$. $\manifold$ is called the
+  planning manifold.}
 
@@
 
-\textbf{We show here how a multiple constraints solver works: its
+\textbf{We show here how a multiple constraint solver works: its
   purpose is to find the root $q$ of a non-linear $C^1$ function
@@
   configuration on a manifold $\manifold$, $f(q)$ can be defined as a
-  vector function that contains the concatenation of all constraints
-  defining $\manifold$. Note that as the intersection of two or more
-  manifolds is also a manifold, this constraint solver allows us also
-  to generate configurations that lie at the intersection of several
-  manifolds.}
+  vector valued function that contains the concatenation of all
+  constraints defining $\manifold$. Note that as the intersection of
+  two or more manifolds is also a manifold, this constraint solver
+  allows us also to generate configurations that lie at the
+  intersection of several manifolds.}
 

### 2012-12-06 paper.tex
Proposal to take into account jpl comments.
  * Comment about constrained RRT (un)usefulness not addressed yet.

@@
   exact models of a humanoid robot and its environment. It is used to
-  solve navigation and manipulation problems. Our planner generates
-  collision-free, statically balanced geometric paths, i.e. paths such
-  that for any configuration along the path, the robot is balanced on
-  its feet with zero speed. We also present a constructive formal
-  proof that these paths can be approximated by collision-free,
-  dynamically balanced walk trajectories, i.e. trajectories satisfying
-  the Newton-Euler equations. Our planner is designed for perfectly
+  solve navigation and manipulation problems. Our planner is a
+  two-step algorithm: a first collision-free path is computed in the
+  quasi-static configuration space \footnote{A quasi-static
+    configuration is a configuration where the robot is balanced on
+    its feet with zero speed.}, then this first path is approximated
+  by a sequence of dynamic walking trajectories. The proof of
+  correctness of the algorithm is based on the concept of small-space
+  controllability. Such property allows to approximate any \textit{non
+    necessarily admissible} path, by a sequence of admissible ones. In
+  our context we prove that dynamic walking makes humanoid robots
+  small-space controllable. Our planner is designed for perfectly
   modeled indoor environments, where the floor is horizontal and
-  flat. Also, because we do not compute explicitly footstep positions
+  flat. Also, because we do not explicitly compute footprint positions
   at the planning stage, the presented planner is unable to plan
@@
 
-\textbf{In this work, we show that, under some assumptions, it is
-  sufficient to plan in the configuration space in a first step, even
-  though the final output of our algorithm is guaranteed to be a
-  dynamically balanced walk motion. This result was first presented in
-  \cite{dalibard2011small}.}
+\textbf{We show in this work that, under some assumptions, it is
+  sufficient to plan a first path in the quasi-static configuration
+  space of a humanoid robot, and then use the small-space
+  controllability property to approximate this path by an admissible
+  trajectory, i.e. a dynamically balanced walking motion. This result
+  was first presented in \cite{dalibard2011small}.}
 
@@
 
-Another contribution of this paper is the formal proof that this first
-path can always be approximated by a dynamically balanced,
-collision-free walking trajectory. We have implemented this
-well-grounded method, and the results have been validated on the HRP-2
-robot.
+Another contribution of this paper is the formal proof that ynamic
+walking makes humanoid robots small-space controllable, whith the
+direct implication that this first path can always be approximated by
+a dynamically balanced, collision-free walking trajectory. We have
+implemented this well-grounded method, and the results have been
+validated on the HRP-2 robot.
 
@@
 \textbf{This section presents an algorithm for constrained motion
-  planning in the configuration space $\mathcal{C}$. In the particular
-  case of a humanoid robot which is an under-actuated system,
-  $\mathcal{C}=\mathcal{Q} \times SE(3)$, where $\mathcal{Q}$
-  represents the kinematic tree actuators and $SE(3)$ represents the
-  position in space of the root of the tree. If the robot has $n$
-  actuated DoFs, then $dim(\mathcal{Q})=n$ and
+  planning on a submanifold $\manifold$ of the configuration space
+  $\mathcal{C}$. In the particular case of a humanoid robot which is
+  an under-actuated system, $\mathcal{C}=\mathcal{Q} \times SE(3)$,
+  where $\mathcal{Q}$ represents the kinematic tree actuators and
+  $SE(3)$ represents the position in space of the root of the tree. If
+  the robot has $n$ actuated DoFs, then $dim(\mathcal{Q})=n$ and
   $dim(\mathcal{C})=n+6$. A configuration $q$ of $\mathcal{C}$ is said
-  to be valid iff, besides being collision-free, it lies on a specific
-  submanifold $\manifold$ of $\mathcal{C}$. $\manifold$ is called the
-  planning manifold.}
+  to be valid iff, besides being collision-free, it lies on the
+  manifold $\manifold$; we call $\manifold$ the planning manifold.}
 
@@
   reality a legged robot cannot slide on a regular floor, such paths
-  are physically unfeasible.They are, however, easier to generate than
-  feasible dynamic walking trajectories because only geometric
+  are physically unfeasible. They are, however, easier to generate
+  than feasible dynamic walking trajectories because only geometric
   constraints are considered at planning time.}

### 2012-12-07 paper.tex
Add more modifications from jpl comments.

@@
   two-step algorithm: a first collision-free path is computed in the
-  quasi-static configuration space \footnote{A quasi-static
-    configuration is a configuration where the robot is balanced on
-    its feet with zero speed.}, then this first path is approximated
-  by a sequence of dynamic walking trajectories. The proof of
-  correctness of the algorithm is based on the concept of small-space
-  controllability. Such property allows to approximate any \textit{non
-    necessarily admissible} path, by a sequence of admissible ones. In
-  our context we prove that dynamic walking makes humanoid robots
-  small-space controllable. Our planner is designed for perfectly
-  modeled indoor environments, where the floor is horizontal and
-  flat. Also, because we do not explicitly compute footprint positions
-  at the planning stage, the presented planner is unable to plan
-  motions in which the robot steps over obstacles. These limitations
-  are discussed in the paper.}
+  space of quasi-statically balanced configurations, then this first
+  path is approximated by a sequence of dynamic walking trajectories,
+  i.e. trajectories satisfying the Newton-Euler equations. The proof
+  of correctness of the algorithm is based on the concept of
+  small-space controllability. This property allows, under some
+  assumptions, to approximate any \textit{non necessarily admissible}
+  path, by a sequence of admissible trajectories. In our context we
+  prove that dynamic walking makes humanoid robots small-space
+  controllable. Our planner is designed for perfectly modeled indoor
+  environments, where the floor is horizontal and flat. Also, because
+  we do not explicitly compute footprint positions at the planning
+  stage, the presented planner is unable to plan motions in which the
+  robot steps over obstacles. These limitations are discussed in the
+  paper.}
 
@@
 
-\textbf{The constrained motion planner and the small-space
+\textbf{Our constrained motion planner and the small-space
   controllability property can luckily still be used with different
-  environment representations that might more suited with perception
-  systems and their inherent measurement errors, using for instance
-  point clouds. In the future, we would like to achieve dynamic walk
-  planning using a Kinect sensor that would be mounted on the humanoid
-  robot. This would relax the a priori environment knowledge
-  constraint and bring us one step closer to reactive planning in
-  unknown environments.}
+  environment representations that might be more suited with
+  perception systems. Extensions in that directions
+  \cite{Nakhei4755945, Dang6100889} are currently under development to
+  relax the a priori environment knowledge constraint and achieve
+  reactive planning in unknown environments.}
 

### 2012-12-07 paper.tex
Fix comment about environment representation.

@@
   controllability property can luckily still be used with different
-  environment representations that might be more suited with
-  perception systems. Extensions in that directions
-  \cite{Nakhei4755945, Dang6100889} are currently under development to
-  relax the a priori environment knowledge constraint and achieve
-  reactive planning in unknown environments.}
+  environment representations that might be more suited for perception
+  systems. Extensions \cite{Nakhei4755945,DanLauLam2012} are currently
+  under development to relax the a priori environment knowledge
+  constraint and achieve reactive planning in unknown environments.}
 

### 2012-12-07 paper.tex
(Pending approval) Change author affiliation.
(Pending approval) Modify remark about proof generalization.

@@
     LAAS, F-31400 Toulouse Cedex 4, France. S\'ebastien Dalibard is
-    now with Nanyang Technological University, Singapore.}
+    now with Aldebaran Robotics, France.}
   \footnote{
@@
 \subsubsection{Remarks}
-\paragraph{Generalization to a complete model:} More work is needed to generalize 
-the previous proof to any legged robot model. First, because we use a model with
-a point mass, we did not have to consider the angular momentum of the robot
-around the CoM, while it accounts for dynamic balance. Moreover, when considering
-an actual kinematic and dynamic model,
-the positions of feet and CoM are not controlled DoFs, but inputs 
-for inverse kinematic tasks. Continuity and regularity results about the robot
-controller -- i.e. how much the configuration changes when the input of an inverse
-kinematics task changes--, would be needed to generalize the proof. We have not
-proven such properties for the inverse kinematics controller we use. 
-The  proof presented above still gives the intuition for the
-algorithm presented in the following.
+\paragraph{Generalization to a complete model:} 
+\textbf{We did not extend the previous proof to any legged robot model
+  since empirically, the table cart model is a very good fit for our
+  humanoid robot. Although of little practical interest, the
+  generalization of the proof does not seem very difficult to
+  achieve. As an insight, the difference between the table cart model
+  and the full size humanoid robot is due to the derivative of the
+  angular momentum and to the vertical acceleration of the center of
+  mass. These perturbations can be made as small as desired along the
+  sliding path by following the sliding path as slowly as
+  necessary. The derivatives of the angular momentum produced by the
+  stepping motion can also be made as small as desired by making the
+  step height as small as necessary and by using recent results on
+  properties of joint trajectories induced by end-effector motions
+  \cite{Zanchettin6084763}.}
 

### 2012-12-07 paper.tex
Integrate comments from Florent.

@@
 at a planning phase. When planning motion for humanoid robots, this is a particularly costly
-approach, as the size of the space to explore is augmented with the robot velocity and footstep
+approach, as the size of the space to explore is augmented with the robot velocity and footprint
 positions. The other way is to first plan  a geometric path that can be approximated by a
@@
   space of quasi-statically balanced configurations, then this first
-  path is approximated by a sequence of dynamic walking trajectories,
-  i.e. trajectories satisfying the Newton-Euler equations. The proof
-  of correctness of the algorithm is based on the concept of
-  small-space controllability. This property allows, under some
-  assumptions, to approximate any \textit{non necessarily admissible}
-  path, by a sequence of admissible trajectories. In our context we
-  prove that dynamic walking makes humanoid robots small-space
-  controllable. Our planner is designed for perfectly modeled indoor
-  environments, where the floor is horizontal and flat. Also, because
-  we do not explicitly compute footprint positions at the planning
-  stage, the presented planner is unable to plan motions in which the
-  robot steps over obstacles. These limitations are discussed in the
-  paper.}
+  path is approximated by a sequence of dynamic walking
+  trajectories. The proof of correctness of the algorithm is based on
+  the concept of small-space controllability. This property allows,
+  under some assumptions, to approximate any \textit{non necessarily
+    admissible} path, by a sequence of admissible trajectories. In our
+  context we prove that dynamic walking makes humanoid robots
+  small-space controllable. Our planner is designed for perfectly
+  modeled indoor environments, where the floor is horizontal and
+  flat. Also, because we do not explicitly compute footprint positions
+  at the planning stage, our planner is unable to plan motions in
+  which the robot steps over obstacles. These limitations are
+  discussed in the paper.}
 
@@
 
-This work is based on several fields of humanoid robotics
-research: prioritized inverse kinematics, randomized  whole-body  motion
-planning and walk pattern generators based on the Zero-Moment
-Point (ZMP) formalism. This section summarizes the literature related to each of 
-these fields.
+This work is based on several fields of humanoid robotics research:
+prioritized inverse kinematics, randomized whole-body motion planning
+and walk pattern generators based on the Zero-Moment Point (ZMP). This
+section summarizes the literature related to each of these fields.
 
@@
 dynamically balanced walk patterns. Since the introduction of the ZMP
-formalism  \cite{vukobratovic1969contribution},  several  methods  have  been  proposed  to
-generate  walking  motions efficiently.   One  way  to  deal with  the
-complexity  of a  humanoid  robot  kinematic tree  is  to use  the
-so-called "cart-table" simplified model \cite{kajita2003biped}. Based on this
-model,  planning a trajectory  for the  ZMP is  reduced to  planning a
-trajectory  for  the Center  of  Mass (CoM)  of  the  robot.  Given  a
-trajectory  of  the CoM  and  footprint  positions, inverse  kinematics
-solvers can animate  the whole set of DoFs of the
-robot to generate a dynamically balanced walk trajectory.
+\cite{vukobratovic1969contribution}, several methods have been
+proposed to generate walking motions efficiently.  One way to deal
+with the complexity of a humanoid robot kinematic tree is to use the
+so-called "cart-table" simplified model \cite{kajita2003biped}. Based
+on this model, planning a trajectory for the ZMP is reduced to
+planning a trajectory for the Center of Mass (CoM) of the robot.
+Given a trajectory of the CoM and footprint positions, inverse
+kinematics solvers can animate the whole set of DoFs of the robot to
+generate a dynamically balanced walk trajectory.
 
@@
 Some planning methods for free-climbing robots \cite{bretl2006motion}
-can be seen as a general way 
-to consider quasi-static multi-step planning. They are not however directly applicable to
-humanoid dynamically balanced locomotion.
-Other recent contributions to the field of locomotion planning  include algorithms 
-considering the dynamics at the planning phase \cite{shkolnik2011bounding}. This leads to 
-a growth of algorithmic complexity, particularly costly for high-dimensional
-systems such as humanoid robots. To the authors knowledge such techniques have not been
-used on humanoid robotic platforms so far.
+can be seen as a general way to consider quasi-static multi-step
+planning. They are not however directly applicable to humanoid
+dynamically balanced locomotion.  Other recent contributions to the
+field of locomotion planning include algorithms considering the
+dynamics at the planning phase \cite{shkolnik2011bounding}. This leads
+to a growth of algorithmic complexity, particularly costly for
+high-dimensional systems such as humanoid robots. To the authors'
+knowledge such techniques have not been used on humanoid robotic
+platforms so far.
 
@@
 
-Another contribution of this paper is the formal proof that ynamic
-walking makes humanoid robots small-space controllable, whith the
+Another contribution of this paper is the formal proof that dynamic
+walking makes humanoid robots small-space controllable, with the
 direct implication that this first path can always be approximated by
@@
 \end{enumerate}
-\textbf{During global planning, we will consider several types of
-equality constraints for various reasons:}
+During global planning, we will consider several types of constraints
+for various reasons:
 \begin{itemize}
-\item Static balance: the CoM of the robot stays above
-  the support polygon center, the two feet have fixed position and
-  orientation.
+\item \textbf{Static balance: the CoM of the robot stays above the support
+  polygon center, the two feet are horizontal on the ground.}
 \item End-effector position and orientation: the goals of some problems presented
@@
 controllability. It also uses the fact that balance criteria for dynamic walking are different
-from the ones for static balance. An important point is that our result and method do not hold
-for a biped robot that walks in a quasi-static way.
+from the ones for static balance.
 
@@
   trajectory, derive foot trajectories, and a preview controller
-  outputs a corresponding CoM trajectory. Classic numerical
+  returns the corresponding CoM trajectory. Classic numerical
   Jacobian-based prioritized inverse kinematics methods prove to be
@@
   accomplish secondary tasks, such as following a reference
-  configuration trajectory. We use the same framework which was
-  developed in \cite{yoshida2006tds}.}
-
-The stack of tasks applied to the robot to generate a dynamic walking
-motion is -- in decreasing priority order:
+  configuration trajectory. We use the framework called Generalized
+  inverse kinematics (Gik) and developed in \cite{yoshida2006tds}.}
+
+\textbf{The hierarchy of tasks (referred to as \textit{GikTasks} in
+  Algorithm ~\ref{alg:walk}) applied to the robot to generate a
+  dynamic walking motion is -- in decreasing priority order:}
 
@@
 
-\STATE $StackOfTasks$.initialize()
-
-\STATE $StackOfTasks$.addFootprintTask($Footprints$)
-
-\STATE $StackOfTasks$.addWaistTask()
-
-\STATE $StackOfTasks$.addConfigurationTask($p$)
+\STATE $GikTasks$.addFootprintTask($Footprints$)
+
+\STATE $GikTasks$.addWaistTask()
+
+\STATE $GikTasks$.addConfigurationTask($p$)
 
 \STATE $DynamicTrajectory \leftarrow
-\text{ComputeWalkTrajectory}(StackOfTasks)$
+\text{ComputeWalkTrajectory}(GikTasks)$
 
@@
   executed.}
+
+To get more natural motions in the experiments, we require the foot
+positions to be fixed with respect to each other, and the CoM to be
+projected in the center of the support polygon during the sliding path
+planning stage.
 
@@
 
-\textbf{The presented experiment setup assumes perfect knowledge of
-  the environment. This can be guaranteed during experiments by using
+\textbf{The experimental setup assumes perfect knowledge of the
+  environment. This can be guaranteed during experiments by using
   calibrated objects and motion capture systems. This indeed allows us
@@
   problem, interesting as it is, is thus completely decoupled from the
-  planning problem in our presentation.}
-
-\textbf{Our constrained motion planner and the small-space
-  controllability property can luckily still be used with different
-  environment representations that might be more suited for perception
-  systems. Extensions \cite{Nakhei4755945,DanLauLam2012} are currently
-  under development to relax the a priori environment knowledge
-  constraint and achieve reactive planning in unknown environments.}
+  planning problem in our presentation. From previous experiences
+  however, we are confident that our algorithm will work as well in
+  environments modeled by vision \cite{Nakhei4755945,DanLauLam2012}.}
 
@@
 
-The presented setup also assumes perfect execution of the plan. It can
-be critical here, since non-nominal stepping may cause the robot to
-drift away from the planned trajectory. Future experiments will
-include trajectory following during plan execution.
+The setup also assumes perfect execution of the plan. It can be
+critical here, since non-nominal stepping may cause the robot to drift
+away from the planned trajectory. Future experiments will include
+trajectory following during plan execution.
 

### 2012-12-07 paper.tex
Update constrained extend algorithm.

@@
   purpose is to find the root $q$ of a non-linear $C^1$ function
-  $f(q)$ with a tolerance of $\varepsilon$. If we want to find a
+  $f(q)$ with a tolerance of $\epsilon$. If we want to find a
   configuration on a manifold $\manifold$, $f(q)$ can be defined as a
@@
 
-\textbf{We observed that values of $\varepsilon=10^{-6}$,
+\textbf{We observed that values of $\epsilon=10^{-6}$,
   $\alpha=0.1$, $\alpha_{max}=0.95$ and $w=0.8$ lead to good behavior,
@@
 \begin{algorithm}
-\caption{\texttt{SolveConstraints}($q$, f, $\varepsilon$): find $q$
+\caption{\texttt{SolveConstraints}($q$, f, $\epsilon$): find $q$
   such that $f(q) = 0$}
@@
 \STATE $i=0$
-\WHILE{$\|f(q)\| > \varepsilon$ and $i\leq it_{max}$}
+\WHILE{$\|f(q)\| > \epsilon$ and $i\leq it_{max}$}
 \STATE {\color{red} // $\left(.\right)^{+}$ denotes the Moore-Penrose pseudo-inverse}
@@
 \ENDWHILE
-\IF {$\|f(q)\| \leq \varepsilon$}
+\IF {$\|f(q)\| \leq \epsilon$}
 \STATE return $q$
@@
 \draw [dashed,thin] (qn) -- (450,277);
+\node [right = 0.2cm] at (450,277) {\small{$T_q\manifold$}};
 \node [draw,circle,inner sep=2pt] (qr2) at (367,277) {};
@@
     Algorithm \ref{alg:constrained}: $q_{rand}$ is first projected on
-    the tangent space of $\manifold$. $q_{rand}'$ is then projected
-    onto $\manifold$, and a classic RRT extension tries to go as far
-    as possible from $q_{near}$ towards $q_{rand}''$ while remaining
-    on $\manifold$. $q_{new}$ is then returned.}}
+    $T_q\manifold$ the tangent space of $\manifold$. $q_{rand}'$ is
+    then projected onto $\manifold$. A classic RRT extension tries to
+    go as far as possible from $q_{near}$ towards $q_{rand}''$ while
+    remaining on $\manifold$. $q_{new}$ is then returned.}}
 \label{fig:gikrrt}
@@
   \begin{algorithmic}
-    \STATE {\color{red} // $I$ denotes the identity matrix.}
-    \STATE $q_{rand}' \leftarrow q_{near} +   
-    \left(I -  
-    \left(\frac{\partial f}{\partial q}(q_{near})\right)^{+}
-    \frac{\partial f}{\partial q}(q_{near})\right) (q_{rand} - q_{near})$
-    \IF {$\texttt{SolveConstraints}(q_{rand}',f,\epsilon) = q_{rand}''$}
-    \STATE RRT::Extend($\mathcal{T},q_{near},q_{rand}''$)
-    \ENDIF
+    \STATE $d \leftarrow$ \texttt{Distance}($q_{near}, q_{rand}$)
+    \STATE $q \leftarrow$ $q_{near}$
+    \WHILE{$d > \epsilon$}
+    \STATE $q_{rand}' \leftarrow$ \texttt{OrthogonalProject}($q_{rand}, T_q\manifold$)
+    \STATE $q_{rand}'' \leftarrow$ \texttt{SolveConstraints}($q_{rand}',f,\epsilon$)
+    \STATE $d \leftarrow$ \texttt{Distance}($q,q_{rand}''$)
+    \STATE $q \leftarrow q_{rand}''$
+    \ENDWHILE
+    \STATE $q_{new} \leftarrow$ RRT::Extend($\mathcal{T},q_{near},q_{rand}''$)
   \end{algorithmic}

### 2012-12-07 paper.tex
Spellcheck run.

@@
 constrained random motion planning on a fixed dimension manifold is 
-probabilisticaly complete. This proof applies to our algorithm. }
+probabilistically complete. This proof applies to our algorithm. }
 

### 2013-02-06 paper.tex
Update description of benchmark results extension.
Place extension description as first appendix and add ijrr url.
Mention extensions in figures.

@@
 
-\caption{HRP-2 grabs a ball on a shelf, puts it on another shelf, and comes back to 
-  a rest position. Static balance constraints are enforced along the path, and 
-  the intermediary goals consisting in grasping and displacing the ball are defined
-  implicitly as inverse kinematics constraints.}
+\caption{HRP-2 grabs a ball on a shelf, puts it on another shelf, and
+  comes back to a rest position. Static balance constraints are
+  enforced along the path, and the intermediary goals consisting in
+  grasping and displacing the ball are defined implicitly as inverse
+  kinematics constraints. Extension 1 shows a video of the motion.}
 \label{fig:wb-shelves}
@@
   locomotion. When the robot is close to obstacles, the amplitude of
-  the oscillations decreases.}
+  the oscillations decreases. Extension 2 shows a video of the
+  executed trajectory on the real HRP-2.}
 \label{fig:chairs-waist}
@@
 
-\caption{Execution of the walking trajectory by HRP-2 on stage. The robot first goes
-  to the shelves to release the ball, then comes back to a rest position.}
+\caption{Execution of the walking trajectory by HRP-2 on stage. The
+  robot first goes to the shelves to release the ball, then comes back
+  to a rest position. Extension 3 shows a video of this motion.}
 
@@
 \appendix
-\section{Sliding Motion Planning Benchmarks}
-\label{app:bench}
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=0.8\linewidth]{plots/rrt-it.eps}
-\caption{Number of RRT iterations $it$ for the floating objects and the
-  shelf scenarios, using two variants of RRT. Mean $\overline{it}$,
-  standard deviation $\sigma_{it}$, minimum and maximum values are
-  represented.}
-\label{fig:rrt-it}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=0.8\linewidth]{plots/rrt-t.eps}
-\caption{RRT computation time $t$ for the floating objects and the shelf
-  scenarios, using two variants of RRT. Mean $\overline{t}$, standard
-  deviation $\sigma_{t}$, minimum and maximum values are represented.}
-\label{fig:rrt-t}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=0.8\linewidth]{plots/rrt-n.eps}
-\caption{Number of tree nodes $n$ for the floating objects and the
-  shelf scenarios, using two variants of RRT. Mean $\overline{n}$,
-  standard deviation $\sigma_{n}$, minimum and maximum values are
-  represented.}
-\label{fig:rrt-n}
-\end{figure}
-
 \section{Index to Multimedia Extensions}
+
+The multimedia extensions to this article are at:
+\url{http://www.ijrr.org}.
 
@@
 in simulation and on HRP-2. \\
-4 & Data & Raw data of sliding motion planning benchmarks for all
-scenarios. \\
+4 & Data & Raw sliding motion planning benchmark results for Chairs,
+Floating objects and Shelf Scenarios. \\
 \hline
@@
 
+\section{Sliding Motion Planning Benchmarks}
+\label{app:bench}
+
+All plots were generated from data available in Extension 4.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.8\linewidth]{plots/rrt-it.eps}
+\caption{Number of RRT iterations $it$ for the floating objects and the
+  shelf scenarios, using two variants of RRT. Mean $\overline{it}$,
+  standard deviation $\sigma_{it}$, minimum and maximum values are
+  represented.}
+\label{fig:rrt-it}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.8\linewidth]{plots/rrt-t.eps}
+\caption{RRT computation time $t$ for the floating objects and the shelf
+  scenarios, using two variants of RRT. Mean $\overline{t}$, standard
+  deviation $\sigma_{t}$, minimum and maximum values are represented.}
+\label{fig:rrt-t}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=0.8\linewidth]{plots/rrt-n.eps}
+\caption{Number of tree nodes $n$ for the floating objects and the
+  shelf scenarios, using two variants of RRT. Mean $\overline{n}$,
+  standard deviation $\sigma_{n}$, minimum and maximum values are
+  represented.}
+\label{fig:rrt-n}
+\end{figure}
+
 \bibliographystyle{agsm}
